---
title: "Comparison of different models with real data"
author: "David Bolin, Alexandre B. Simas, and Jonas Wallin"
date: "Created: 2024-02-09. Last modified: `r Sys.Date()`."
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Comparison of different models with real data}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
references:
- id: BSW2022a
  title: "Gaussian Whittle--Matérn fields on metric graphs"
  author:
  - family: Bolin
    given: David
  - family: Simas
    given: Alexandre B.
  - family: Wallin
    given: Jonas
  container-title: Bernoulli
  type: article
  issued:
    year: 2023
- id: BSW2022b
  title: "Statistical properties of Gaussian Whittle--Matérn fields on metric graphs"
  author:
  - family: Bolin
    given: David
  - family: Simas
    given: Alexandre B.
  - family: Wallin
    given: Jonas
  container-title: arXiv:2304.10372
  type: preprint
  issued:
    year: 2023
- id: Anderes2020
  title: "Isotropic covariance functions on graphs and their edges"
  author:
  - family: Anderes
    given: Ethan
  - family: Møller
    given: Jesper
  - family: Rasmussen
    given: Jakob G
  container-title: Annals of Statistics
  type: article
  issue: 48
  pages: 2478-2503
  issued:
    year: 2020
- id: Borovitskiy2021
  title: "Matérn Gaussian processes on graphs"
  author:
  - family: Borovitskiy
    given: Viacheslav
  - family: Azangulov
    given: Iskander
  - family: Terenin
    given: Alexander 
  - family: Mostowsky
    given: Peter
  - family: Deisenroth
    given: Marc
  - family: Durrande
    given: Nicolas
  container-title: International Conference on Artificial Intelligence and Statistics
  type: article
  pages: 2593-2601
  issued: 
    year: 2021
- id: LindgrenRue2015
  title: Bayesian Spatial Modelling with R-INLA.
  author:
  - family: Lindgren
    given: Finn
  - family: Rue
    given: Haavard
  container-title: Journal of Statistical Software
  type: article
  issue: 63
  pages: 1-25
  issued:
    year: 2015
- id: inlabru2019
  title: inlabru an R package for Bayesian spatial modelling from ecological survey data
  author:
  - family: Bachl
    given: Fabian E.
  - family: Lindgren
    given: Finn
  - family: Borchers
    given: David L.
  - family: Illian
    given: Janine B.
  container-title: Methods in Ecology and Evolution
  type: article
  issue: 10
  pages: 760-766
  issued:
    year: 2019
- id: sppackage
  title: Applied spatial data analysis with R
  author:
  - family: Bivand
    given: Roger S.
  - family: Pebesma
    given: Edzer
  - family: Gomez-Rubio
    given: Virgilio
  publisher: Springer, NY
  type: book
  issued:
    year: 2013
- id: plotlypackage
  title: Interactive Web-Based Data Visualization with R, plotly, and shiny
  author:
  - family: Sievert
    given: Carson
  publisher: Chapman and Hall/CRC
  type: book
  issued:
    year: 2020
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
library(MetricGraph)
```

# Introduction

In this vignette our goal is to compare the predictive power of different models on metric graphs by cross-validation. More precisely, we consider the Whittle--Matérn fields introduced by [@BSW2022a](https://arxiv.org/abs/2205.06163) and [@BSW2022b](https://arxiv.org/abs/2304.10372), a Gaussian
processes with an isotropic exponential covariance functions [@Anderes2020](https://projecteuclid.org/journals/annals-of-statistics/volume-48/issue-4/Isotropic-covariance-functions-on-graphs-and-their-edges/10.1214/19-AOS1896.full), and Matérn Gaussian processes based on the graph Laplacian [@Borovitskiy2021](http://proceedings.mlr.press/v130/borovitskiy21a/borovitskiy21a.pdf). 

# The dataset

For this example we will consider the `pems` data contained in the `MetricGraph` package. The data consists of traffic speed observations on highways in the city of San Jose, California. The variable `y` contains the traffic speeds.

```{r}
 pems_graph <- metric_graph$new(edges = pems$edges, longlat=TRUE)
 pems_graph$add_observations(data = pems$data, normalized=TRUE)
 pems_graph$prune_vertices()
```

Let us take a look at the observations:

```{r}
pems_graph$plot(data = "y", vertex_size = 0)
```

# The models

Let us fit the different models. We start by fitting a Whittle-Matérn field with `alpha = 1`:

```{r}
fit_alpha1 <- graph_lme(y ~ -1, graph=pems_graph, model = list(type = "WhittleMatern", alpha = 1))
```

and look at its summary:

```{r}
summary(fit_alpha1)
```

Now, we will fit a Whittle-Matérn field with `alpha = 2`:

```{r}
fit_alpha2 <- graph_lme(y ~ -1, graph=pems_graph, model = list(type = "WhittleMatern", alpha = 2))
```

and its summary:

```{r}
summary(fit_alpha2)
```

We will now fit Whittle-Matérn fields with `alpha = 1` and `alpha=2`, and by performing a boundary correction on vertices of degree 1. To such an end, we set `BC=1`:


```{r}
fit_alpha1_bc <- graph_lme(y ~ -1, graph=pems_graph, BC = 1,
        model = list(type = "WhittleMatern", alpha = 1))

fit_alpha2_bc <- graph_lme(y ~ -1, graph=pems_graph, BC = 1,
        model = list(type = "WhittleMatern", alpha = 2))
```

Now, let us look at the summaries:

```{r}
summary(fit_alpha1_bc)
```
and
```{r}
summary(fit_alpha2_bc)
```

Similarly, let us now fit a Matérn Gaussian model based on the graph Laplacian for `alpha=1` and `alpha=2`:

```{r}
fit_GL1 <- graph_lme(y ~ -1, graph=pems_graph, model = list(type = "graphLaplacian", alpha = 1))

fit_GL2 <- graph_lme(y ~ -1, graph=pems_graph, model = list(type = "graphLaplacian", alpha = 2))
```

and look at their summaries:

```{r}
summary(fit_GL1)
```

and

```{r}
summary(fit_GL2)
```

Observe that the default optimizer (L-BFGS-B) failed to converge, thus an alternative optimizer was used, when fitting the graph Laplacian model with `alpha=2`.

Let us now fit a Gaussian field with isotropic exponential covariance function:

```{r}
fit_isoexp <- graph_lme(y ~ -1, graph=pems_graph, model = list(type = "isoCov"))
```

Observe the warning, message. This message tells us that we did not check if the metric graph has Euclidean edges. Let us check now:

```{r}
pems_graph$check_euclidean()
```

Now, we can look at the graph's characteristics contained in this summary:

```{r}
summary(pems_graph)
```

Thus, this graph does not have Euclidean edges. This means, when we fit the model, we actually modify the graph, when we add the obvservations as vertices. By default, when fitting isotropic models, the `graph_lme()` function checks it the metric graph after adding the observations as vertices has Euclidean edges. We can check by looking at the `euclidean` element of the fitted model:

```{r}
fit_isoexp$euclidean
```

Therefore, even after adding the observations as vertices, the metric graph is still not Euclidean.

Let us check if we can fit an isotropic model with isotropic Matérn covariance function with smoothness parameter `nu=1.5` (which corresponds to `alpha=2` for Whittle-Matérn fields and Matérn Gaussian models based on the graph Laplacian). Recall that by the results in [@Anderes2020](https://projecteuclid.org/journals/annals-of-statistics/volume-48/issue-4/Isotropic-covariance-functions-on-graphs-and-their-edges/10.1214/19-AOS1896.full), it is not guaranteed that such a function will be a valid covariance function even on an metric graph with Euclidean edges.

We start by defining the Matérn covariance:

```{r}
cov_mat <- function(h,theta){
    nu <- 1.5
    sigma <- theta[1]
    kappa <- theta[2]
    (sigma^2/(2^(nu - 1) * gamma(nu))) * ((kappa * abs(h))^nu) * 
            besselK(kappa * abs(h), nu)
        }
```

Let us now fit the model:

```{r, error=TRUE}
fit_isomat <- graph_lme(y ~ -1, graph = pems_graph, model = list(type = "isoCov", cov_function = cov_mat), 
                                    model_options = list(start_par_vec = c(1,1)))
```

Indeed, the model could not be fitted.

# Comparison by cross-validation

We will now use the function `posterior_crossvalidation()` to perform 
leave-one-out cross validation based on the estimated parameters and compare
the results:

```{r}
fitted_models_list <- list("alpha=1" = fit_alpha1, 
                    "alpha=2" = fit_alpha2, 
                    "isoExp" = fit_isoexp, 
                    "GL1" = fit_GL1, "GL2" = fit_GL2,
                    "alpha=1 bc" = fit_alpha1_bc,
                    "alpha=2 bc" = fit_alpha2_bc)

posterior_crossvalidation(fitted_models_list, factor = 1000)[["scores"]]
```
