---
title: "Working with disconnected metric graphs"
author: "David Bolin, Alexandre B. Simas, and Jonas Wallin"
date: "Created: `r Sys.Date()`. Last modified: `r Sys.Date()`."
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Working with disconnected metric graphs}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, fig.align = "center")
library(MetricGraph)
set.seed(1)
```

# Introduction

A metric graph need not be connected: real-world networks (street segments,
isolated stretches of river, fragmented sensor coverage, etc.) often consist of
several disjoint components. The `metric_graph` class supports this case
directly — pass `check_connected = FALSE` to skip the connectedness warning at
construction time, and then use the same methods (`add_observations`,
`compute_geodist`, `compute_resdist`, `spde_precision`, `graph_lme`,
`sample_spde`, ...) as on a connected graph. Distance methods produce `Inf` for
vertex pairs in different components and SPDE precision matrices come out
block-diagonal by construction, so no special bookkeeping is required.

This vignette walks through the disconnected workflow: constructing the graph,
inspecting and routing across components, computing distances and SPDE models,
and per-component visualisation.

For the basics of constructing and working with a single connected metric
graph, see [Working with metric graphs](metric_graph.html).

> **Note.** A previous release shipped a separate `graph_components` class for
> this purpose. It is deprecated and now emits a warning at construction; the
> functionality demonstrated below replaces it.

# Constructing a disconnected metric graph

We build a small graph with three disjoint components:

* **Component A** is a small "L"-shape with a curved arc.
* **Component B** is a triangle.
* **Component C** is a single straight edge.

```{r, message=FALSE}
# Component A: L-shape with a curved arc
edgeA1 <- rbind(c(0, 0), c(1, 0))
edgeA2 <- rbind(c(0, 0), c(0, 1))
edgeA3 <- rbind(c(0, 1), c(-1, 1))
theta  <- seq(from = pi, to = 3 * pi / 2, length.out = 20)
edgeA4 <- cbind(sin(theta), 1 + cos(theta))

# Component B: a triangle, translated to the right
edgeB1 <- rbind(c(4, 0), c(5, 0))
edgeB2 <- rbind(c(5, 0), c(4.5, 1))
edgeB3 <- rbind(c(4.5, 1), c(4, 0))

# Component C: a single straight edge, placed below
edgeC1 <- rbind(c(1, -2), c(3, -2))

edges <- list(edgeA1, edgeA2, edgeA3, edgeA4,
              edgeB1, edgeB2, edgeB3,
              edgeC1)
```

Setting `check_connected = FALSE` suppresses the "graph is disconnected"
message:

```{r, message=FALSE}
graph <- metric_graph$new(edges = edges, check_connected = FALSE)
graph$nE
graph$nV
```

The graph behaves exactly like any other `metric_graph`. Observations,
meshes, and SPDE models all work without further setup; the rest of this
vignette focuses on the bits that are specific to the disconnected case.

# Inspecting the components

`get_components()` returns the connected components as a list of
`metric_graph` objects. Components are sorted by total edge length,
descending:

```{r}
comps <- graph$get_components()
length(comps)
sapply(comps, function(g) g$nE)
sapply(comps, function(g) sum(as.numeric(g$edge_lengths)))
```

The largest component is just `comps[[1]]`:

```{r}
g_largest <- comps[[1]]
g_largest$nE
```

For a connected graph, `get_components()` simply returns `list(self)`, so the
same call works uniformly.

## Determining the component of a spatial point

`which_component()` snaps each spatial point to the nearest network location
and returns the component index:

```{r}
XY <- rbind(c(0.5,  0.0),   # near component A
            c(4.5,  0.5),   # inside the triangle (component B)
            c(2.0, -2.0))   # on the bottom edge (component C)
graph$which_component(XY)
```

This is the same routing rule that `add_observations()` uses internally for
spatial input.

# Plotting

`plot()` accepts a `components` argument that draws every component in a
different colour:

```{r, message=FALSE}
graph$plot(components = TRUE)
```

You can also pass an `n × 3` matrix of RGB values (one row per component) for
explicit colour control:

```{r, message=FALSE}
cols <- rbind(c(0.85, 0.10, 0.10),
              c(0.10, 0.45, 0.85),
              c(0.10, 0.65, 0.20))
graph$plot(components = cols)
```

# Adding observations

Observations are added with the usual `add_observations()` method. For
`data_coords = "spatial"` each point is snapped to the nearest edge
automatically:

```{r}
n_obs <- 30
df_spatial <- data.frame(
  coord_x = c(runif(n_obs, -1, 1),
              runif(n_obs,  4,  5),
              runif(n_obs,  1,  3)),
  coord_y = c(runif(n_obs,  0, 1),
              runif(n_obs,  0, 1),
              rep(-2, n_obs)),
  y       = rnorm(3 * n_obs)
)

graph$add_observations(data = df_spatial, data_coords = "spatial",
                       verbose = 0, tolerance = 3)
```

`get_data()` returns the observations exactly as for a connected graph. Pair
with `which_component()` if you need a per-component breakdown:

```{r}
obs <- graph$get_data()
table(graph$which_component(cbind(obs$.coord_x, obs$.coord_y)))
```

For `data_coords = "PtE"`, the usual `(edge_number, distance_on_edge)` columns
are used; edge numbering is global across the whole graph.

# Distances on disconnected graphs

`compute_geodist()` and `compute_resdist()` are aware that vertices in
different components have infinite distance and produce `Inf` on the
cross-component blocks (with finite values within each component):

```{r}
graph$compute_resdist(full = TRUE)
R <- graph$res_dist[[".complete"]]
# Cross-component pairs are Inf
sum(is.infinite(R)) > 0
# Within-component pairs are finite
all(is.finite(diag(R)))
```

This means isotropic models such as `graph_lme(model = "isoExp")` produce
exactly zero correlation between components without any special handling.

# SPDE models

The SPDE precision matrix `spde_precision()` builds a per-edge precision
matrix and assembles it via vertex indices in `graph$E`. Because edges in
different components touch disjoint vertex sets, the resulting matrix is
block-diagonal automatically:

```{r}
Q <- spde_precision(kappa = 2, tau = 1, alpha = 1, graph = graph)
dim(Q)
# Verify block structure: vertices of component k only connect within k
membership <- igraph::components(
  igraph::make_graph(c(t(graph$E)), directed = FALSE), mode = "weak"
)$membership
cross <- outer(membership, membership, FUN = "!=")
max(abs(as.matrix(Q)[cross]))
```

Fitting a Whittle--Matérn model is no different from the connected case:

```{r, message=FALSE, warning=FALSE}
graph$clear_observations()
df_fit <- data.frame(
  coord_x = c(runif(50, -1, 1), runif(50, 4, 5), runif(50, 1, 3)),
  coord_y = c(runif(50,  0, 1), runif(50, 0, 1), rep(-2, 50)),
  y       = rnorm(150)
)
graph$add_observations(data = df_fit, data_coords = "spatial",
                       verbose = 0, tolerance = 3)

fit <- graph_lme(y ~ 1, graph = graph, model = "WM1")
as.numeric(stats::logLik(fit))
```

`sample_spde()` likewise samples a Whittle--Matérn field on a disconnected
mesh:

```{r}
graph$build_mesh(h = 0.1)
graph$compute_fem()
samp <- sample_spde(graph = graph, alpha = 1, kappa = 5, tau = 1,
                    type = "mesh")
graph$plot_function(X = samp, vertex_size = 1)
```

# Plotting functions on the components

`plot_function()` works on disconnected graphs, both for stored data
(`data = "y"`) and for an arbitrary `newdata` indexed by the global
`.edge_number`:

```{r, message=FALSE, warning=FALSE}
graph$plot_function(data = "y")
```

```{r, message=FALSE, warning=FALSE}
n_per_edge <- 30
newdata <- do.call(rbind, lapply(seq_len(graph$nE), function(e) {
  data.frame(.edge_number      = e,
             .distance_on_edge = seq(0, 1, length.out = n_per_edge))
}))
newdata$f <- with(newdata, sin(2 * pi * .distance_on_edge) + 0.5 * .edge_number)
class(newdata) <- c("metric_graph_data", class(newdata))
graph$plot_function(data = "f", newdata = newdata)
```
