---
title: "Working with disconnected metric graphs"
author: "David Bolin, Alexandre B. Simas, and Jonas Wallin"
date: "Created: `r Sys.Date()`. Last modified: `r Sys.Date()`."
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Working with disconnected metric graphs}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, fig.align = "center")
library(MetricGraph)
set.seed(1)
```

# Introduction

A metric graph need not be connected: real-world networks (street segments,
isolated stretches of river, fragmented sensor coverage, etc.) often consist of
several disjoint components.  In `MetricGraph`, the `graph_components` class
represents such a graph as a collection of `metric_graph` objects, one per
connected component.

`graph_components` mirrors the public interface of `metric_graph`. The same
methods used to add data, build a mesh, compute finite-element matrices, and
plot functions are available, and they are dispatched to the appropriate
component automatically. This vignette illustrates this functionality on a
small example.

For the basics of constructing and working with a single connected metric
graph, see [Working with metric graphs](metric_graph.html).

# Constructing a disconnected metric graph

We build a small graph with three disjoint components:

* **Component A** is a small "L"-shape with a curved arc, similar to the
  example used in the other vignettes.
* **Component B** is a triangle.
* **Component C** is a single straight edge.

```{r, message=FALSE}
# Component A: L-shape with a curved arc
edgeA1 <- rbind(c(0, 0), c(1, 0))
edgeA2 <- rbind(c(0, 0), c(0, 1))
edgeA3 <- rbind(c(0, 1), c(-1, 1))
theta  <- seq(from = pi, to = 3 * pi / 2, length.out = 20)
edgeA4 <- cbind(sin(theta), 1 + cos(theta))

# Component B: a triangle, translated to the right
edgeB1 <- rbind(c(4, 0), c(5, 0))
edgeB2 <- rbind(c(5, 0), c(4.5, 1))
edgeB3 <- rbind(c(4.5, 1), c(4, 0))

# Component C: a single straight edge, placed below
edgeC1 <- rbind(c(1, -2), c(3, -2))

edges <- list(edgeA1, edgeA2, edgeA3, edgeA4,
              edgeB1, edgeB2, edgeB3,
              edgeC1)
```

We construct the `graph_components` object directly from the list of edges:

```{r, message=FALSE}
graphs <- graph_components$new(edges = edges)
```

The number of components, the number of vertices in each, and the total edge
length per component are exposed as fields:

```{r}
graphs$n
graphs$sizes
graphs$lengths
```

By default, components are sorted by total edge length (set
`by_length = FALSE` in the constructor to sort by vertex count instead). The
underlying `metric_graph` objects are stored in `graphs$graphs`:

```{r}
class(graphs$graphs[[1]])
```

The `plot` method draws every component on a single ggplot, using a different
colour for each:

```{r, message=FALSE}
graphs$plot()
```

If only the largest component is of interest, `get_largest()` returns it as a
standalone `metric_graph` object that can be used with the rest of the package
exactly as before:

```{r}
g_largest <- graphs$get_largest()
class(g_largest)
g_largest$nE
```

# Determining the component of a spatial point

For a set of spatial coordinates, `which_component()` returns the index of the
component each point is closest to (in Euclidean snap distance):

```{r}
XY <- rbind(c(0.5,  0.0),   # near component A
            c(4.5,  0.5),   # inside the triangle (component B)
            c(2.0, -2.0))   # on the bottom edge (component C)
graphs$which_component(XY)
```

This is the routing rule used internally when adding observations in spatial
coordinates.

# Adding observations

`graph_components$add_observations()` accepts the same arguments as
`metric_graph$add_observations()`, with two extensions:

* When `data_coords = "spatial"`, each observation is routed to the component
  whose nearest network location is closest. No further bookkeeping is needed
  from the user.
* When `data_coords = "PtE"`, the data must include a column identifying the
  component, alongside the usual `edge_number` and `distance_on_edge`
  columns. The convention is the triple
  `(component number, edge number, distance on edge)`.

## Adding spatial observations

We sample some random locations near each component and let
`add_observations()` route them automatically:

```{r}
n_obs <- 30
df_spatial <- data.frame(
  coord_x = c(runif(n_obs, -1, 1),       # near component A
              runif(n_obs,  4,  5),      # near component B
              runif(n_obs,  1,  3)),     # on component C
  coord_y = c(runif(n_obs,  0, 1),
              runif(n_obs,  0, 1),
              rep(-2, n_obs)),
  y       = rnorm(3 * n_obs)
)

graphs$add_observations(data = df_spatial, data_coords = "spatial",
                        verbose = 0)
```
Note that some observations were not added as they were too far away from the graph. To add all observations,
we can increase the `tolerance` argument which is passed to the addition in the different components. 
```{r}
graphs$add_observations(data = df_spatial, 
                        data_coords = "spatial",
                        clear_obs = TRUE,
                        verbose = 0,
                        tolerance = 3)
```
`get_data()` retrieves all observations as a tibble; the additional
`.component` column tells us which component each row was routed to:

```{r}
obs <- graphs$get_data()
obs
table(obs$.component)
```

The same data can be visualised on the graph with `plot(data = ...)`:

```{r, message=FALSE}
graphs$plot(data = "y")
```

Note: `plot()` automatically skips components that do not contain the
specified data column, so this works seamlessly even if some components have
no observations.

## Adding observations in graph coordinates

When the data is already specified in graph coordinates, supply a column with
the component index. Here we add four observations, one on each component
(component C only has one edge):

```{r}
df_pte <- data.frame(
  component        = c(1, 1, 2, 3),
  edge_number      = c(1, 2, 1, 1),
  distance_on_edge = c(0.25, 0.75, 0.5, 0.5),
  z                = c(10, 20, 30, 40)
)

graphs$add_observations(data = df_pte, data_coords = "PtE",
                        normalized = TRUE, verbose = 0)
```

The new observations are added to the existing ones (set `clear_obs = TRUE`
to wipe the existing data first). We can confirm the routing by inspecting
`get_PtE()`, which returns a matrix of `(component, edge, distance)` triples:

```{r}
tail(graphs$get_PtE(), 4)
```

To remove all observations from every component, use `clear_observations()`:

```{r}
graphs$clear_observations()
```

We re-add the spatial observations for the rest of the vignette:

```{r}
graphs$add_observations(data = df_spatial, data_coords = "spatial",
                        verbose = 0)
```

## Grouped observations

The `group` argument behaves exactly as in `metric_graph$add_observations()`.
The `get_groups()` method returns the union of group identifiers across all
components:

```{r}
df_grouped <- data.frame(
  coord_x = c(0.2, 0.6, 4.5, 2.5),
  coord_y = c(0.0, 0.0, 0.5, -2.0),
  rep     = c("a", "b", "a", "b"),
  y2      = c(1, 2, 3, 4)
)

graphs$add_observations(data = df_grouped, data_coords = "spatial",
                        group = "rep", verbose = 0)
graphs$get_groups()
```

# Converting between coordinate systems

The `coordinates()` method converts between graph coordinates (now expressed
as `(component, edge, distance)`) and spatial coordinates.

From PtE to XY:

```{r}
PtE_in <- rbind(c(1, 1, 0.5),   # midpoint of edge 1 of component 1
                c(2, 1, 0.5),   # midpoint of edge 1 of component 2
                c(3, 1, 0.5))   # midpoint of the only edge of component 3
graphs$coordinates(PtE = PtE_in, normalized = TRUE)
```

From XY to PtE — note that the result has three columns, with the first being
the component index:

```{r}
graphs$coordinates(XY = rbind(c(0.5, 0.0),
                              c(4.5, 0.5),
                              c(2.0, -2.0)))
```

The two operations round-trip:

```{r}
all.equal(
  graphs$coordinates(XY = graphs$coordinates(PtE = PtE_in, normalized = TRUE)),
  PtE_in,
  tolerance = 1e-8
)
```

# Building meshes and FEM matrices

`build_mesh()` and `compute_fem()` are forwarded to every component, so the
same call equips all of them at once:

```{r}
graphs$build_mesh(h = 0.1)
graphs$compute_fem()

# Inspect the per-component mesh sizes
sapply(graphs$graphs, function(g) nrow(g$mesh$V))
```

Each component now has its own mass matrix `C`, stiffness matrix `G`, and so
on, ready to be used by the FEM-based models elsewhere in the package.

# Plotting functions on the components

`plot_function()` works in two modes, both mirroring
`metric_graph$plot_function()`.

## Plotting a stored data column

Pass `data = "<column name>"` to plot an observed quantity. Components without
that column are skipped silently:

```{r, message=FALSE, warning = FALSE}
graphs$plot_function(data = "y")
```

## Plotting a function specified at arbitrary locations

To plot values at user-supplied locations, pass `newdata` as a
`metric_graph_data` object containing the columns `.component`,
`.edge_number`, `.distance_on_edge`, and the value column. Below we draw a
smooth function on each component:

```{r, message=FALSE, warning = FALSE}
make_locs <- function(component_id, n_e, n_per_edge = 30) {
  do.call(rbind, lapply(seq_len(n_e), function(e) {
    data.frame(.component        = component_id,
               .edge_number      = e,
               .distance_on_edge = seq(0, 1, length.out = n_per_edge))
  }))
}

newdata <- rbind(
  make_locs(1, n_e = graphs$graphs[[1]]$nE),
  make_locs(2, n_e = graphs$graphs[[2]]$nE),
  make_locs(3, n_e = graphs$graphs[[3]]$nE)
)
newdata$f <- with(newdata, sin(2 * pi * .distance_on_edge) +
                            0.5 * .component)
class(newdata) <- c("metric_graph_data", class(newdata))

graphs$plot_function(data = "f", newdata = newdata)
```

# Modelling with `graph_components`

The SPDE machinery (`graph_lme`, `graph_spde`, `sample_spde`,
`spde_precision`, `spde_variance`, `spde_covariance`,
`make_Q_spacetime`, `make_Q_euler`, `simulate_spacetime`) accepts a
`graph_components` object directly. Internally each function either
block-stacks the per-component matrices (for example, the precision
matrix is `Matrix::bdiag()` of the per-component precision matrices) or
delegates the work to each component independently. No
`metric_graph$new()` constructor pipeline runs in the process, so
calling these functions on a `graph_components` is essentially as cheap
as calling them on a single connected `metric_graph`.

For example, fitting a Whittle--Matérn model with the same syntax as on
a regular metric graph:
```{r, message=FALSE, warning=FALSE}
graphs$clear_observations()
df_fit <- data.frame(
  coord_x = c(runif(50, -1, 1), runif(50, 4, 5), runif(50, 1, 3)),
  coord_y = c(runif(50, 0, 1), runif(50, 0, 1), rep(-2, 50)),
  y       = rnorm(150)
)
graphs$add_observations(data = df_fit, data_coords = "spatial",
                        verbose = 0)

fit <- graph_lme(y ~ 1, graph = graphs, model = "WM1")
```

Sampling a Whittle--Matérn field on the mesh of every component in one
call:
```{r}
sample <- sample_spde(graph = graphs, alpha = 1,
                      kappa = 5, tau = 1,
                      type = "mesh")
length(sample)
```

For "manual" sampling at user-specified locations, pass a 3-column
`PtE` matrix `(component, edge, distance)`:
```{r}
PtE_in <- rbind(c(1, 1, 0.25), c(2, 2, 0.5), c(3, 1, 0.5))
sample_spde(graph = graphs, alpha = 1,
            kappa = 5, tau = 1,
            PtE = PtE_in, type = "manual")
```

The marginal variance of the field at every mesh location:
```{r}
v <- spde_variance(kappa = 5, tau = 1, alpha = 1, graph = graphs)
length(v)
```

# Avoiding repeated assembly with `as_metric_graph()`

When passing a `graph_components` to one of these functions, the
package internally calls `as_metric_graph()` to produce a single
combined `metric_graph` view of the components. The conversion bypasses
the regular `metric_graph$new()` constructor — it stacks the
per-component fields directly, with edge and vertex indices shifted by
the cumulative offsets, and block-diagonally combines the FEM matrices.
The cost is therefore $O(N)$ memcopy rather than the
$O(N \log N)$ vertex deduplication of the full constructor.

For repeated downstream calls (e.g., several `sample_spde()` invocations
or fitting several models on the same data), call `as_metric_graph()`
once yourself to amortise even that small assembly cost:

```{r}
mg <- graphs$as_metric_graph()
class(mg)
mg$is_disconnected()
```

The returned object is an ordinary `metric_graph` and can be passed to
the same downstream functions:
```{r, warning=FALSE, message=FALSE}
fit2 <- graph_lme(y ~ 1, graph = mg, model = "WM1")
```

The two fits are numerically equivalent (the second simply re-uses the
already-assembled view):
```{r}
all.equal(as.numeric(stats::logLik(fit)),
          as.numeric(stats::logLik(fit2)))
```

# The `disconnected` flag

A `metric_graph` returned by `as_metric_graph()` carries a
`disconnected = TRUE` flag, queryable via `metric_graph$is_disconnected()`:
```{r}
mg$is_disconnected()
```

The flag protects against a class of subtle errors: the edge numbering
of `mg` is the *combined* one across components, not the
per-component numbering of the original `graph_components`. Adding
observations directly to `mg` is therefore disallowed:

```{r, error=TRUE}
mg$add_observations(
  data = data.frame(edge_number = 1, distance_on_edge = 0.5, y = 1),
  verbose = 0
)
```

The intended workflow is to add data to the original
`graph_components` and call `as_metric_graph()` again:

```{r, warning=FALSE, message=FALSE}
graphs$add_observations(
  data = data.frame(coord_x = 1.5, coord_y = -2, y3 = 99),
  data_coords = "spatial", verbose = 0
)
mg <- graphs$as_metric_graph()
```

# Summary of the API mirrored from `metric_graph`

The table below summarises the methods that `graph_components` adds to mirror
the `metric_graph` API. Each method dispatches to the underlying components
and either combines the results (for query methods) or routes the input to
the appropriate component (for write methods).

| Method                    | Behaviour                                                         |
|---------------------------|-------------------------------------------------------------------|
| `which_component(XY)`     | Closest component for each spatial point                          |
| `add_observations(...)`   | Routes spatial points; splits PtE data by `component` column      |
| `clear_observations()`    | Clears observations on every component                            |
| `get_data(...)`           | Combined data with an extra `.component` column                   |
| `get_PtE()`               | Matrix of `(component, edge, distance)` triples                   |
| `get_groups()`            | Union of group identifiers across components                      |
| `coordinates(PtE, XY)`    | Converts between `(component, edge, distance)` and `XY`           |
| `build_mesh(...)`         | Builds a mesh on every component                                  |
| `compute_fem(...)`        | Computes FEM matrices on every meshed component                   |
| `plot(...)`               | Plots all components; supports `data = ...`                       |
| `plot_function(...)`      | Plots a function on each component (uses `.component` in newdata) |
| `as_metric_graph()`       | Cheap, block-stacked view as a single (disconnected) `metric_graph` |

These methods are designed so that, in most use cases, working with a
disconnected graph requires no more code than working with a connected one,
and at no extra cost — `graph_lme`, `graph_spde`, `sample_spde`, and friends
all accept a `graph_components` directly.
