---
title: "INLA and inlabru interfaces"
author: "David Bolin, Alexandre B. Simas, and Jonas Wallin"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{INLA and inlabru interfaces}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
references:
- id: BSW2022a
  title: "Gaussian Whittle--Matéern fields on metric graphs"
  author:
  - family: Bolin
    given: David
  - family: Simas
    given: Alexandre B.
  - family: Wallin
    given: Jonas
  container-title: arXiv:2205.06163
  type: preprint
  issued:
  year: 2022
- id: BSW2022b
  title: "Statistical properties of Gaussian Whittle--Matérn fields on metric graphs"
  author:
  - family: Bolin
    given: David
  - family: Simas
    given: Alexandre B.
  - family: Wallin
    given: Jonas
  container-title: arXiv:??
  type: preprint
  issued:
  year: 2022
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
set.seed(1)
library(GPGraph)
```

# Introduction

In this vignette we will present our `R-INLA` and `inlabru` 
interfaces to Whittle--Matérn fields. The underlying theory
for this approach is provided in 
[@BSW2022a](https://arxiv.org/abs/2205.06163) and [@BSW2022b](??). 

For an introduction to the `metric_graph` class, please see 
the [Working with metric graphs](metric_graph.html) vignette.

In the [Gaussian random fields on metric graphs](random_fields.html) vignette,
we introduce all the models in metric graphs contained in this package,
as well as, how to perform statistical tasks on these models, but without
the `R-INLA` or `inlabru` interfaces.

We will present our `R-INLA` and `inlabru` interfaces to 
the Whittle-Matérn fields by providing a step-by-step illustration.

The Whittle--Matérn fields are specified as solutions to the stochastic
differential equation 
$$
  (\kappa^2 - \Delta)^{\alpha} \tau u = \mathcal{W}
$$
on the metric graph $\Gamma$. We can work with these models without
any approximations if the smoothness parameter $\alpha$ is an integer, and this 
is what we focus on in this vignette. For details on the case of a general 
smoothness parameter, see [Whittle--Matérn fields with general smoothness](fem_models.html). 

# A toy dataset 

Let us begin by loading the `GPGraph` package
and creating a metric graph:

```{r}
library(GPGraph)

line1 <- Line(rbind(c(0,0),c(1,0)))
line2 <- Line(rbind(c(0,0),c(0,1)))
line3 <- Line(rbind(c(0,1),c(-1,1)))
theta <- seq(from=pi,to=3*pi/2,length.out = 20)
line4 <- Line(cbind(sin(theta),1+ cos(theta)))
Lines = sp::SpatialLines(list(Lines(list(line1),ID="1"),
                              Lines(list(line2),ID="2"),
                              Lines(list(line4),ID="3"),
                              Lines(list(line3),ID="4")))
graph <- metric_graph$new(Lines = Lines)
```

Let us add 50 random locations in each edge where we will 
have observations:

```{r}
obs_per_edge <- 50
obs_loc <- NULL
for(i in 1:(graph$nE)) {
  obs_loc <- rbind(obs_loc,
                   cbind(rep(i,obs_per_edge), 
                   runif(obs_per_edge)))
}
```

We will now sample in these observation locations
and plot the latent field:

```{r, fig.show='hold',fig.align = "center",fig.width=7,echo=TRUE}
sigma <- 2
alpha <- 1
nu <- alpha - 0.5
r <- 0.15 # r stands for range
kappa <- sqrt(8 * nu) / r

u <- sample_spde(kappa = kappa, sigma = sigma, alpha = alpha,
                 graph = graph, PtE = obs_loc)
graph$plot(X = u, X_loc = obs_loc)
```

Let us now generate the observed responses, which we will
call `y`. We will also plot the observed responses on the
metric graph.

```{r, fig.show='hold',fig.align = "center",fig.width=7,echo=TRUE}
n_obs <- length(u)
sigma.e <- 0.1

y <- u + sigma.e * rnorm(n_obs)
graph$plot(X = y, X_loc = obs_loc)
```

# `R-INLA` implementation

We are now in a position to fit the model with our
`R-INLA` implementation. To this end,
we need to add the observations to the graph, which we 
will do with the `add_observations2()` method. Then,
we need to turn the observations into vertices of the 
graph with the `observation_to_vertex()` method. 

Before adding the observations we will grab a copy of
the graph with the `clone()` method for two reasons: i. to make it better to 
visualization; ii. we will need to provide another
approach of kriging as we will see in the *Kriging
with `R-INLA` implementation* section.

```{r, fig.show='hold',fig.align = "center",fig.width=7,echo=TRUE}
# Cloning the graph
graph_bkp <- graph$clone()
# Adding observations and turning them to vertices
graph$add_observations2(y, PtE = obs_loc, normalized=TRUE)
graph$observation_to_vertex()
graph$plot(data=TRUE)
```

Now, we load the `R-INLA` package and
create the `inla` model object with the
`gpgraph_spde` function. By default we have
`alpha=1`. 

```{r}
library(INLA)
spde_model <- gpgraph_spde(graph)
```

Now, we need to create the index object with 
the `graph_spde_make_index()` function:

```{r}
spde_index <- graph_spde_make_index(name="field", graph=graph)
```

Observe that, as in `R-INLA`, we need to name the field.

The next step is to create the observation matrix, to which
we use the `graph_spde_make_A()` function:

```{r}
A <- graph_spde_make_A(graph)
```

The remaining is standard in `R-INLA`. We create the
formula object and the
`inla.stack` object with the `inla.stack()` function:

```{r}
f.s <- y ~ -1 + Intercept + f(field, model = spde_model)

stk.dat <- inla.stack(data = list(y=as.vector(y)), 
                        A = A, 
                        effects = c(
      spde_index,
      list(Intercept = 1)
    ))
```

Now, we use the `graph_stack()` function that behaves in 
an analogous manner to the `inla.stack.data()` function:
```{r}
data_stk <- graph_stack(stk.dat, "field")
```

Finally, we fit the model:

```{r}
spde_fit <- inla(f.s, data = data_stk)
```

Let us now obtain the estimates in the original scale
by using the `spde_metric_graph_result()` function, then
taking a `summary()`:

```{r}
spde_result <- spde_metric_graph_result(spde_fit, "field", spde_model)

summary(spde_result)
```

We will now compare the means of the estimated values with
the true values:

```{r}
  result_df <- data.frame(
    parameter = c("std.dev", "range"),
    true = c(sigma, r),
    mean = c(
      spde_result$summary.sigma$mean,
      spde_result$summary.range$mean
    ),
    mode = c(
      spde_result$summary.sigma$mode,
      spde_result$summary.range$mode
    )
  )
  print(result_df)
```


We can also plot the posterior marginal densities with the 
help of the `gg_df()` function:
```{r, fig.show='hold',fig.align = "center",echo=TRUE, fig.width=7, warning=FALSE}
  posterior_df_fit <- gg_df(spde_result)

  ggplot(posterior_df_fit) + geom_line(aes(x = x, y = y)) + 
  facet_wrap(~parameter, scales = "free") + labs(y = "Density")
```

## Kriging with the `R-INLA` implementation

We can have two approaches to do kriging with our `R-INLA`
implementation of Whittle-Matérn fields on metric graphs.

* In the first we add a second set of `NA` add_observations
at the locations in which we want to do kriging to the 
graph in which we fitted the model. Then, fit a new model,
and obtain the posterior mean at the desired locations.
Observe that on this method, if we use the
`add_observations2()` method, we will need to provide the locations
in which we want to predict in terms of the positions of
the new graph, which contains the observed set of 
observations as vertices.

* In the second method we add the locations in which we
want to predict together with the observations and 
observed locations. In this method we provide the locations we
want to predict in terms of the same base graph, as the
observations are not vertices of the base graph.

We will now do kriging at an evenly spaced mesh across
the metric graph.

## Kriging by adding new observation locations to the fitted graph

Since the fitted graph contains the observed locations as vertices,
we will use our backup graph `graph_bkp` to obtain the spatial 
locations of the points in which we want to do kriging.
We will create a copy of `graph_bkp` as we will still need
`graph_bkp` for the second approach of kriging.

```{r}
graph_tmp <- graph_bkp$clone()
```

Let us obtain an evenly spaced mesh with respect to the base
graph:

```{r}
obs_per_edge_prd <- 50
obs_loc_prd <- NULL
loc_tmp <- 0:(obs_per_edge_prd+1)/(obs_per_edge_prd+1)
loc_tmp <- loc_tmp[2:(obs_per_edge_prd+1)]
for(i in 1:(graph_tmp$nE)) {
    obs_loc_prd <- rbind(obs_loc_prd,
                 cbind(rep(i,obs_per_edge_prd), 
                 loc_tmp))
}
```
Let us add the locations as vertices and plot the resulting graph:
```{r, fig.show='hold',fig.align = "center",echo=TRUE, fig.width=7, warning=FALSE}
n_obs_prd <- obs_per_edge_prd * graph_tmp$nE
y_prd <- rep(NA, n_obs_prd)
graph_tmp$add_observations2(y_prd, obs_loc_prd,
                            normalized = TRUE)
graph_tmp$observation_to_vertex()
graph_tmp$plot()
```
We will now extract the spatial locations and use
these to add these new locations to the graph
we used to fit the `R-INLA` model:

```{r}
spatial_loc_prd <- graph_tmp$Points
```

We will now add them to the graph we fitted the
`R-INLA` model:

```{r}
graph$add_observations(Spoints = spatial_loc_prd)
graph$observation_to_vertex()
```

Let us plot the resulting graph (the gray dots indicate
the new locations):

```{r, fig.show='hold',fig.align = "center",echo=TRUE, fig.width=7, warning=FALSE}
graph$plot(data=TRUE)
```

We will now fit a new model with `R-INLA` with this new
graph that contains the prediction locations. To this end,
we create a new model object with the `gpgraph_spde()` function:

```{r}
spde_model_prd <- gpgraph_spde(graph)
```

Now, let us create a new index object and a new observation (`A`):
```{r}
spde_index_prd <- graph_spde_make_index(name="field", graph=graph)

A_prd <- graph_spde_make_A(graph)
```

We will create a new vector of response variables,
concatenating `y` to `y_prd`, then create a new
formula object and the
`inla.stack` object:

```{r}
y_cmp <- c(y, y_prd)

f_s_prd <- y ~ -1 + Intercept + f(field, model = spde_model_prd)

stk_dat_prd <- inla.stack(data = list(y=y_cmp), 
                        A = A_prd, 
                        effects = c(
      spde_index_prd,
      list(Intercept = 1)
    ))
```

Now, we use the `graph_stack()` function and fit the model:
```{r}
data_stk_prd <- graph_stack(stk_dat_prd, "field")

spde_fit_prd <- inla(f_s_prd, data = data_stk_prd)
```

Let us now obtain the estimates in the original scale:

```{r}
spde_result_prd <- spde_metric_graph_result(spde_fit_prd, "field", spde_model_prd)

summary(spde_result_prd)
```

We will now extract the means at the prediction locations:
```{r}
idx_prd <- which(is.na(y_cmp))

m_prd <- spde_fit_prd$summary.fitted.values$mean[idx_prd]
```

To improve visualization, we will plot the posterior means
on `graph_bkp`:
```{r, fig.show='hold',fig.align = "center",echo=TRUE, fig.width=7, warning=FALSE}
graph_bkp$plot(X = m_prd, X_loc = obs_loc_prd)
```

We will now plot the predictions together with
the data. To this end we need to create a new 
vector `y` which is a concatenation of `y`
and `m_prd`, then use it as the vector of 
responses of `graph`, which is done by using the
`add_responses()` method. Then, we plot.

Let us begin by concatenating the vectors and adding
the responses to `graph`:
```{r}
y_new <- c(y, m_prd)
graph$add_responses(y_new)
```

Now, we plot:
```{r, fig.show='hold',fig.align = "center",echo=TRUE, fig.width=7, warning=FALSE}
graph$plot(data=TRUE)
```

## Kriging by adding the prediction locations from the very beginning

This method is much simpler than the previous one. In this method
we concatenate the observation locations with the prediction locations,
assign `NA` as the response values at the prediction locations,
then proceed as in the first part of this vignette.

We start by getting a new copy of `graph_bkp` to be 
used in this second
prediction approach.

```{r}
graph_prd <- graph_bkp$clone()
```

We will use `obs_loc` as the observation locations and 
`obs_loc_prd` as the prediction locations. To this end,
we create a new matrix of locations called `obs_loc_prd2`
by vertically concatenating these two location matrices:
```{r}
obs_loc_prd2 <- rbind(obs_loc, obs_loc_prd)
```

Recall that the response variable is the concatenation
of `y` and a vector of `NA`. We already have this vector,
which we called `y_cmp` in the first prediction part.

We will now add these observations and turn them to
vertices:

```{r, fig.show='hold',fig.align = "center",echo=TRUE, fig.width=7, warning=FALSE}
graph_prd$add_observations2(y_cmp, PtE = obs_loc_prd2, 
                            normalized=TRUE)
graph_prd$observation_to_vertex()
graph_prd$plot(data=TRUE)
```

Now, we proceed as in the first part. We first create
the model:
```{r}
spde_model_prd2 <- gpgraph_spde(graph_prd)
```

We now create the observation (`A`) matrix, the index object
and the stack object:

```{r}
A_prd2 <- graph_spde_make_A(graph_prd)

spde_index_prd2 <- graph_spde_make_index(name="field", 
                                        graph=graph_prd)

stk_dat_prd2 <- inla.stack(data = list(y=y_cmp), 
                        A = A_prd2, 
                        effects = c(
      spde_index_prd2,
      list(Intercept = 1)
    ))
```

Now we create the formula object:
```{r}
f_s_prd2 <- y ~ -1 + Intercept + f(field, model = spde_model_prd2)
```
Now, we use the `graph_stack()` function:
```{r}
data_stk_prd2 <- graph_stack(stk_dat_prd2, "field")
```

Finally, we fit the model:

```{r}
spde_fit_prd2 <- inla(f_s_prd2, data = data_stk_prd2)
```

Let us now obtain the estimates in the original scale:

```{r}
spde_result_prd2 <- spde_metric_graph_result(spde_fit_prd2, 
                                    "field", spde_model_prd2)

summary(spde_result_prd2)
```


We will now extract the means at the prediction locations:
```{r}
idx_prd <- which(is.na(y_cmp))

m_prd2 <- spde_fit_prd2$summary.fitted.values$mean[idx_prd]
```

To improve visualization, we will plot the posterior means
on `graph_bkp`:
```{r, fig.show='hold',fig.align = "center",echo=TRUE, fig.width=7, warning=FALSE}
graph_bkp$plot(X = m_prd2, X_loc = obs_loc_prd)
```

We will now plot the predictions together with
the data. To this end we need to create a new 
vector `y` which is a concatenation of `y`
and `m_prd`, then use it as the vector of 
responses of `graph`, which is done by using the
`add_responses()` method. Then, we plot.

Let us begin by concatenating the vectors and adding
the responses to `graph`:
```{r}
y_new2 <- c(y, m_prd2)
graph_prd$add_responses(y_new2)
```

Now, we plot:
```{r, fig.show='hold',fig.align = "center",echo=TRUE, fig.width=7, warning=FALSE}
graph_prd$plot(data=TRUE)
```

## Fitting `R-INLA` models with replicates

We will now illustrate how to use our `R-INLA` implementation to 
fit models with replicates.

To simplify exposition, we will use the same base graph.
So, we begin by creating a copy:
```{r}
graph_rep <- graph_bkp$clone()
```

We will use the same observation locations as for the
previous cases. Let us sample 30 replicates:
```{r}
sigma_rep <- 1.5
alpha_rep <- 1
nu_rep <- alpha_rep - 0.5
r_rep <- 0.2 # r stands for range
kappa_rep <- sqrt(8 * nu_rep) / r_rep

n_repl <- 30

u_rep <- sample_spde(kappa = kappa_rep, sigma = sigma_rep,
                 alpha = alpha_rep,
                 graph = graph_rep, PtE = obs_loc,
                 nsim = n_repl)
```

Let us now generate the observed responses, which we will
call `y_rep`.

```{r, fig.show='hold',fig.align = "center",fig.width=7,echo=TRUE}
n_obs_rep <- nrow(u_rep)
sigma_e <- 0.1

y_rep <- u_rep + sigma_e * matrix(rnorm(n_obs_rep * n_repl),
                                    ncol=n_repl)
```

Since we are considering a model with replicates, we will
add only the first replicate to the graph object:
```{r, fig.show='hold',fig.align = "center",fig.width=7,echo=TRUE}
graph_rep$add_observations2(y_rep[,1], 
                    PtE = obs_loc, normalized=TRUE)
graph_rep$observation_to_vertex()
graph_rep$plot(data=TRUE)
```

Let us now create the model object:

```{r}
spde_model_rep <- gpgraph_spde(graph_rep)
```

We now create the index object by using the
`graph_spde_make_index()` function with the
argument `n.repl` set to the number of replicates:
```{r}
spde_index_rep <- graph_spde_make_index(name="field", 
                    graph=graph_rep, n.repl = n_repl)
```

Similarly, we create the observation (`A`) matrix with the
function `graph_spde_make_A()` setting the `n.repl` argument
to the number of replicates:
```{r}
A_rep <- graph_spde_make_A(graph_rep, n.repl = n_repl)
```

Next, we create the stack object, remembering that
we need to input the response variable `y` as a vector
and the `sample_spde()` function returns a matrix.
```{r}
stk_dat_rep <- inla.stack(data = list(y=as.vector(y_rep)), 
                        A = A_rep, 
                        effects = c(
      spde_index_rep,
      list(Intercept = 1)
    ))
```

We now create the formula object, adding the
name of the field (in our case `field`)
attached with `.repl` a the `replicate` argument
inside the `f()` function.
```{r}
f_s_rep <- y ~ -1 + Intercept + 
    f(field, model = spde_model_rep, 
        replicate = field.repl)
```
Then, we create the stack object with The
`graph_stack()` function:
```{r}
data_stk_rep <- graph_stack(stk_dat_rep, "field")
```
Now, we fit the model:
```{r}
spde_fit_rep <- inla(f_s_rep, data = data_stk_rep)
```

Let us see the estimated values in the original scale:
```{r}
spde_result_rep <- spde_metric_graph_result(spde_fit_rep, 
                        "field", spde_model_rep)

summary(spde_result_rep)
```

Let us compare with the true values:

```{r}
  result_df_rep <- data.frame(
    parameter = c("std.dev", "range"),
    true = c(sigma_rep, r_rep),
    mean = c(
      spde_result_rep$summary.sigma$mean,
      spde_result_rep$summary.range$mean
    ),
    mode = c(
      spde_result_rep$summary.sigma$mode,
      spde_result_rep$summary.range$mode
    )
  )
  print(result_df_rep)
```

# `inlabru` implementation

We will now present our `inlabru` implementation
of the Whittle-Matérn fields for metric graphs.
It has the advantage of not requiring the user
to provide observation matrices, indices nor 
stack objects.

Similarly to our `R-INLA` implementation,
we need to add the observations to the graph, which we 
will do with the `add_observations2()` method. Then,
we need to turn the observations into vertices of the 
graph with the `observation_to_vertex()` method. 

Before adding the observations, we will proceed
as we did for the `R-INLA` implementation and 
grab a copy of the base graph (`graph_bkp`) with the 
`clone()` method.
In this section we will also use the toy dataset,
so the response variable is `y` and the observation
locations are `obs_loc`.

```{r, fig.show='hold',fig.align = "center",fig.width=7,echo=TRUE}
# Cloning the graph
graph_bru <- graph_bkp$clone()
# Adding observations and turning them to vertices
graph_bru$add_observations2(y, PtE = obs_loc, normalized=TRUE)
graph_bru$observation_to_vertex()
graph_bru$plot(data=TRUE)
```

Now, we load the `inlabru` package. We will also
need to create the `inla` model object with the
`gpgraph_spde` function. By default we have
`alpha=1`. 

```{r}
library(inlabru)
spde_model_bru <- gpgraph_spde(graph_bru)
```
The next step is to create the data list containing
the reponse variables and the observation locations:
```{r}
data_list <- list(y = y, loc = obs_loc)
```

Now, we create `inlabru`'s component, which is a 
formula-like object:
```{r}
cmp <-
    y ~ -1 + Intercept(1) + field(loc, 
                            model = spde_model_bru)
```

Now, we directly fit the model:

```{r}
spde_bru_fit <-
    bru(cmp, data=data_list)
```

Let us now obtain the estimates in the original scale
by using the `spde_metric_graph_result()` function, then
taking a `summary()`:

```{r}
spde_bru_result <- spde_metric_graph_result(spde_bru_fit, 
                    "field", spde_model_bru)

summary(spde_bru_result)
```

We will now compare the means of the estimated values with
the true values:

```{r}
  result_df_bru <- data.frame(
    parameter = c("std.dev", "range"),
    true = c(sigma, r),
    mean = c(
      spde_bru_result$summary.sigma$mean,
      spde_bru_result$summary.range$mean
    ),
    mode = c(
      spde_bru_result$summary.sigma$mode,
      spde_bru_result$summary.range$mode
    )
  )
  print(result_df_bru)
```

We can also plot the posterior marginal densities with the 
help of the `gg_df()` function:
```{r, fig.show='hold',fig.align = "center",echo=TRUE, fig.width=7, warning=FALSE}
  posterior_df_bru_fit <- gg_df(spde_bru_result)

  ggplot(posterior_df_bru_fit) + geom_line(aes(x = x, y = y)) + 
  facet_wrap(~parameter, scales = "free") + labs(y = "Density")
```

## Kriging with the `inlabru` implementation

Unfortunately, our `inlabru` implementation is not 
compatible with `inlabru`'s `predict()` method. This has
to do with the nature of the metric graph's object.

Therefore, the approach to do kriging is very similar 
to `R-INLA`'s implementation. However, it is much simpler 
since it does not require the user to handle 
observation matrices, indices nor on stack objects.

Therefore, in the same manner as in `R-INLA`'s
implementation, we have two approaches to do 
kriging with our `inlabru`
implementation of Whittle-Matérn fields on metric graphs.

* In the first we add a second set of `NA` add_observations
at the locations in which we want to do kriging to the 
graph in which we fitted the model. Then, fit a new model,
and obtain the posterior mean at the desired locations.
Observe that on this method, if we use the
`add_observations2()` method, we will need to provide the locations
in which we want to predict in terms of the positions of
the new graph, which contains the observed set of 
observations as vertices.

* In the second method we add the locations in which we
want to predict together with the observations and 
observed locations. In this method we provide the locations we
want to predict in terms of the same base graph, as the
observations are not vertices of the base graph.

As in `R-INLA`'s illustration, we will do kriging at an evenly spaced mesh across
the metric graph.

## Kriging by adding new observation locations to the fitted graph

Since the fitted graph contains the observed locations as vertices,
we will use our backup graph `graph_bkp` to obtain the spatial 
locations of the points in which we want to do kriging.
We will create a copy of `graph_bkp` as we will still need
`graph_bkp` for the second approach of kriging.

```{r}
graph_tmp <- graph_bkp$clone()
```

Let us obtain an evenly spaced mesh with respect to the base
graph:

```{r}
obs_per_edge_prd <- 50
obs_loc_prd <- NULL
loc_tmp <- 0:(obs_per_edge_prd+1)/(obs_per_edge_prd+1)
loc_tmp <- loc_tmp[2:(obs_per_edge_prd+1)]
for(i in 1:(graph_tmp$nE)) {
    obs_loc_prd <- rbind(obs_loc_prd,
                 cbind(rep(i,obs_per_edge_prd), 
                 loc_tmp))
}
```
Let us add the locations as vertices and plot the resulting graph:
```{r, fig.show='hold',fig.align = "center",echo=TRUE, fig.width=7, warning=FALSE}
n_obs_prd <- obs_per_edge_prd * graph_tmp$nE
y_prd <- rep(NA, n_obs_prd)
graph_tmp$add_observations2(y_prd, obs_loc_prd,
                            normalized = TRUE)
graph_tmp$observation_to_vertex()
graph_tmp$plot()
```
We will now extract the spatial locations and use
these to add these new locations to the graph
we used to fit the `inlabru` model:

```{r}
spatial_loc_prd <- graph_tmp$Points
```

We will now add them to the graph we fitted the
`inlabru` model:

```{r}
graph_bru$add_observations(Spoints = spatial_loc_prd)
graph_bru$observation_to_vertex()
```

Let us plot the resulting graph (the gray dots indicate
the new locations):

```{r, fig.show='hold',fig.align = "center",echo=TRUE, fig.width=7, warning=FALSE}
graph_bru$plot(data=TRUE)
```

We will now fit a new model with `inlabru` with this new
graph that contains the prediction locations. To this end,
we create a new model object with the `gpgraph_spde()` function:

```{r}
spde_model_bru_prd <- gpgraph_spde(graph_bru)
```

We will create a new vector of response variables,
concatenating `y` to `y_prd`. We will also
concatenate the observation and prediction
locations vertically. Then, create 
a new data list
```{r}
y_cmp <- c(y, y_prd)
obs_loc_cmp <- rbind(obs_loc, obs_loc_prd)

data_list_prd <- list(y = y_cmp, loc = obs_loc_cmp)
```

Now, let us create a new
`inlabru` component:

```{r}
cmp_prd <- y ~ -1 + Intercept(1) + field(loc, 
                            model = spde_model_bru_prd)
```

Now, we fit the model:
```{r}
spde_bru_fit_prd <-
    bru(cmp_prd, data=data_list_prd)
```

Let us now obtain the estimates in the original scale:

```{r}
spde_result_bru_prd <- spde_metric_graph_result(spde_bru_fit_prd, 
            "field", spde_model_bru_prd)

summary(spde_result_bru_prd)
```

We will now extract the means at the prediction locations:
```{r}
idx_prd <- which(is.na(y_cmp))

m_prd_bru <- spde_bru_fit_prd$summary.fitted.values$mean[idx_prd]
```

To improve visualization, we will plot the posterior means
on `graph_bkp`:
```{r, fig.show='hold',fig.align = "center",echo=TRUE, fig.width=7, warning=FALSE}
graph_bkp$plot(X = m_prd_bru, X_loc = obs_loc_prd)
```

We will now plot the predictions together with
the data. To this end we need to create a new 
vector `y` which is a concatenation of `y`
and `m_prd`, then use it as the vector of 
responses of `graph`, which is done by using the
`add_responses()` method. Then, we plot.

Let us begin by concatenating the vectors and adding
the responses to `graph`:
```{r}
y_new_bru <- c(y, m_prd_bru)
graph_bru$add_responses(y_new_bru)
```

Now, we plot:
```{r, fig.show='hold',fig.align = "center",echo=TRUE, fig.width=7, warning=FALSE}
graph_bru$plot(data=TRUE)
```

## Kriging by adding the prediction locations from the very beginning

As for the `R-INLA` implementation, 
this method is much simpler than the previous one. 
In this method
we concatenate the observation locations with the 
prediction locations,
assign `NA` as the response values at 
the prediction locations,
then proceed as in the first part of this vignette.

We start by getting a new copy of `graph_bkp` to be 
used in this second
prediction approach.

```{r}
graph_bru_prd <- graph_bkp$clone()
```

We will use `obs_loc` as the observation locations and 
`obs_loc_prd` as the prediction locations. To this end,
we create a new matrix of locations called `obs_loc_prd2`
by vertically concatenating these two location matrices:
```{r}
obs_loc_prd2 <- rbind(obs_loc, obs_loc_prd)
```

Recall that the response variable is the concatenation
of `y` and a vector of `NA`. We already have this vector,
which we called `y_cmp` in the first prediction part.

We will now add these observations and turn them to
vertices:

```{r, fig.show='hold',fig.align = "center",echo=TRUE, fig.width=7, warning=FALSE}
graph_bru_prd$add_observations2(y_cmp, PtE = obs_loc_prd2, 
                            normalized=TRUE)
graph_bru_prd$observation_to_vertex()
graph_bru_prd$plot(data=TRUE)
```

Now, we proceed as in the first part. We first create
the model:
```{r}
spde_model_bru_prd2 <- gpgraph_spde(graph_bru_prd)
```
Next, we create the new data list:
```{r}
data_list_prd2 <- list(y = y_cmp, 
                            loc = obs_loc_prd2)
```

Now we create the component object:
```{r}
cmp_prd2 <- y ~ -1 + Intercept(1) + field(loc, 
                            model = spde_model_bru_prd2)
```
Finally, we fit the model:

```{r}
spde_bru_fit_prd2 <-
    bru(cmp_prd2, data=data_list_prd2)
```

Let us now obtain the estimates in the original scale:

```{r}
spde_result_bru_prd2 <- spde_metric_graph_result(spde_bru_fit_prd2, 
                                    "field", spde_model_bru_prd2)

summary(spde_result_bru_prd2)
```


We will now extract the means at the prediction locations:
```{r}
idx_prd <- which(is.na(y_cmp))

m_prd2_bru <- spde_bru_fit_prd2$summary.fitted.values$mean[idx_prd]
```

To improve visualization, we will plot the posterior means
on `graph_bkp`:
```{r, fig.show='hold',fig.align = "center",echo=TRUE, fig.width=7, warning=FALSE}
graph_bkp$plot(X = m_prd2_bru, X_loc = obs_loc_prd)
```

We will now plot the predictions together with
the data. To this end we need to create a new 
vector `y` which is a concatenation of `y`
and `m_prd`, then use it as the vector of 
responses of `graph`, which is done by using the
`add_responses()` method. Then, we plot.

Let us begin by concatenating the vectors and adding
the responses to `graph`:
```{r}
y_new2 <- c(y, m_prd2)
graph_bru_prd$add_responses(y_new2)
```

Now, we plot:
```{r, fig.show='hold',fig.align = "center",echo=TRUE, fig.width=7, warning=FALSE}
graph_bru_prd$plot(data=TRUE)
```

## Fitting `inlabru` models with replicates

We will now illustrate how to use our `inlabru` implementation to 
fit models with replicates.

To simplify exposition, we will use the same base graph.
So, we begin by creating a copy:
```{r}
graph_bru_rep <- graph_bkp$clone()
```

We will use the same observation locations and simulated
data from our illustration in `R-INLA`'s
implementation. Recall that we have 30 replicates, which
is given in the variable `n_repl`.

Recall that the response variables are given in `y_rep`,
which is a matrix with 30 columns, one per replicate.

Since we are considering a model with replicates, we will
add only the first replicate to the graph object:
```{r, fig.show='hold',fig.align = "center",fig.width=7,echo=TRUE}
graph_bru_rep$add_observations2(y_rep[,1], 
                    PtE = obs_loc, normalized=TRUE)
graph_bru_rep$observation_to_vertex()
graph_bru_rep$plot(data=TRUE)
```

Let us now create the model object:

```{r}
spde_model_bru_rep <- gpgraph_spde(graph_bru_rep)
```

Let us now create a new data list, where we provide
the response variables (as a vector) and the 
observation locations (for our current implementation,
the observation locations must be same throughout
the replicates):
```{r}
obs_loc_rep <- obs_loc
for(i in 2:n_repl){
    obs_loc_rep <- rbind(obs_loc_rep, obs_loc)
}   
data_list_rep <- list(y = as.vector(y_rep),
                            loc = obs_loc_rep)
```

To create the component, we need to create a vector
of replicate indexes. Recall that the number of 
observations is stored in the variable `n_obs_rep`:
```{r}
repl <- rep(1:n_repl, each=n_obs_rep)
```

Now, we can create the component, setting the argument
`replicate`, of the component `field`, to our vector `repl`:
```{r}
cmp_rep <-
    y ~ -1 + Intercept(1) + field(loc, 
                        model = spde_model_bru_rep,
                        replicate = repl)
```

Now, we fit the model:
```{r}
spde_bru_fit_rep <-
    bru(cmp_rep,
        data=data_list_rep)
```

Let us see the estimated values in the original scale:
```{r}
spde_result_bru_rep <- spde_metric_graph_result(spde_bru_fit_rep, 
                        "field", spde_model_bru_rep)

summary(spde_result_bru_rep)
```

Let us compare with the true values:

```{r}
  result_df_bru_rep <- data.frame(
    parameter = c("std.dev", "range"),
    true = c(sigma_rep, r_rep),
    mean = c(
      spde_result_bru_rep$summary.sigma$mean,
      spde_result_bru_rep$summary.range$mean
    ),
    mode = c(
      spde_result_bru_rep$summary.sigma$mode,
      spde_result_bru_rep$summary.range$mode
    )
  )
  print(result_df_bru_rep)
```