---
title: "INLA interface for Whittle--Matérn fields on metric graphs"
author: "David Bolin, Alexandre B. Simas, and Jonas Wallin"
date: "Created: 2022-11-23. Last modified: `r Sys.Date()`."
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{INLA interface for Whittle--Matérn fields on metric graphs}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
references:
- id: BSW2022a
  title: "Gaussian Whittle--Matérn fields on metric graphs"
  author:
  - family: Bolin
    given: David
  - family: Simas
    given: Alexandre B.
  - family: Wallin
    given: Jonas
  container-title: Bernoulli
  type: article
  issued:
    year: 2023
- id: BSW2022b
  title: "Statistical properties of Gaussian Whittle--Matérn fields on metric graphs"
  author:
  - family: Bolin
    given: David
  - family: Simas
    given: Alexandre B.
  - family: Wallin
    given: Jonas
  container-title: arXiv:2304.10372
  type: preprint
  issued:
    year: 2023
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
set.seed(1)
library(MetricGraph)
```

# Introduction

In this vignette we will present our `R-INLA`
interface to Whittle--Matérn fields. The underlying theory
for this approach is provided in 
[@BSW2022a](https://arxiv.org/abs/2205.06163) and [@BSW2022b](https://arxiv.org/abs/2304.10372). 

For an introduction to the `metric_graph` class, please see 
the [Working with metric graphs](metric_graph.html) vignette.

For a simplification of the `R-INLA` interface, see
the [inlabru interface of Whittle--Matérn fields](inlabru_interface.html) vignette.

In the [Gaussian random fields on metric graphs](random_fields.html) vignette,
we introduce all the models in metric graphs contained in this package,
as well as, how to perform statistical tasks on these models, but without
the `R-INLA` or `inlabru` interfaces.

We will present our `R-INLA` interface to 
the Whittle-Matérn fields by providing a step-by-step illustration.

The Whittle--Matérn fields are specified as solutions to the stochastic
differential equation 
$$
  (\kappa^2 - \Delta)^{\alpha} \tau u = \mathcal{W}
$$
on the metric graph $\Gamma$. We can work with these models without
any approximations if the smoothness parameter $\alpha$ is an integer, and this 
is what we focus on in this vignette. For details on the case of a general 
smoothness parameter, see [Whittle--Matérn fields with general smoothness](fem_models.html). 

# A toy dataset 

Let us begin by loading the `MetricGraph` package
and creating a metric graph:

```{r}
library(MetricGraph)

edge1 <- rbind(c(0,0),c(1,0))
edge2 <- rbind(c(0,0),c(0,1))
edge3 <- rbind(c(0,1),c(-1,1))
theta <- seq(from=pi,to=3*pi/2,length.out = 20)
edge4 <- cbind(sin(theta),1+ cos(theta))
edges = list(edge1, edge2, edge3, edge4)
graph <- metric_graph$new(edges = edges)
```

Let us add 50 random locations in each edge where we will 
have observations:

```{r}
obs_per_edge <- 50
obs_loc <- NULL
for(i in 1:(graph$nE)) {
  obs_loc <- rbind(obs_loc,
                   cbind(rep(i,obs_per_edge), 
                   runif(obs_per_edge)))
}
```

We will now sample in these observation locations
and plot the latent field:

```{r, fig.show='hold',fig.align = "center",echo=TRUE}
sigma <- 2
alpha <- 1
nu <- alpha - 0.5
r <- 0.15 # r stands for range


u <- sample_spde(range = r, sigma = sigma, alpha = alpha,
                 graph = graph, PtE = obs_loc)
graph$plot(X = u, X_loc = obs_loc)
```

Let us now generate the observed responses, which we will
call `y`. We will also plot the observed responses on the
metric graph.

```{r, fig.show='hold',fig.align = "center",echo=TRUE}
n_obs <- length(u)
sigma.e <- 0.1

y <- u + sigma.e * rnorm(n_obs)
graph$plot(X = y, X_loc = obs_loc)
```

# `R-INLA` implementation

We are now in a position to fit the model with our
`R-INLA` implementation. To this end,
we need to add the observations to the graph, which we 
will do with the `add_observations()` method. 

```{r, fig.show='hold',fig.align = "center",echo=TRUE}
df_graph <- data.frame(y = y, edge_number = obs_loc[,1],
                        distance_on_edge = obs_loc[,2])
# Adding observations and turning them to vertices
graph$add_observations(data=df_graph, normalized=TRUE)
graph$plot(data="y")
```

Now, we load the `R-INLA` package and
create the `inla` model object with the
`graph_spde` function. By default we have
`alpha=1`. 

```{r, message=FALSE}
library(INLA)
spde_model <- graph_spde(graph)
```

Now, we need to create the index object with 
the `graph_spde_make_index()` function:

```{r}
spde_index <- graph_spde_make_index(name="field", graph_spde=spde_model)
```

Observe that, as in `R-INLA`, we need to name the field.

The next step is to create the observation matrix, to which
we use the `graph_spde_make_A()` function:

```{r}
A <- graph_spde_make_A(spde_model)
```

The remaining is standard in `R-INLA`. We create the
formula object and the
`inla.stack` object with the `inla.stack()` function. 
The data needs to be in the `graph` (by using the `add_observations()` 
method) and should be supplied to the stack by using the
`graph_spde_data()` function. 

```{r}
f.s <- y ~ -1 + Intercept + f(field, model = spde_model)

stk_dat <- inla.stack(data = graph_data_spde(spde_model), 
                        A = A, 
                        effects = c(
      spde_index,
      list(Intercept = 1)
    ))
```

Now, we use the `inla.stack.data()` function:
```{r}
data_stk <- inla.stack.data(stk_dat)
```

Finally, we fit the model:

```{r}
spde_fit <- inla(f.s, data = data_stk)
```

Let us now obtain the estimates in the original scale
by using the `spde_metric_graph_result()` function, then
taking a `summary()`:
      
```{r}
spde_result <- spde_metric_graph_result(spde_fit, "field", spde_model)

summary(spde_result)
```

We will now compare the means of the estimated values with
the true values:

```{r}
  result_df <- data.frame(
    parameter = c("std.dev", "range"),
    true = c(sigma, r),
    mean = c(
      spde_result$summary.sigma$mean,
      spde_result$summary.range$mean
    ),
    mode = c(
      spde_result$summary.sigma$mode,
      spde_result$summary.range$mode
    )
  )
  print(result_df)
```


We can also plot the posterior marginal densities with the 
help of the `gg_df()` function:
```{r, fig.show='hold',fig.align = "center",echo=TRUE,  warning=FALSE}
  posterior_df_fit <- gg_df(spde_result)

  library(ggplot2)

  ggplot(posterior_df_fit) + geom_line(aes(x = x, y = y)) + 
  facet_wrap(~parameter, scales = "free") + labs(y = "Density")
```


## Kriging with our INLA implementation

Let us begin by obtaining an evenly spaced mesh with respect to the base
graph:

```{r}
obs_per_edge_prd <- 50
graph$build_mesh(n = obs_per_edge_prd)
```

Let us plot the resulting graph:

```{r, fig.show='hold',fig.align = "center",echo=TRUE,  warning=FALSE}
graph$plot(mesh=TRUE)
```

We will now add the observations on the mesh
locations to the graph we fitted the
`R-INLA` model. To this end we will use the
`add_mesh_observations()` method. We will enter the response variables
as `NA`. We can get the number of mesh nodes by counting 
the number of rows of the `mesh$PtE` attribute.

```{r}
n_obs_mesh <- nrow(graph$mesh$PtE)
y_prd <- rep(NA, n_obs_mesh)
data_mesh <- data.frame(y = y_prd)
graph$add_mesh_observations(data = data_mesh)
```

We will now fit a new model with `R-INLA` with this new
graph that contains the prediction locations. To this end,
we create a new model object with the `graph_spde()` function:

```{r}
spde_model_prd <- graph_spde(graph)
```

Now, let us create a new index object and a new observation (`A`):
```{r}
spde_index_prd <- graph_spde_make_index(name="field", 
                            graph_spde=spde_model_prd)

A_prd <- graph_spde_make_A(graph_spde = spde_model_prd)
```

We will create a new vector of response variables,
concatenating `y` to `y_prd`, then create a new
formula object and the
`inla.stack` object:

```{r}
f_s_prd <- y ~ -1 + Intercept + f(field, model = spde_model_prd)

stk_dat_prd <- inla.stack(data = graph_data_spde(spde_model_prd), 
                        A = A_prd, 
                        effects = c(
      spde_index_prd,
      list(Intercept = 1)
    ))
```

Now, we use the `inla.stack.data()` function and fit the model:
```{r}
data_stk_prd <- inla.stack.data(stk_dat_prd)

spde_fit_prd <- inla(f_s_prd, data = data_stk_prd)
```

We will now extract the means at the prediction locations:
```{r}
idx_prd <- which(is.na(graph_data_spde(spde_model_prd)[["y"]]))

m_prd <- spde_fit_prd$summary.fitted.values$mean[idx_prd]
```

To improve visualization, we will plot the posterior means
using the `plot()` method:
```{r, fig.show='hold',fig.align = "center",echo=TRUE,  warning=FALSE}
graph$plot_function(m_prd, vertex_size = 0, edge_width = 2)
```

Finally, we can plot the predictions together with
the data:
```{r, fig.show='hold',fig.align = "center",echo=TRUE,  warning=FALSE, message=FALSE}
p <- graph$plot_function(X = m_prd, vertex_size = 0, edge_width = 1)
graph$plot(data="y", vertex_size = 0, data_size = 2, p = p, edge_width = 0)
```

## Fitting `R-INLA` models with replicates

We will now illustrate how to use our `R-INLA` implementation to 
fit models with replicates.

To simplify exposition, we will use the same base graph.
So, we begin by clearing the observations.
```{r}
graph$clear_observations()
```

We will use the same observation locations as for the
previous cases. Let us sample 30 replicates:
```{r}
sigma_rep <- 1.5
alpha_rep <- 1
nu_rep <- alpha_rep - 0.5
r_rep <- 0.2 # r stands for range
kappa_rep <- sqrt(8 * nu_rep) / r_rep

n_repl <- 30

u_rep <- sample_spde(range = r_rep, sigma = sigma_rep,
                 alpha = alpha_rep,
                 graph = graph, PtE = obs_loc,
                 nsim = n_repl)
```

Let us now generate the observed responses, which we will
call `y_rep`.

```{r, fig.show='hold',fig.align = "center",echo=TRUE}
n_obs_rep <- nrow(u_rep)
sigma_e <- 0.1

y_rep <- u_rep + sigma_e * matrix(rnorm(n_obs_rep * n_repl),
                                    ncol=n_repl)
```

The `sample_spde()` function returns a matrix in which each replicate
is a column. We need to stack the columns together and a column to
indicate the replicate:
```{r, fig.show='hold',fig.align = "center",echo=TRUE}
dl_graph <- lapply(1:ncol(y_rep), function(i){data.frame(y = y_rep[,i],
                                          edge_number = obs_loc[,1],
                                          distance_on_edge = obs_loc[,2],
                                          repl = i)})
dl_graph <- do.call(rbind, dl_graph)
```
We can now add the the observations by setting the `group`
argument to `repl`:
```{r}
graph$add_observations(data = dl_graph, normalized=TRUE, 
                            group = "repl",
                            edge_number = "edge_number",
                            distance_on_edge = "distance_on_edge")
```

By definition the `plot()` method plots the first replicate.
We can select the other replicates with the `group` argument.
See the [Working with metric graphs](metric_graphs.html)
for more details.

```{r}
graph$plot(data="y")
```

Let us plot another replicate:

```{r}
graph$plot(data="y", group=2)
```

Let us now create the model object:

```{r}
spde_model_rep <- graph_spde(graph)
```

Let us first consider a case in which we do not use all
replicates. Then, we consider the case in which we use all
replicates. 

Thus, let us assume we want only to consider replicates
1, 3, 5, 7 and 9. To this end,
we the index object by using the
`graph_spde_make_index()` function with the
argument `n.repl` set to the number of replicates we want,
in this case 5:
```{r}
spde_index_rep <- graph_spde_make_index(name="field", 
                    graph_spde=spde_model_rep, 
                    n.repl = 5)
```

Similarly, we create the observation (`A`) matrix with the
function `graph_spde_make_A()` setting the `repl` argument
to the replicates we want:
```{r}
A_rep <- graph_spde_make_A(graph_spde=spde_model_rep,
                             repl = c(1,3,5,7,9))
```

Next, we create the stack object, remembering that
we need to input the `data` argument
with the function `graph_data_spde()`. There, we 
determine the replicates we want with the `repl` argument.
```{r}
stk_dat_rep <- inla.stack(data = graph_data_spde(spde_model_rep,
                                  repl = c(1,3,5,7,9)), 
                        A = A_rep, 
                        effects = c(
      spde_index_rep,
      list(Intercept = 1)
    ))
```

We now create the formula object, adding the
name of the field (in our case `field`)
attached with `.repl` a the `replicate` argument
inside the `f()` function.
```{r}
f_s_rep <- y ~ -1 + Intercept + 
    f(field, model = spde_model_rep, 
        replicate = field.repl)
```
Then, we create the stack object with The
`inla.stack.data()` function:
```{r}
data_stk_rep <- inla.stack.data(stk_dat_rep)
```
Now, we fit the model:
```{r}
spde_fit_rep <- inla(f_s_rep, data = data_stk_rep)
```

Let us see the estimated values in the original scale:
```{r}
spde_result_rep <- spde_metric_graph_result(spde_fit_rep, 
                        "field", spde_model_rep)

summary(spde_result_rep)
```

Let us compare with the true values:

```{r}
  result_df_rep <- data.frame(
    parameter = c("std.dev", "range"),
    true = c(sigma_rep, r_rep),
    mean = c(
      spde_result_rep$summary.sigma$mean,
      spde_result_rep$summary.range$mean
    ),
    mode = c(
      spde_result_rep$summary.sigma$mode,
      spde_result_rep$summary.range$mode
    )
  )
  print(result_df_rep)
```


Now, let us consider the case with all replicates. 
We create a new index object by using the
`graph_spde_make_index()` function with the
argument `n.repl` set to the total number of replicates:
```{r}
spde_index_rep <- graph_spde_make_index(name="field", 
                    graph_spde=spde_model_rep, 
                    n.repl = n_repl)
```

Similarly, we create a new observation (`A`) matrix with the
function `graph_spde_make_A()` setting the `repl` argument
to `__all`:
```{r}
A_rep <- graph_spde_make_A(graph_spde=spde_model_rep,
                            repl="__all")
```

Now the stack, where we set the `repl` argument in 
the function `graph_data_spde()` to `__all`:
```{r}
stk_dat_rep <- inla.stack(data = graph_data_spde(spde_model_rep,
                                          repl = "__all"), 
                        A = A_rep, 
                        effects = c(
      spde_index_rep,
      list(Intercept = 1)
    ))
```

We now create the formula object in the same way as before:
```{r}
f_s_rep <- y ~ -1 + Intercept + 
    f(field, model = spde_model_rep, 
        replicate = field.repl)
```
Then, we create the stack object with The
`inla.stack.data()` function:
```{r}
data_stk_rep <- inla.stack.data(stk_dat_rep)
```
Now, we fit the model:
```{r}
spde_fit_rep <- inla(f_s_rep, data = data_stk_rep)
```

Let us see the estimated values in the original scale:
```{r}
spde_result_rep <- spde_metric_graph_result(spde_fit_rep, 
                        "field", spde_model_rep)

summary(spde_result_rep)
```

Let us compare with the true values:

```{r}
  result_df_rep <- data.frame(
    parameter = c("std.dev", "range"),
    true = c(sigma_rep, r_rep),
    mean = c(
      spde_result_rep$summary.sigma$mean,
      spde_result_rep$summary.range$mean
    ),
    mode = c(
      spde_result_rep$summary.sigma$mode,
      spde_result_rep$summary.range$mode
    )
  )
  print(result_df_rep)
```

