---
title: "inlabru interface of Whittle--Matérn fields"
author: "David Bolin, Alexandre B. Simas, and Jonas Wallin"
date: "Created: 2022-11-23. Last modified: `r Sys.Date()`."
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{inlabru interface of Whittle--Matérn fields}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
references:
- id: BSW2022a
  title: "Gaussian Whittle--Matérn fields on metric graphs"
  author:
  - family: Bolin
    given: David
  - family: Simas
    given: Alexandre B.
  - family: Wallin
    given: Jonas
  container-title: Bernoulli
  type: article
  issued:
    year: 2023
- id: BSW2022b
  title: "Statistical properties of Gaussian Whittle--Matérn fields on metric graphs"
  author:
  - family: Bolin
    given: David
  - family: Simas
    given: Alexandre B.
  - family: Wallin
    given: Jonas
  container-title: arXiv:2304.10372
  type: preprint
  issued:
    year: 2023
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
set.seed(1)
library(MetricGraph)
```

# Introduction

In this vignette we will present our `inlabru`
interface to Whittle--Matérn fields. The underlying theory
for this approach is provided in 
[@BSW2022a](https://arxiv.org/abs/2205.06163) and [@BSW2022b](https://arxiv.org/abs/2304.10372). 

For an introduction to the `metric_graph` class, please see 
the [Working with metric graphs](metric_graph.html) vignette.

For our `R-INLA` interface, see
the [INLA interface of Whittle--Matérn fields](inla_interface.html) vignette.

In the [Gaussian random fields on metric graphs](random_fields.html) vignette,
we introduce all the models in metric graphs contained in this package,
as well as, how to perform statistical tasks on these models, but without
the `R-INLA` or `inlabru` interfaces.

We will present our `inlabru` interface to 
the Whittle-Matérn fields by providing a step-by-step illustration.

The Whittle--Matérn fields are specified as solutions to the stochastic
differential equation 
$$
  (\kappa^2 - \Delta)^{\alpha} \tau u = \mathcal{W}
$$
on the metric graph $\Gamma$. We can work with these models without
any approximations if the smoothness parameter $\alpha$ is an integer, and this 
is what we focus on in this vignette. For details on the case of a general 
smoothness parameter, see [Whittle--Matérn fields with general smoothness](fem_models.html). 

# A toy dataset 

Let us begin by loading the `MetricGraph` package
and creating a metric graph:

```{r}
library(MetricGraph)

edge1 <- rbind(c(0,0),c(1,0))
edge2 <- rbind(c(0,0),c(0,1))
edge3 <- rbind(c(0,1),c(-1,1))
theta <- seq(from=pi,to=3*pi/2,length.out = 20)
edge4 <- cbind(sin(theta),1+ cos(theta))
edges = list(edge1, edge2, edge3, edge4)
graph_bru <- metric_graph$new(edges = edges)
```

Let us add 50 random locations in each edge where we will 
have observations:

```{r}
obs_per_edge <- 50
obs_loc <- NULL
for(i in 1:(graph_bru$nE)) {
  obs_loc <- rbind(obs_loc,
                   cbind(rep(i,obs_per_edge), 
                   runif(obs_per_edge)))
}
```

We will now sample in these observation locations
and plot the latent field:

```{r, fig.show='hold',fig.align = "center",echo=TRUE}
sigma <- 2
alpha <- 1
nu <- alpha - 0.5
r <- 0.15 # r stands for range

u <- sample_spde(range = r, sigma = sigma, alpha = alpha,
                 graph = graph_bru, PtE = obs_loc)
graph_bru$plot(X = u, X_loc = obs_loc)
```

Let us now generate the observed responses, which we will
call `y`. We will also plot the observed responses on the
metric graph.

```{r, fig.show='hold',fig.align = "center",echo=TRUE}
n_obs <- length(u)
sigma.e <- 0.1

y <- u + sigma.e * rnorm(n_obs)
graph_bru$plot(X = y, X_loc = obs_loc)
```


# `inlabru` implementation

We will now present our `inlabru` implementation
of the Whittle-Matérn fields for metric graphs.
It has the advantage, over our `R-INLA`
implementation, of not requiring the user
to provide observation matrices, indices nor 
stack objects.

We are now in a position to fit the model with our
`inlabru` implementation. To this end,
we need to add the observations to the graph, which we 
will do with the `add_observations()` method. 

```{r, fig.show='hold',fig.align = "center",echo=TRUE}
# Creating the data frame
df_graph <- data.frame(y = y, edge_number = obs_loc[,1],
                      distance_on_edge = obs_loc[,2])
# Adding observations and turning them to vertices
graph_bru$add_observations(data = df_graph, normalized=TRUE)
graph_bru$plot(data="y")
```

Now, we load `INLA` and `inlabru` packages. We will also
need to create the `inla` model object with the
`graph_spde` function. By default we have
`alpha=1`. 

```{r, message=FALSE}
library(INLA)
library(inlabru)
spde_model_bru <- graph_spde(graph_bru)
```

Now, we create `inlabru`'s component, which is a 
formula-like object. The index parameter in `inlabru`
is not used in our implementation, thus, we replace it 
by the `repl` argument, which tells which 
replicates to use. If there is no replicates, we
supply `NULL`.
```{r}
cmp <-
    y ~ -1 + Intercept(1) + field(loc,
                    model = spde_model_bru)
```

Now, we create the data object to be passed to the `bru()` function:

```{r}
data_spde_bru <- graph_data_spde(spde_model_bru, loc = "loc")
```

we directly fit the model by providing the `data` component of the `data_spde_bru` list:

```{r}
spde_bru_fit <-
    bru(cmp, data=data_spde_bru[["data"]])
```

Let us now obtain the estimates in the original scale
by using the `spde_metric_graph_result()` function, then
taking a `summary()`:

```{r}
spde_bru_result <- spde_metric_graph_result(spde_bru_fit, 
                    "field", spde_model_bru)

summary(spde_bru_result)
```

We will now compare the means of the estimated values with
the true values:

```{r}
  result_df_bru <- data.frame(
    parameter = c("std.dev", "range"),
    true = c(sigma, r),
    mean = c(
      spde_bru_result$summary.sigma$mean,
      spde_bru_result$summary.range$mean
    ),
    mode = c(
      spde_bru_result$summary.sigma$mode,
      spde_bru_result$summary.range$mode
    )
  )
  print(result_df_bru)
```

We can also plot the posterior marginal densities with the 
help of the `gg_df()` function:
```{r, fig.show='hold',fig.align = "center",echo=TRUE,  warning=FALSE}
  posterior_df_bru_fit <- gg_df(spde_bru_result)

  library(ggplot2)

  ggplot(posterior_df_bru_fit) + geom_line(aes(x = x, y = y)) + 
  facet_wrap(~parameter, scales = "free") + labs(y = "Density")
```

## Kriging with the `inlabru` implementation

Unfortunately, our `inlabru` implementation is not 
compatible with `inlabru`'s `predict()` method. This has
to do with the nature of the metric graph's object.

To this end, we have provided a different `predict()` method.
We will now show
how to do kriging with the help of this function. 

We begin by creating a data list
with the positions we want the predictions. In this case,
we will want the predictions on a mesh.

Let us begin by obtaining an evenly spaced mesh with respect to the base
graph:

```{r}
obs_per_edge_prd <- 50
graph_bru$build_mesh(n = obs_per_edge_prd)
```

Let us plot the resulting graph:

```{r, fig.show='hold',fig.align = "center",echo=TRUE,  warning=FALSE}
graph_bru$plot(mesh=TRUE)
```

The positions we want are the mesh positions, which
can be obtained by using the `get_mesh_locations()` method.
We also set `bru=TRUE` and `loc="loc"` to obtain a data list suitable to be used with `inlabru`.
```{r}
data_list <- graph_bru$get_mesh_locations(bru = TRUE,
                                            loc = "loc")
```

We can now obtain the predictions by using the `predict()` method. Observe
that our `predict()` method for graph models is a bit different from
`inlabru`'s standard `predict()` method. Indeed, the first argument is
the model created with the `graph_spde()` function, the second is
`inlabru`'s component, and the remaining is as done with the standard `predict()`
method in `inlabru`.

```{r}
field_pred <- predict(spde_model_bru, 
                                cmp,
                                spde_bru_fit, 
                                newdata = data_list,
                                formula = ~field)
```

Finally, we can plot the predictions together with
the data:
```{r, fig.show='hold',fig.align = "center",echo=TRUE,  warning=FALSE, message=FALSE}
plot(field_pred)
```

We can also obtain a 3d plot by setting `plotly` to `TRUE`:

```{r, fig.show='hold',fig.align = "center",echo=TRUE,  warning=FALSE, message=FALSE}
plot(field_pred, plotly = TRUE)
```

## Fitting `inlabru` models with replicates

We will now illustrate how to use our `inlabru` implementation to 
fit models with replicates.

To simplify exposition, we will use the same base graph.
So, we begin by clearing the observations:
```{r}
graph_bru$clear_observations()
```

We will use the same observation locations as for the
previous cases. Let us sample 30 replicates:
```{r}
sigma_rep <- 1.5
alpha_rep <- 1
nu_rep <- alpha_rep - 0.5
r_rep <- 0.2 # r stands for range

n_repl <- 30

u_rep <- sample_spde(range = r_rep, sigma = sigma_rep,
                 alpha = alpha_rep,
                 graph = graph_bru, PtE = obs_loc,
                 nsim = n_repl)
```

Let us now generate the observed responses, which we will
call `y_rep`.

```{r, fig.show='hold',fig.align = "center",echo=TRUE}
n_obs_rep <- nrow(u_rep)
sigma_e <- 0.1

y_rep <- u_rep + sigma_e * matrix(rnorm(n_obs_rep * n_repl),
                                    ncol=n_repl)
```


We can now add the the observations by setting the `group`
argument to `repl`:
```{r, fig.show='hold',fig.align = "center",echo=TRUE}
dl_rep_graph <- lapply(1:ncol(y_rep), function(i){data.frame(y = y_rep[,i],
                                          edge_number = obs_loc[,1],
                                          distance_on_edge = obs_loc[,2],
                                          repl = i)})
dl_rep_graph <- do.call(rbind, dl_rep_graph)

graph_bru$add_observations(data = dl_rep_graph, normalized=TRUE,
                                    group = "repl")
```

By definition the `plot()` method plots the first replicate.
We can select the other replicates with the `group` argument.
See the [Working with metric graphs](metric_graphs.html)
for more details.

```{r}
graph_bru$plot(data="y")
```

Let us plot another replicate:

```{r}
graph_bru$plot(data="y", group=2)
```


Let us now create the model object:

```{r}
spde_model_bru_rep <- graph_spde(graph_bru)
```

Let us first create a model using the replicates
1, 3, 5, 7 and 9. To this end, we provide the vector
of the replicates we want as the `input` argument to the `field`.
The `graph_data_spde()` acts as a helper function when building this vector. All we need to do, is to use the `repl` component of the list created when using the `graph_data_spde()`
```{r}
data_spde_bru <- graph_data_spde(spde_model_bru_rep, loc = "loc",
        repl=c(1,3,5,7,9))

repl <- data_spde_bru[["repl"]]
cmp_rep <-
    y ~ -1 + Intercept(1) + field(loc, 
                        model = spde_model_bru_rep,
                        replicate = repl)
```

Now, we fit the model:
```{r}
spde_bru_fit_rep <-
    bru(cmp_rep,
        data=data_spde_bru[["data"]])
```

Let us see the estimated values in the original scale:
```{r}
spde_result_bru_rep <- spde_metric_graph_result(spde_bru_fit_rep, 
                        "field", spde_model_bru_rep)

summary(spde_result_bru_rep)
```

Let us compare with the true values:

```{r}
  result_df_bru_rep <- data.frame(
    parameter = c("std.dev", "range"),
    true = c(sigma_rep, r_rep),
    mean = c(
      spde_result_bru_rep$summary.sigma$mean,
      spde_result_bru_rep$summary.range$mean
    ),
    mode = c(
      spde_result_bru_rep$summary.sigma$mode,
      spde_result_bru_rep$summary.range$mode
    )
  )
  print(result_df_bru_rep)
```


We will now show how to fit the model considering all replicates.
To this end, we simply set the `repl` argument in `graph_data_spde()`
function to `__all`.

```{r}
data_spde_bru_rep <- graph_data_spde(spde_model_bru_rep, loc = "loc",
        repl="__all")

repl <- data_spde_bru_rep[["repl"]]

cmp_rep <-  y ~ -1 + Intercept(1) + field(loc, 
                        model = spde_model_bru_rep,
                        replicate = repl)
```

Similarly, we fit the model, by setting the `repl` argument 
to "__all" inside the
`graph_data_spde()` function:
```{r}
spde_bru_fit_rep <-
    bru(cmp_rep,
        data=data_spde_bru_rep[["data"]])
```

Let us see the estimated values in the original scale:
```{r}
spde_result_bru_rep <- spde_metric_graph_result(spde_bru_fit_rep, 
                        "field", spde_model_bru_rep)

summary(spde_result_bru_rep)
```

Let us compare with the true values:

```{r}
  result_df_bru_rep <- data.frame(
    parameter = c("std.dev", "range"),
    true = c(sigma_rep, r_rep),
    mean = c(
      spde_result_bru_rep$summary.sigma$mean,
      spde_result_bru_rep$summary.range$mean
    ),
    mode = c(
      spde_result_bru_rep$summary.sigma$mode,
      spde_result_bru_rep$summary.range$mode
    )
  )
  print(result_df_bru_rep)
```

