---
title: "On isotropic covariances on metric graphs with non-Euclidean edges"
author: "David Bolin, Alexandre B. Simas, and Jonas Wallin"
date: "Created: 2024-02-11. Last modified: `r Sys.Date()`."
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{On isotropic covariances on metric graphs with non-Euclidean edges}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
  - id: Anderes2020
  title: "Isotropic covariance functions on graphs and their edges"
  author:
  - family: Anderes
    given: Ethan
  - family: Møller
    given: Jesper
  - family: Rasmussen
    given: Jakob G
  container-title: Annals of Statistics
  type: article
  issue: 48
  pages: 2478-2503
  issued:
    year: 2020
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```


# Introduction

In this vignette we will investigate the behavior of isotropic covariance functions (with respect to the resistance metric) on metric graphs with non-Euclidean edges. This boils down to studying the resistance metric on metric graphs with non-Euclidean edges.

More precisely, we will show that when dealing with metric graphs with non-Euclidean edges, the addition or removal of vertices of degree 2 can have a severe impact on the resistance metric and, a fortiori, on the covariance function. 

The effect of this study is that if one fits an isotropic model on a metric graph with non-Euclidean edges by turning the observations into vertices, then the resulting model is different from what one would expect. Indeed, the model might change when doing predictions (as one would need to add further observations as vertices to obtain predictions). Furthermore, if one does not add observations as vertices, then the actual covariance function might not be valid  [@Anderes2020](https://projecteuclid.org/journals/annals-of-statistics/volume-48/issue-4/Isotropic-covariance-functions-on-graphs-and-their-edges/10.1214/19-AOS1896.full). 

# A simple example

We start by creating a metric graph with multiple edges. This metric graph does not have Euclidean edges:

```{r}
edge1 <- rbind(c(0,0),c(1,0))
theta1 <- seq(from=pi/2,to=0,length.out = 50)
theta2 <- seq(from=pi,to=3*pi/2,length.out = 50)
edge2 <- cbind(sin(theta1)-1,cos(theta1))
edge3 <- cbind(sin(theta2),1+ cos(theta2))
edges = list(edge1, edge2, edge3)

graph <- metric_graph$new(edges = edges)

graph$plot()
```

One can also check if a metric graph has Euclidean edges by calling the `check_euclidean()` method, then calling the `summary()`:

```{r}
graph$check_euclidean()
summary(graph)
```

Let us now compute the resistance distances between the vertices using the `compute_resdist()` method:

```{r}
graph$compute_resdist()
```

We can now observe the resistance distances in the `res_dist` element of `graph`:

```{r}
res_dist_graph <- graph$res_dist[[1]]
```

In reality, what is being actually done to the metric graph, is that one of the edges is, in practice, being ignored. Indeed, let us create a metric graph with Euclidean edges obtained by removing one of the edges, and compute the resistance distance between its vertices. Thus, we start by creating such a metric graph:

```{r}
edge1 <- rbind(c(0,0),c(1,0))
theta1 <- seq(from=pi/2,to=0,length.out = 50)
edge2 <- cbind(sin(theta1)-1,cos(theta1))
edges2 = list(edge1, edge2)

graph2 <- metric_graph$new(edges = edges2)

graph2$plot()
```

Observe that it has Euclidean edges:

```{r}
graph2$check_euclidean()
summary(graph2)
```

Let us now compute the resistance distances:

```{r}
graph2$compute_resdist()
```

Now, let us compare the resistance distances:

```{r}
identical(res_dist_graph, graph2$res_dist[[1]])
```

Therefore, indeed, one of the edges is being ignored in practice. 

Let us now see what happens when we have observations on both edges, and compute the resistance distances on the resulting metric graph (in which the observations were turned into vertices).

To this end, let us create some observations and add them to `graph`:

```{r}
df_data <- data.frame(edge_number = c(1,2,3), distance_on_edge = c(0.5,0.5,0.5), y = c(1,2,3))

graph$add_observations(data = df_data, normalized = TRUE)

graph$plot(data="y")
```

Let us now compute the resistance distances between the original vertices. To this end we will use the `compute_resdist()` method with the arguments `full=TRUE` to compute the resistance distance at all available locations, and `include_vertices=TRUE` to include the original vertices of the metric graph in the result. 

```{r}
graph$compute_resdist(full=TRUE, include_vertices = TRUE)
```

Since we have 3 vertices, the resistance distances between the original vertices are given by the elements 1 to 3 of the resulting matrix:

```{r}
res_dist_orig <- graph$res_dist[[1]][1:3,1:3]
res_dist_orig
```

We can observe that the resistance distances are very different. Thus, the impact of adding observations when computing resistance distances on non-Euclidean metric graphs is severe. 

In particular, the impact on the covariance function is severe. For example, let us consider an exponential covariance function with `sigma=1` and `kappa=1`:

```{r}
exp_covariance(res_dist_graph, theta = c(1,1))
```
and
```{r}
exp_covariance(res_dist_orig, theta = c(1,1))
```

This shows that these are two different models. One considering the original metric graph, and another, considering the metric graph with the added observations. Therefore, when one considers a model with isotropic covariance function on a metric graph with non-Euclidean edges to perform statistical analysis, different models are considered in the different stages of the statistical analysis. More precisely, one believes the model being considered only depends on the original vertices of the metric graph, when, in reality, when fitting the model, the resistance distances are computed on a different metric graph that was obtained when the observations were turned into vertices of a new metric graph. Finally, when one does prediction, yet another metric graph will be considered, namely, the metric graph in which the prediction locations were added as vertices of yet another metric graph.

Thus, we do not recommend using isotropic covariance models on metric graphs with non-Euclidean edges, even if the model is capable of being fit, as there will be several inconsistencies between the original model, the fitted model, and the model used to obtain predictions.