---
title: "MetricGraph: Random Fields on Metric Graphs"
author: "David Bolin, Alexandre B. Simas, and Jonas Wallin"
date: "`r Sys.Date()`"
output:
  rmarkdown::html_vignette:
    number_sections: true
vignette: >
  %\VignetteIndexEntry{MetricGraph: Random Fields on Metric Graphs}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
references:
- id: BSW2022a
  title: "Gaussian Whittle--Matérn fields on metric graphs"
  author:
  - family: Bolin
    given: David
  - family: Simas
    given: Alexandre B.
  - family: Wallin
    given: Jonas
  container-title: arXiv:2205.06163
  type: preprint
  issued:
    year: 2022
- id: BSW2022b
  title: "Statistical properties of Gaussian Whittle--Matérn fields on metric graphs"
  author:
  - family: Bolin
    given: David
  - family: Simas
    given: Alexandre B.
  - family: Wallin
    given: Jonas
  container-title: arXiv:??
  type: preprint
  issued:
    year: 2022
- id: Anderes2020
  title: "Isotropic covariance functions on graphs and their edges"
  author:
  - family: Anderes
    given: Ethan
  - family: Møller
    given: Jesper
  - family: Rasmussen
    given: Jakob G
  container-title: Annals of Statistics
  type: article
  issue: 48
  pages: 2478-2503
  issued:
    year: 2020
- id: Borovitskiy2021
  title: "Matérn Gaussian processes on graphs"
  author:
  - family: Borovitskiy
    given: Viacheslav
  - family: Azangulov
    given: Iskander
  - family: Terenin
    given: Alexander 
  - family: Mostowsky
    given: Peter
  - family: Deisenroth
    given: Marc
  - family: Durrande
    given: Nicolas
  container-title: International Conference on Artificial Intelligence and Statistics
  type: article
  pages: 2593-2601
  issued: 
    year: 2021
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(MetricGraph)
library(sp)
library(osmdata)
set.seed(1)
```

# Introduction

There has lately been much interest in statistical modeling of data on compact metric graphs such as street or river networks based on Gaussian random fields.


The `R` package `MetricGraph` contains functions for working with data and 
random fields on compact metric graphs. The main functionality is contained in 
the `metric_graph` class, which is used for specifying metric graphs, adding data to them, visualization, and other basic functions that are needed for working with data and random fields on metric graphs. The package also implements three types
of Gaussian processes on metric graphs: The Whittle--Matérn fields introduced 
by [@BSW2022a](https://arxiv.org/abs/2205.06163) and [@BSW2022b](??), Gaussian
processes with isotropic covariance functions [@Anderes2020](https://projecteuclid.org/journals/annals-of-statistics/volume-48/issue-4/Isotropic-covariance-functions-on-graphs-and-their-edges/10.1214/19-AOS1896.full), and Gaussian models based on the graph Laplacian [@Borovitskiy2021](http://proceedings.mlr.press/v130/borovitskiy21a/borovitskiy21a.pdf). 

Basic statistical tasks such as likelihood evaluation and prediction is implemented for these three types of models in `MetricGraph`. Further, the package also contains interfaces to `R-INLA` (REF, package available from [http://R-INLA.
org/download/](http://R-INLA.org/download/) and `inlabru` (REF) that facilitates using those packages for full Bayesian inference of general Latent Gaussian Models (LGMs) that includes Whittle--Matérn fields on metric graphs. 

The package is available to install from the Comprehensive R Archive
Network (CRAN) at \url{https://CRAN.R-project.org/package=MetricGraph}. A development version of the package is available via the repository [https://github.com/davidbolin/MetricGraph](https://github.com/davidbolin/MetricGraph). The development version is updated more frequently, and can easily be installed directly in R as described on the repository homepage. 

The following sections describe the main functionality of the package and summarizes some of the required theory. Section 2 introduces metric graphs and the `metric_graph` class, Section 3 shows how to work with random fields on metric graphs, and Section 4 introduces the `inlabru` interface of the package through an application to real data. For a complete introduction to the functionality of the package, we refer to the Vignettes available on the package homepage [https://davidbolin.github.io/MetricGraph/](https://davidbolin.github.io/MetricGraph/). In particular, that contains an introduction to the `INLA` interface, the implementation of Whittle--Matérn fields with general smoothness, and further details and examples for all methods in the package. 


# The `metric_graph` class

A compact metric graph $\Gamma$ consists of a set of finitely many vertices $\mathcal{V}=\{v_i\}$ and a finite set $\mathcal{E}=\{e_j\}$ of edges connecting the vertices. Each edge $e$ is defined by a pair of vertices $(v_i,v_k)$ and a finite length $l_e \in (0,\infty)$. An edge in the graph is a curve parametrized by arc-length, and a location $s\in \Gamma$ is a position on an edge, and can thus be represented as a touple $(e,t)$ where $t\in[0,l_e]$. 

## Basic constructions 	
A metric graph is represented in the `MetricGraph` package through the class `metric_graph`. An object of this class can be constructed in two ways. The first is to specify the vertex matrix `V` and the edge matrix `E`, and it is then assumed that all edges are straight lines. The second, more flexible, option is to specify the object from a `SpatialLines` object using the `sp` package (REF). 

To illustrate this, we use the `osmdata` package to download data from OpenStreetMap. In the following code, we extract the streets on the campus of 
King Abdullah University of Science and Technology (KAUST) as a `SpatialLines` object:

```{r}
call <- opq(bbox = c(39.0884, 22.33, 39.115, 22.3056))
call <- add_osm_feature(call, key = "highway",value=c("motorway",
                                                      "primary","secondary",
                                                      "tertiary",
                                                      "residential"))
data <- osmdata_sp(call)
lines <- SpatialLines(data$osm_lines@lines)
```

We can now create the metric graph as follows. In the command we set the argument `longlat = TRUE` since the coordinates of the vertices are given in Longitude and Latitude.
```{r, fig.show='hold',fig.align = "center",fig.width=7,echo=TRUE}
graph <- metric_graph$new(lines, longlat = TRUE)
graph$plot(vertex_size = 0.5)
```

We can note the warning that the graph is not connected, so let us create a 
`graph_components` object that contains all connected components as graphs and then extract the largest connected component to work with
```{r, fig.show='hold',fig.align = "center",fig.width=7,echo=TRUE}
graphs <- graph_components$new(lines = lines, longlat = TRUE)
graph <- graphs$get_largest()
graph$plot(vertex_size = 0)
```

The `graph` object now contains all important features of the graph, such as the vertex matrix `graph$V`, the number of vertices `graph$nV`, the edge matrix `graph$E`, the number of edges  `graph$nE`, and the vector of all edge lengths `graph$edge_lengths` given in the unit km (since we specified `longlat = TRUE` in the construction). Thus, we can obtain the range of the edge lengths in the unit m as:
```{r}
range(graph$edge_lengths) * 1000
```

We will also remove the vertices of degree 2 by using the `prune_vertices()` method:

```{r, fig.align = "center",fig.width=7,echo=TRUE}
graph$prune_vertices()
graph$plot()
```

## Understanding coordinates on graphs
The locations of the vertices are specified in Euclidean coordinates. However, 
when specifying a position on the graph, it is not practical to work with 
Euclidean coordinates since not all locations in Euclidean space are locations
on the graph. It is instead better to specify a location on the graph by the 
touple $(i, t)$, where $i$ denotes the number of the edge and $t$ is the 
location on the edge. The location $t$ can either be specified as the 
distance from the start of the edge (and then takes values between 0 and the 
length of the edge) or as the normalized distance from the start of the edge
(and then takes values between 0 and 1). The function `coordinates` can be used
to convert between coordinates in Euclidean space and locations on the graph. 
For example the location at normalized distance 0.2 from the start of the second edge is:
```{r}
graph$coordinates(PtE = matrix(c(2, 0.2), 1,2), normalized = TRUE)
```

The function can also be used to find the closest location on the graph for 
a location in Euclidean space: 
```{r}
graph$coordinates(XY = matrix(c(39.117, 22.319), 1,2))
```
In this case, the `normalized` argument decides whether the returned value 
should be given in normalized distance or not. 

## Adding data to the graph 

Given that we have constructed the metric graph, we can now add data to it. 
As an example, let us sample some locations on edges at random and add data to them in the graph: 
```{r, fig.show='hold',fig.align = "center",fig.width=7,echo=TRUE}
n.obs <- 10
data <- data.frame(edge_number = sample(1:graph$nE, n.obs),
                   distance_on_edge = runif(n.obs),
                   y = rnorm(n.obs))
graph$add_observations(data = data, normalized = TRUE)
graph$plot(data = "y", vertex_size = 0)
```

One should note here that one needs to specify `normalized = TRUE` in the function to specify that the locations are in normalized distance on the edges. If this command is not set, the distances are interpreted as not being normalized. The `add_observations()` function accepts multiple types of inputs. One scenario that can be common in applications is to have the data as `SpatialPoints` objects, and one can then add the observations as a `SpatialPointsDataFrame`. To illustrate this, let us again sample some locations at random on the graph, then use the function `coordinates` to transform those to Longitude Latitude coordinates, which we then use to create a `SpatialPointsDataFrame` object which we add to the graph:
```{r, fig.show='hold',fig.align = "center",fig.width=7,echo=TRUE}
obs.loc <- cbind(sample(1:graph$nE, n.obs), runif(n.obs))
obs.lonlat <- graph$coordinates(PtE = obs.loc, normalized = TRUE)
obs <- rnorm(n.obs)
points <- SpatialPointsDataFrame(coords = obs.lonlat,
                                 data = data.frame(y = obs))
graph$add_observations(points)
graph$plot(data = "y", vertex_size = 0)
```

If we want to replace the data in the object, we can use `clear_observations()`
to remove all current data.

## Working with functions on metric graphs

When working with data on metric graphs, one often wants to display functions
on the graph. The best way to visualize functions on the graph is to evaluate
them on a fine mesh over the graph and then use `plot_function`. To illustrate
this procedure, let us construct a mesh on the graph:
```{r, fig.show='hold',fig.align = "center",fig.width=7,echo=TRUE}
graph$build_mesh(h = 100/1000)
graph$plot(mesh=TRUE)
```

In the command `build_mesh`, the argument `h` decides the largest spacing between nodes in the mesh. Above we chose that as 100m which is a bit coarse. So let us reduce the value of `h` to 10m and rebuild the mesh:
```{r}
  graph$build_mesh(h = 10/1000)
```

Suppose now that we want to display the function $f(s) = \text{Longitude}(s) - \text{Latitude}(s)$ on this graph. We then first evaluate it on the vertices of the mesh and then use the  function `plot_function` to display it:
```{r, fig.show='hold',fig.align = "center",fig.width=7,echo=TRUE}
lon <- graph$mesh$V[, 1]
lat <- graph$mesh$V[, 2]
f <- lon - lat
graph$plot_function(f, vertex_size = 0, edge_width = 0.5)
```

Alternatively, we can set `plotly = TRUE` in the plot command to get a 3D 
visualization of the function by using the `plotly` (REF) package. 
When the first argument of `plot_function` is a vector, the function assumes that 
the values in the vector are the values of the function evaluated at the vertices
of the mesh. As an alternative, one can also provide the first argument as a 
matrix consisting of the triplets $(i, t, f(i, t))$, where $i$ denotes the 
edge number, $t$ the location on the edge, and $f(i, t)$ the value at that point.

# Random fields on metric graphs
Having defined the metric graph, we are now ready to specify Gaussian processes on it. In this section, we will briefly cover the three main types of Gaussian processes that are supported. Be begin by the main class of models, the Whittle--Matérn fields, then consider Gaussian processes with isotropic covariance functions, and finally look at discrete models based on the graph Laplacian.

## Whittle--Matérn fields 
The Gaussian Whittle--Matérn fields are specified as solutions to the stochastic
differential equation 
$$
  (\kappa^2 - \Delta)^{\alpha/2} \tau u = \mathcal{W}
$$
on the metric graph $\Gamma$. We can work with these models without
and approximations if the smoothness parameter $\alpha$ is an integer, and this 
is what we focus on in this vignette. For details on the case of a general 
smoothness parameter, see [Whittle--Matérn fields with general smoothness](fem_models.html). 

### Sampling 

As an example, let us simulate the field $u$ on the graph using $\alpha = 1$. 
To do so, we again draw some locations at random, then sample the field at these locations and plot the result
```{r, fig.align = "center",fig.width=7,echo=TRUE}
sigma <- 1.3
r <- 0.15 # range parameter
kappa <- sqrt(8*0.5)/r
sigma_e <- 0.1

n.obs <- 200
obs.loc <- cbind(sample(1:graph$nE, n.obs, replace=TRUE), runif(n.obs))

# The graph has very small edges, so we will use the 
# Q method to improve numerical stability
u <- sample_spde(kappa = kappa, sigma = sigma, alpha = 1,
                 graph = graph, PtE = obs.loc, method = "Q")
graph$clear_observations()
graph$add_observations(data = data.frame(edge_number = obs.loc[, 1],
                                         distance_on_edge = obs.loc[, 2],
                                         u = u),
                       normalized = TRUE)
graph$plot(data = "u", vertex_size = 0)
```

We can also sample the field at the mesh on the graph as follows:
```{r, fig.show='hold',fig.align = "center",fig.width=7,echo=TRUE}
u <- sample_spde(kappa = kappa, sigma = sigma, alpha = 1,
                 graph = graph, type = "mesh", method = "Q")
graph$plot_function(u, vertex_size = 0, edge_width = 0.5)
```

Since $\alpha=1$, these sample paths are continuous but not differentiable. 
To visualize the correlation structure of the field, we can compute and plot
the covariances between some point and all other points in the graph as follows:
```{r, fig.show='hold',fig.align = "center",fig.width=7,echo=TRUE}
C <- covariance_alpha1_mesh(c(200, 0.1), kappa = kappa, sigma = sigma, 
                            graph = graph)
graph$plot_function(C, vertex_size = 0, edge_width = 0.5)
```

To obtain a field with differentiable sample paths, we can change to $\alpha=2$ in the code above.


### Inference

Suppose that we have data of the form 
$$
y_i = \beta_1\text{lon}(s_i) +  \beta_1\text{lat}(s_i) +  u(s_i) + \varepsilon_i, \quad i=1,\ldots,n
$$
where $s_i\in \Gamma$ are observation locations, lon and lat are the longitude and latitude of these locations, $\beta_1,\beta_2$ are some regression coefficients, and 
$\varepsilon_i$ are independent centered Gaussian variables $N(0,\sigma_e^2)$ representing measurement noise. 

Let us create such observations, clear the current data in the graph, and finally add the data on the graph:
```{r, fig.show='hold',fig.align = "center",fig.width=7,echo=TRUE}
# The graph has very small edges, so we will use the 
# Q method to improve numerical stability

sigma <- 2
r <- 0.2 # range parameter
kappa <- sqrt(8*0.5)/r
sigma_e <- 0.1

n.obs.1 <- 400 # all edges
n.obs.2 <- 100 # long edges

n.obs <- n.obs.1 + n.obs.2

obs.loc <- cbind(sample(1:graph$nE, n.obs.1, replace=TRUE), runif(n.obs.1))

# Let us now add some locations on long edges:
long.edges <- graph$edge_lengths > 0.5

obs.loc <- rbind(obs.loc, cbind(sample(which(long.edges), n.obs.2, replace=TRUE), runif(n.obs.2)))

u <- sample_spde(kappa = kappa, sigma = sigma, alpha = 1,
                 graph = graph, PtE = obs.loc, method = "Q")

beta0 = -1
beta1 = 1
beta2 = 2
lonlat <- graph$coordinates(PtE = obs.loc)
scaled_lonlat <- scale(lonlat)
y <- beta0 + beta1 * scaled_lonlat[, 1] + beta2 * scaled_lonlat[, 2] + u + sigma_e*rnorm(n.obs)

data <- data.frame(edge_number = obs.loc[, 1],
                   distance_on_edge = obs.loc[, 2],
                   lon = scaled_lonlat[, 1],
                   lat = scaled_lonlat[, 2],
                   y = y)

graph$clear_observations()
graph$add_observations(data = data, normalized = TRUE)
```

```{r}
res <- graph_lme(y ~ lon + lat, graph = graph)
```

Let us get the summary of the result:
```{r}
summary(res)
```

We will now compare with the true values of the random effects:
```{r}
sigma_e_est <- res$coeff$measurement_error[[1]]
sigma_est <- res$coeff$random_effects[1]
range_est <- res$coeff$random_effects[2]
results <- data.frame(sigma_e = c(sigma_e, sigma_e_est),
                      sigma = c(sigma, sigma_est),
                      range = c(r, range_est),
                      row.names = c("Truth", "Estimate"))
print(results)
```
Given these estimated parameters, we can now do kriging to estimate the 
field at locations in the graph. As an example, we now obtain predictions 
on the regular mesh that we previously constructed. First, we obtain the covariates
on the mesh locations.




Now, we can compute the predictions for $y$. First, let us compute the posterior mean for the field at the observation locations and plot the residuals between the data and the field

```{r, fig.show='hold',fig.align = "center",fig.width=7,echo=TRUE}
pred_u <- predict(res, data = data, normalized = TRUE, only_latent = TRUE)
graph$plot(X = pred_u$mean - u, X_loc = data[,1:2], vertex_size = 0)
```

Let us now obtain predictions of the field on a mesh of equally spaced nodes on the graph. First, let us create the mesh and `data.frame`:
```{r}
graph$build_mesh(h = 50/1000)
lonlat_mesh <- graph$coordinates(PtE = graph$mesh$VtE)
scaled_lonlat_mesh <- scale(lonlat_mesh, 
                            center = attr(scaled_lonlat, "scaled:center"),
                            attr(scaled_lonlat, "scaled:scale"))
data_mesh_pred <- data.frame(lon = scaled_lonlat_mesh[,1],
                              lat = scaled_lonlat_mesh[,2],
                              edge_number = graph$mesh$VtE[,1],
                              distance_on_edge = graph$mesh$VtE[,2])
```

Now, let us obtain the predictions. We can obtain estimates for the latent field by setting `only_latent` to `TRUE`:

```{r, fig.show='hold',fig.align = "center",fig.width=7,echo=TRUE}
pred <- predict(res, data=data_mesh_pred, normalized=TRUE, only_latent=TRUE)
u_est <- pred$mean
graph$plot_function(u_est, vertex_size = 0, edge_width = 0.5)
```

Finally, let us obtain predictions for the observed values on the mesh. In this case we set
`only_latent` to `FALSE`:

```{r, fig.show='hold',fig.align = "center",fig.width=7,echo=TRUE}
pred_y <- predict(res, data=data_mesh_pred, normalized=TRUE, only_latent=FALSE)
y_est <- pred_y$mean
graph$plot_function(y_est, vertex_size = 0, edge_width = 0.5)
```

The same procedure can be done with $\alpha = 2$. One can also estimate $\alpha$
from data as described in the vignette 
[Whittle--Matérn fields with general smoothness](fem_models.html).


## Isotropic Gaussian fields

For metric graphs with Euclidean edges, [@Anderes2020](https://projecteuclid.org/journals/annals-of-statistics/volume-48/issue-4/Isotropic-covariance-functions-on-graphs-and-their-edges/10.1214/19-AOS1896.full)
showed that one can define  valid Gaussian processes through various isotropic
covariance functions if the distances between points are measured in the 
so-called resistance metric $d(\cdot,\cdot)$. One example of a valid covariance
function is the isotropic exponential covariance
function 
$$
r(d(s,t)) = \sigma^2\exp(-\kappa d(s,t)).
$$
This covariance is very similar to that of the 
Whittle--Mat\'ern fields with $\alpha = 1$. Between the two, we recommend
using the Whittle--Matérn model since it has Markov properties which makes inference much faster. Further, that covariance is well-defined for any
compact metric graph, whereas the isotropic exponential is only guaranteed to be
positive definite if the graph has Euclidean edges. See [@BSW2022b](??) for 
further comparisons. 

However, let us now illustrate how to use it for the data that we generated above. To work with the covariance function, the only cumbersome thing is to compute the metric. The `metric_graph` class has 
built in support for this, and we can obtain the distances between the observation locations as 
```{r}
graph$compute_resdist()
```

However, if the goal is to fit a model using this covariance function, there is no need for the user to compute it. It is done internally when one uses the `graph_lme()` function. We need to set the `model` argument in `graph_lme()` as a list with type `"isoCov"` (there is no need to add additional arguments, as the exponential covariance is the default). Let us fit a linear regression model with a random effect given by a Gaussian field with an isotropic exponential covariance function:

```{r}
res_exp <- graph_lme(y ~ lon + lat, graph = graph, model = list(type = "isoCov"))
summary(res_exp)
```

Let us now compute the posterior mean for the field at the observation 
locations and plot the residuals between the data and the field:

```{r, fig.show='hold',fig.align = "center",fig.width=7,echo=TRUE}
u_est_exp <- predict(res_exp, data = data, normalized = TRUE, only_latent = TRUE)
graph$plot(X = u_est_exp$mean - u, X_loc = data[,1:2], vertex_size = 0)
```

To perform kriging prediction to other locations, one can use the `predict()` method along with a `data.frame` containing the locations in which one wants to obtain predictions and the corresponding covariate values at these locations. In this example we will use the `data_mesh_pred` from the previous example.
We start by obtaining predictions for the latent model:

```{r, fig.align = "center",fig.width=7,echo=TRUE}
pred_exp <- predict(res_exp, data = data_mesh_pred, 
                    normalized=TRUE, only_latent = TRUE)
u_est_exp <- pred_exp$mean
graph$plot_function(u_est_exp, vertex_size = 0, edge_width = 0.5)
```

Finally, let us estimated the observed values at the mesh locations:

```{r, fig.align = "center",fig.width=7,echo=TRUE}
pred_exp_y <- predict(res_exp, data = data_mesh_pred, 
                    normalized=TRUE, only_latent = FALSE)
y_est_exp <- pred_exp_y$mean
graph$plot_function(y_est_exp, vertex_size = 0, edge_width = 0.5)
```

## Models based on the Graph Laplacian
A final set of Gaussian models that is supported by `MetricGraph` is the Matérn type
processes based on the graph Laplacian introduced
by [@Borovitskiy2021](http://proceedings.mlr.press/v130/borovitskiy21a/borovitskiy21a.pdf). These are multivariate Gaussian distributions, which are defined in the
vertices through the equation
$$
(\kappa^2\mathbf{I} - \mathbf{\Delta}_\Gamma)^{\alpha/2}\mathbf{u} = \mathbf{W}
$$
Here $\mathbf{W}\sim N(0,\sigma^2\mathbf{I})$ is a vector with independent Gaussian
variables and $\mathbf{\Delta}_\Gamma$ is the graph Laplacian. Further, $\mathbf{u}$
is a vector with the values of the process in the vertices of $\Gamma$, which
by definition has precision matrix
$$
\mathbf{Q} = \sigma^{-2}(\kappa^2\mathbf{I} - \mathbf{\Delta}_\Gamma)^{\alpha}
$$
Thus, to define these models, the only "difficult" thing is to compute the 
graph Laplacian. The (weighted) graph Laplacian, where the weights are specified
by the edge lengths can be computed by the function `compute_laplacian()` in the
`metric_graph` object. Suppose that we want to fit the data that we defined
above with this model. We can use the `graph_lme()` function. Also, observe that there is no need to 
use the `compute_laplacian()` function, as it is done internally. We now set the `model` argument as a list with the `type` being `"GraphLaplacian"`:

```{r}
res_gl <- graph_lme(y ~ lon + lat, graph = graph, model = list(type = "GraphLaplacian"))
summary(res_gl)
```

We can now obtain prediction at the observed locations by using the `predict()` method. Let us compute the posterior mean for the field at the observation locations and plot the residuals between the data and the field:

```{r, fig.show='hold',fig.align = "center",fig.width=7,echo=TRUE}
pred_u_GL <- predict(res_gl, data = data, normalized = TRUE, only_latent = TRUE)
graph$plot(X = pred_u_GL$mean - u, X_loc = data[,1:2], vertex_size = 0)
```

Now, if we do predictions outside of the observation locations on a graph Laplacian model, we need to modify the graph. This modifies the model in its entirety. Thus, we need to refit the model with all the observation locations we want to do predictions. However, if we use the `predict()` method with observations outside of the observation locations, the `predict()` will return predictions together with a `warning` that one should refit the model to obtain proper predictions:

```{r, fig.align = "center",fig.width=7,echo=TRUE}
pred_GL <- predict(res_gl, data = data_mesh_pred, 
                    normalized=TRUE, only_latent = TRUE)
u_est_GL <- pred_GL$mean
graph$plot_function(u_est_GL, vertex_size = 0, edge_width = 0.5)
```
and here the (incorrect way of obtaining) predictions of the observed data:
```{r, fig.align = "center",fig.width=7,echo=TRUE}
pred_GL_y <- predict(res_gl, data = data_mesh_pred, 
                    normalized=TRUE, only_latent = FALSE)
y_est_GL <- pred_GL_y$mean
graph$plot_function(y_est_GL, vertex_size = 0, edge_width = 0.5)
```




# The `inlabru` interface
In this vignette we will present our `inlabru`
interface to Whittle--Matérn fields. The `MetricGraph` package also has a similar interface to`R-INLA`, which is described in detail in
the [INLA interface of Whittle--Matérn fields](inla_interface.html) vignette.

## Basic setup and estimation
We will use the same graph and data as before. The `inlabru` implementation requires the observation locations to be added to the graph. However, note that for the Whittle--Matérn fields (contrary to the models based on the graph Laplacian) we are not changing the model by adding vertices at observation locations. We already created the extended graph above, so we can use that. Now, we load `INLA` and `inlabru` packages. We will also need to create the `inla` model object with the
`graph_spde` function. By default we have `alpha=1`. 

```{r, message=FALSE}
# library(INLA)
# library(inlabru)
# spde_model <- graph_spde(graph_extended)
```
The next step is to create the data list containing
the reponse variables and the `PtE` attribute:
```{r}
# data_list <- list(y = y, loc = graph_extended$PtE)
```

Now, we create `inlabru`'s component, which is a  formula-like object:
```{r}
# cmp <- y ~ -1 + field(loc, model = spde_model)
```

This formula is very simple since we are not assuming mean zero, so that we do not need an intercept, and we do not have any other covariates or model components. However, the setup is exactly  the same for more complicated models, with the only exception that we would have more terms in the formla. Now, we directly fit the model:

```{r}
# spde_bru_fit <- bru(cmp, data=data_list)
```

The advantage / difference between the estimates we obtain here and those above is that the `bru` function does full Bayesian inference (assuming priors for the model parameters). We used the default priors when creating the `graph_spde` model (see the help text for that function). The advantage now is that we do not only obtain point estimates but entire posterior distributions for the parameters. To view the estimates we can use the `spde_metric_graph_result()` function, then
taking a `summary()`:

```{r}
# spde_bru_result <- spde_metric_graph_result(spde_bru_fit, 
#                     "field", spde_model)

# summary(spde_bru_result)
```

Here we are showing the estimate of the practical correlation range ($2/\kappa$) instead of $\kappa$ since that is easier to interpret. We now compare the means of the estimated values with the true values:

```{r}
  # result_df_bru <- data.frame(
  #   parameter = c("std.dev", "range"),
  #   true = c(sigma, 2 / kappa),
  #   mean = c(
  #     spde_bru_result$summary.sigma$mean,
  #     spde_bru_result$summary.range$mean
  #   ),
  #   mode = c(
  #     spde_bru_result$summary.sigma$mode,
  #     spde_bru_result$summary.range$mode
  #   )
  # )
  # print(result_df_bru)
```

We can also plot the posterior marginal densities with the 
help of the `gg_df()` function:
```{r, fig.show='hold',fig.align = "center",echo=TRUE, fig.width=7, warning=FALSE}
  # posterior_df_bru_fit <- gg_df(spde_bru_result)

  # library(ggplot2)

  # ggplot(posterior_df_bru_fit) + geom_line(aes(x = x, y = y)) + 
  # facet_wrap(~parameter, scales = "free") + labs(y = "Density")
```


## Kriging with the `inlabru` implementation

There are two approaches to do kriging with our `inlabru` implementation of Whittle-Matérn fields on metric graphs. The simplest is to just add the locations where we want the predictions before fitting the model, and letting the data at those locations be `NA`. Again, since the Whittle--Matérn fields are invariant to the addition of vertices of degree two, this will not change the model. However, it will make the estimation slower since we need to work with a larger graph in `inlabru`. The second alternative, which is faster but also more complicated, is to do the fit without the prediction locations, and then add them at a second stage. We now illustrate how to do this. 


## An example with real data and replicates

# References {-}

