---
title: "Working with metric graphs"
author: "David Bolin, Alexandre B. Simas, and Jonas Wallin"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Working with metric graphs}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(GPGraph)
set.seed(1)
```

# Introduction

In this vignette we will introduce the `metric_graph` class. We will first show
how to construct metric graphs and then how to add data to them. For details 
about how to work with Gaussian processes on metric graphs, we refer to the
Vignettes 

* [Gaussian random fields on metric graphs](random_fields.html)
* [INLA interface of Whittle--Matérn fields](inla_interface.html)
* [inlabru interface of Whittle--Matérn fields](inlabru_interface.html)
* [Whittle--Matérn fields with general smoothness](fem_models.html)

# Constructing metric graphs

## Basic constructions
A metric graph can be constructed in two ways. The first is to specify all edges
in the graph as a `SpatialLines` object using the `sp` package. To illustrate
this, we first load the `sp` package
```{r}
library(sp)
```
and construct the following lines
```{r}
line1 <- Line(rbind(c(0,0),c(1,0)))
line2 <- Line(rbind(c(0,0),c(0,1)))
line3 <- Line(rbind(c(0,1),c(-1,1)))
theta <- seq(from=pi,to=3*pi/2,length.out = 20)
line4 <- Line(cbind(sin(theta),1+ cos(theta)))
Lines = sp::SpatialLines(list(Lines(list(line1),ID="1"),
                              Lines(list(line2),ID="2"),
                              Lines(list(line3),ID="3"),
                              Lines(list(line4),ID="4")))
```
We can now create the graph based on the `Lines` object as follows

```{r}
graph <- metric_graph$new(Lines = Lines)
graph$plot()
```

The `metric_graph` class then automatically constructs the list of edges, 
vertices, and edge lengths. These are available as public elements in the 
object: 
```{r}
V <- graph$V
E <- graph$E
edge_lengths <- graph$edge_lengths
```
An alternative method is to specify these in the construction:
```{r}
graph2 <- metric_graph$new(P = V, E = E, edge_lengths = edge_lengths)
graph2$plot()
```

This graph has the same properties as the first, but since we have not provided
the lines for each edge, it is visualized as having straight lines. If we know
that the graph only has straight lines, we may exclude the edge length
information in the construction: 
```{r}
graph3 <- metric_graph$new(P = V, E = E)
graph3$plot()
```

We now have three graphs with the same edges and vertices. The first two are 
identical from a modelling point of view, since they also have the same 
edge lengths. The third, is different since it assumes straight edges and 
therefore has a different edge length for one of the edges. 
```{r}
print(cbind(graph$edge_lengths, graph2$edge_lengths, graph3$edge_lengths))
```
Thus, from a modelling perspective, it is enough to specify the vertices,
edges, and edge lengths, but to obtain correct visualizations, the spatial
lines for the edges need to be specified. 

## An example with data from OpenStreetMap
This format of constructing graphs is compatible with data from OpenStreetMap,
and to illustrate this we use the `osmdata` to construct a metric graph 
consting of highways in Copenhagen:
```{r}
library(osmdata)
set_overpass_url("https://maps.mail.ru/osm/tools/overpass/api/interpreter")
call <- opq(bbox = c(12.4,55.5,12.65,55.9))
call <- add_osm_feature(call, key = "highway",value=c("motorway",
                                                        "primary","secondary"))
data <- osmdata_sp(call)

graph4 <- metric_graph$new(SpatialLines(data$osm_lines@lines))
graph4$plot(vertex_size = 0)
```
For complex networks like this, one might be interested in looking only at
removing lines that are not connected to the main network. This can be 
done by using the function `largest_component` on the lines object as follows:
```{r}
graph5 <- graph_components(SpatialLines(data$osm_lines@lines))
graph5$plot(vertex_size = 0)
```

We can also use the function to extract all connected components as follows:
```{r message=FALSE, warning=FALSE}
graphs <- graph_components(SpatialLines(data$osm_lines@lines),
                           only_largest =  FALSE)
p <- NULL
for(i in 1:(length(graphs)-1)){
p <- graphs[[i]]$plot(vertex_size = 0, 
                      edge_color = rgb(runif(1),runif(1),runif(1)),
                      p = p)  
}
graphs[[length(graphs)]]$plot(vertex_size = 0,
                              edge_color = rgb(runif(1),runif(1),runif(1)),
                              p = p)
```

# Adding data to the graph 

Given that we have constructed the metric graph, we can now add data to it. 
As an example, let us consider the first graph again and suppose that we have 
observations at a distance 0.5 from the start of each edge. One way of specifying
this is as follows 
```{r}
obs.loc <- cbind(1:4, rep(0.5, 4))
obs <- c(1,2,3,4)
graph$add_observations2(obs, PtE = obs.loc)
graph$plot(data = TRUE)
```

Thus, each row in `PtE` specifies the edge of the observation and the distance
along that edge where the observation location is. In certain situations, 
it might be easier to specify the relative distances on the edges, so that 
0 represents the start and 1 the end of the edge (instead of the edge length). 
To do so, we can simply specify `normalized = TRUE` when adding the 
observations. For example, let us add one more observation at the midpoint 
of the fourth edge:
```{r}
obs.loc <- matrix(c(4, 0.5),1,2)
obs <- c(5)
graph$add_observations2(obs, PtE = obs.loc, normalized = TRUE)
graph$plot(data = TRUE)
```

An alternative method is to specify the observations as spatial points objects,
where the locations are given in Euclidean coordinates: 
```{r}
obs.loc <- rbind(c(0.7, 0), c(0, 0.2))
obs <- c(6,7)
points <- SpatialPointsDataFrame(coords = obs.loc,
                                 data = data.frame(y = obs))
graph$add_observations(points)
graph$plot(data = TRUE)
```

If we want to replace the data in the object, we can use `clear_observations()`
to remove all current data, and then add the new data:
```{r}
graph$clear_observations()
graph$add_observations(points)
graph$plot(data = TRUE)
```

In some cases, we might want to add the observation locations as vertices 
in the graph. This can be done as follows: 
```{r}
graph$observation_to_vertex()
graph$plot(data = TRUE)
```
