---
title: "Working with metric graphs"
author: "David Bolin, Alexandre B. Simas, and Jonas Wallin"
date: "Created: 2022-11-23. Last modified: `r Sys.Date()`."
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Working with metric graphs}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(MetricGraph)
set.seed(1)
```

# Introduction

Networks such as street or river networks are examples of metric graphs. 
A compact metric graph $\Gamma$ consists of a set of finitely many vertices $\mathcal{V}=\{v_i\}$ and a finite set $\mathcal{E}=\{e_j\}$ of edges connecting the vertices. Each edge $e$ is a curve of finite length $l_e$ that connects two vertices. These curves are parameterized by arc length and a location $s\in \Gamma$ is a position on an edge, and can thus be represented as a touple $(e,t)$ where $t\in[0,l_e]$. 
Compared to regular graphs, where one typically defines functions on the vertices, we are for metric graphs interested in function that are defined on both the vertices and the edges. 

In this vignette we will introduce the `metric_graph` class of the `MetricGraph` package. This class provides a user friendly representation of metric graphs, and we will show how to use the class to construct and visualize metric graphs, add data to them, and work with functions defined on the graphs.

For details about Gaussian processes and inference on metric graphs, we refer to the Vignettes 

* [Gaussian random fields on metric graphs](random_fields.html)
* [INLA interface of Whittle--Matérn fields](inla_interface.html)
* [inlabru interface of Whittle--Matérn fields](inlabru_interface.html)
* [Whittle--Matérn fields with general smoothness](fem_models.html)

# Constructing metric graphs

## Basic constructions and properties
A metric graph can be constructed in two ways. The first is to specify all edges
in the graph as a `list` object, where each entry is a matrix. To illustrate
this, we first construct the following edges
```{r}
edge1 <- rbind(c(0,0),c(1,0))
edge2 <- rbind(c(0,0),c(0,1))
edge3 <- rbind(c(0,1),c(-1,1))
theta <- seq(from=pi,to=3*pi/2,length.out = 50)
edge4 <- cbind(sin(theta),1+ cos(theta))
edges = list(edge1, edge2, edge3, edge4)
```
We can now create the graph based on the `edges` object as follows

```{r, fig.show='hold',fig.align = "center",echo=TRUE}
graph <- metric_graph$new(edges = edges)
graph$plot()
```

The plot function that was used to create the plot above has various parameters
to set the sizes and colors of the vertices and edges, and it has a `plotly`
argument to visualize the graph in 3D. For this to work, the `plotly` library
must be installed.
```{r, message=FALSE, fig.show='hold',fig.align = "center",echo=TRUE}
graph$plot(plotly = TRUE, vertex_size = 5, vertex_color = "blue",
           edge_color = "red", edge_width = 2)
```

It is also important to know that the `2d` version of the
`plot()` method returns
a `ggplot2` object and can be modified as such. For instance:

```{r}
p <- graph$plot()
p + ggplot2::labs(title = "Metric graph",
          x = "long", y = "lat")
```

Similarly, the `3d` version of the `plot()` method returns a 
`plotly` object that can also be modified. For instance:
```{r}
p <- graph$plot(plotly = TRUE)
p <- plotly::layout(p, title = "Metric graph", 
              scene = list(xaxis=
              list(title = "Long"),yaxis=list(title = "Lat")))
p
```

We can now obtain various properties of the graph: The vertex matrix, which
specifies the Euclidian coordinates of the vertices is
```{r}
graph$V
```
and the edge matrix that specified the edges of the graph (i.e., which vertices
that are connected by edges) is
```{r}
graph$E
```
To obtain the geodesic (shortest path) distance between the vertices, we can use
the function `compute_geodist`:
```{r}
graph$compute_geodist()
graph$geo_dist
```
The second option it to construct the graph using two matrices `V` and `E` that
specify the locations (in Euclidean space) of the vertices and the edges. In this
case, it is assumed that the graph only has straight edges:
```{r, fig.show='hold',fig.align = "center",echo=TRUE}
V <- rbind(c(0, 0), c(1, 0), c(0, 1), c(-1, 1))
E <- rbind(c(1, 2), c(1, 3), c(3, 4), c(4, 1))
graph2 <- metric_graph$new(V = V, E = E)
graph2$plot()
```

A third option is to create a graph from a `SpatialLines` object:

```{r, message=FALSE, fig.show='hold',fig.align = "center",echo=TRUE}
library(sp)
line1 <- Line(rbind(c(0,0),c(1,0)))
line2 <- Line(rbind(c(0,0),c(0,1)))
line3 <- Line(rbind(c(0,1),c(-1,1)))
theta <- seq(from=pi,to=3*pi/2,length.out = 50)
line4 <- Line(cbind(sin(theta),1+ cos(theta)))
Lines = sp::SpatialLines(list(Lines(list(line1),ID="1"),
                              Lines(list(line2),ID="2"),
                              Lines(list(line3),ID="3"),
                              Lines(list(line4),ID="4")))

graph <- metric_graph$new(edges = Lines)                    
graph$plot()
```

A final option is to create from a `MULTILINESTRING` object:

```{r, message=FALSE, fig.show='hold',fig.align = "center",echo=TRUE}
library(sf)
line1 <- st_linestring(rbind(c(0,0),c(1,0)))
line2 <- st_linestring(rbind(c(0,0),c(0,1)))
line3 <- st_linestring(rbind(c(0,1),c(-1,1)))
theta <- seq(from=pi,to=3*pi/2,length.out = 50)
line4 <- st_linestring(cbind(sin(theta),1+ cos(theta)))
multilinestring = st_multilinestring(list(line1, line2, line3, line4))

graph <- metric_graph$new(edges = multilinestring)                    
graph$plot()
```

## Tolerances for merging vertices and edges
The constructor of the graph has one argument `tolerance` which is used for connecting edges that are close in Euclidean space. Specifically, the `tolerance` argument is given as a list with three elements:

* `vertex_vertex` vertices that are closer than this number are merged (the default value is `1e-7`)
* `vertex_edge` if a vertex at the end of one edge is closer than this number to another edge, this vertex is connected to that edge (the default value is `1e-7`)
* `edge_edge` if two edges at some point are closer than this number, a new
vertex is added at that point and the two edges are connected (the default value is `0` which means that the option is not used)

These options are often needed when constructing graphs based on real data, for example from OpenStreetMap as we will see later. 
To illustrate these options, suppose that we want to construct a graph from the 
following three edges:
```{r, fig.show='hold',fig.align = "center",echo=TRUE}
edge1 <- rbind(c(0,0),c(1,0))
edge2 <- rbind(c(0,0.03),c(0,0.75))
edge3 <- rbind(c(-1,1),c(0.5,0.03))
edge4 <- rbind(c(0,0.75), c(0,1))
edges =  list(edge1, edge2, edge3, edge4)
graph3 <- metric_graph$new(edges = edges)
graph3$plot(degree = TRUE)
print(graph3$nV)
```

We added the option `degree=TRUE` to the plot here to visualize the degrees of each vertex. As expected, one sees that all vertices have degree 1, and none of the three edges are connected. If these are streets in a street network, one might suspect that the two vertices at $(0,0)$ and $(0,0.03)$ really should be the same vertex so that the two edges are connected. This can be adjusted by increasing the `vertex_vertex` tolerance:
```{r, fig.show='hold',fig.align = "center",echo=TRUE}
graph3 <- metric_graph$new(edges = edges, tolerance = list(vertex_vertex = 0.05))
graph3$plot(degree = TRUE)
```

One might also want to add the vertex at $(0.5, 0.03)$ as a vertex on the first edge, so that the two edges there are connected. This can be done by adjusting the `vertex_edge` tolerance: 
```{r, fig.show='hold',fig.align = "center",echo=TRUE}
graph3 <- metric_graph$new(edges = edges, tolerance = list(vertex_vertex = 0.05,
                                                           vertex_edge = 0.1))
graph3$plot(degree = TRUE)
```

We can see that the vertex at $(0.5,0)$ was indeed connected with the edge from $(0,0)$ to $(1,0)$ and that vertex now has degree 3 since it is connected with three edges. One can also note that the `edges` object that was used to create the graph is modified internally in the `metric_graph` object so that the connections are visualized correctly in the sense that all edges are actually shown as connected edges. 

Finally, to add a vertex at the intersection between `edge2` and `edge3` we can adjust the `edge_edge` tolerance: 
```{r, fig.show='hold',fig.align = "center",echo=TRUE}
graph3 <- metric_graph$new(edges = edges, tolerance = list(vertex_vertex = 0.2,
                                                           vertex_edge = 0.1,
                                                           edge_edge = 0.001))
graph3$plot(degree = TRUE)
```

Now, the structure of the metric graph does not change if we add or remove vertices of degree 2. 
Because of this,one might want to remove vertices of degree 2 since this can reduce computational costs.
This can be done by setting the `remove_deg2` argument while creating the graph:
```{r, fig.show='hold',fig.align = "center",echo=TRUE}
graph3 <- metric_graph$new(edges = edges, tolerance = list(vertex_vertex = 0.2,
                                                           vertex_edge = 0.1,
                                                           edge_edge = 0.001),
                           remove_deg2 = TRUE)
graph3$plot(degree = TRUE)
```

Observe that one vertex of degree 2 was not removed. This is due to the fact that if we consider the direction, this vertex has directions incompatible for removal. We can see this by setting the argument `direction=TRUE` in the `plot()` method:

```{r, fig.show='hold',fig.align = "center",echo=TRUE}
graph3$plot(direction = TRUE)
```

## Characteristics of the graph

A brief summary of the main characteristics of the graph can be obtained by the `compute_characteristics()` method:

```{r}
graph3$compute_characteristics()
```

We can, then, view those characteristics by simply printing the `metric_graph` object:

```{r}
graph3
```

Observe that we do not know that the graph has Euclidean edges. We can check by using the `check_euclidean()` method:

```{r}
graph3$check_euclidean()
```

Let us view the characteristics again:

```{r}
graph3
```

It can happen that we have that the graph does not have Euclidean edges without the need to check the distance consistency. However, if one wants to check if the graph satisfies the distance consistency assumption, one can run the `check_distance_consistency()` function:

```{r}
graph3$check_distance_consistency()
```

Further, the individual characteristics can be accessed (after running the `compute_characteristics()` method) through the `characteristics` component of the graph object:

```{r}
graph3$characteristics
```

## Summaries of metric graphs

We can also obtain a summary of the informations contained on a metric graph object by using the `summary()` method. 
Let us obtain a summary of informations of `graph3`:

```{r}
summary(graph3)
```

Observe that there are some quantities that were not computed on the summary. We can see how to compute them by setting the argument `messages` to `TRUE`:

```{r}
summary(graph3, messages = TRUE)
```

Finally, the `summary()` can also be accessed from the metric graph object:

```{r}
graph3$summary()
```

## Vertices and edges

The metric graph object has the `edges` and `vertices` elements in its list. We can get a quick summary from them by calling these elements:

```{r}
graph3$vertices
```

and

```{r}
graph3$edges
```


## Understanding coordinates on graphs

The locations of the vertices are specified in Euclidean coordinates. However, 
when specifying a position on the graph, it is not practical to work with 
Euclidean coordinates since not all locations in Euclidean space are locations
on the graph. It is instead better to specify a location on the graph by the 
touple $(i, t)$, where $i$ denotes the number of the edge and $t$ is the 
location on the edge. The location $t$ can either be specified as the 
distance from the start of the edge (and then takes values between 0 and the 
length of the edge) or as the normalized distance from the start of the edge
(and then takes values between 0 and 1). The function `coordinates` can be used
to convert between coordinates in Euclidean space and locations on the graph. 
For example the location at distance 0.2 from the start of the second edge is:
```{r}
graph$coordinates(PtE = matrix(c(2, 0.2), 1,2), normalized = FALSE)
```
In this case, since the edge has length 1, the location of the point at 
normalized distance 0.2 from the start of the edge is the same:
```{r}
graph$coordinates(PtE = matrix(c(2, 0.2), 1,2), normalized = TRUE)
```

The function can also be used to find the closest location on the graph for 
a location in Euclidean space: 
```{r}
graph$coordinates(XY = matrix(c(0, 0.2), 1,2))
```
In this case, the `normalized` argument decides whether the returned value 
should be given in normalized distance or not. 

## Methods for working with real data

To illustrate the useage of `metric_graph` on some real data, we use the 
`osmdata` package to download data from OpenStreetMap. In the following code,
we extract highways in the city of Copenhagen:
```{r, fig.show='hold',fig.align = "center",fig.height = 7, echo=TRUE}
library(osmdata)
set_overpass_url("https://maps.mail.ru/osm/tools/overpass/api/interpreter")
call <- opq(bbox = c(12.4,55.5,12.65,55.9))
call <- add_osm_feature(call, key = "highway",value=c("motorway", "primary",
                                                      "secondary"))
data <- osmdata_sp(call)

graph5 <- metric_graph$new(SpatialLines(data$osm_lines@lines))
graph5$plot(vertex_size = 0)
```

There are a few things to note about data like this. The first is that the 
coordinates are given in Longitude and Latitude. Because of this, the edge 
lengths are by default given in degrees, which may result in very small 
numbers:
```{r}
range(graph5$get_edge_lengths())
```
This may cause numerical instabilities when dealing with random fields on the
graph, and it also makes it difficult to interpret results (unless one has a 
good intuition about distances in degrees). To avoid such problems, it is 
better to set the `longlat` argument when constructing the graph:
```{r}
graph5 <- metric_graph$new(SpatialLines(data$osm_lines@lines), longlat = TRUE)
```
This tells the constructor that the coordinates are given in Longitude and 
Latitude and that distances should be calculated in km. So if we now look at the
edge lengths, they are given in km:
```{r}
range(graph5$get_edge_lengths())
```
The second thing to note is that the constructor gave a warning that the graph
is not connected. This might not be ideal for modeling and we may want to study
the different connected components separately. If this is not a concern, one can
set the argument `check_connected = FALSE` while creating the graph. In this case
the check is not done and no warning message is printed. To construct all connected components, we can create a `graph_components` object:
```{r message=FALSE, warning=FALSE}
graphs <- graph_components$new(SpatialLines(data$osm_lines@lines), 
                                                    longlat = TRUE)
```
The `graph_components` class contains a list of `metric_graph` objects, one for
each connected component. In this case, we have 
```{r}
graphs$n
```
components in total, and their total edge lengths in km at 
```{r}
graphs$lengths
```
To plot all of them, we can use the `plot` command of the class:
```{r, message=FALSE, warning = FALSE, fig.show='hold',fig.align = "center",fig.height = 7,echo=TRUE}
graphs$plot(vertex_size = 0)
```

One reason for having multiple components here might be that we set the tolerance for merging nodes too low. In fact, 
by looking at the edge lengths we see that we have vertices that are as close 
as 
```{r}
min(graph5$get_edge_lengths(unit='m'))
```
meters. Let us increase the tolerances so that vertices that are at most
a distance of 100 meters (0.1km) apart should be merged:
```{r, message=FALSE, warning = FALSE, fig.show='hold',fig.align = "center",fig.height = 7,echo=TRUE}
graphs <- graph_components$new(SpatialLines(data$osm_lines@lines), longlat = TRUE,
                               tolerance = list(vertex_vertex = 0.1))
graphs$plot(vertex_size = 0)
```

With this choice, let us check the number of connected components:
```{r}
graphs$n
```

We can retrieve the graph as a standard `metric_graph` object to work 
with in further analysis via the `get_largest` command:
```{r}
graph5 <- graphs$get_largest()
```

Let us take a quick look at its vertices and edges:

```{r}
graph5$vertices
```

and

```{r}
graph5$edges
```

It is also relevant to obtain a summary of this graph:

```{r}
summary(graph5)
```

We have that this graph does not have Euclidean edges. Also observe that we have the units for the vertices and lengths, and also the coordinate reference system being used by this object.

# Adding data to the graph 

Given that we have constructed the metric graph, we can now add data to it. For further details on data manipulation on metric graphs, see 
[Data manipulation on metric graphs](metric_graph_data.html). 
As an example, let us consider the first graph again and suppose that we have 
observations at a distance 0.5 from the start of each edge. One way of specifying
this is as follows 
```{r, fig.show='hold',fig.align = "center",echo=TRUE}
obs.loc <- cbind(1:4, rep(0.5, 4))
obs <- c(1,2,3,4)
df_graph <- data.frame(y = obs, edge_number = obs.loc[,1], 
                        distance_on_edge = obs.loc[,2])
graph$add_observations(data = df_graph)
graph$plot(data = "y", data_size = 2)
```

In certain situations, it might be easier to specify the relative distances on 
the edges, so that 0 represents the start and 1 the end of the edge (instead of 
the edge length). To do so, we can simply specify `normalized = TRUE` when adding 
the observations. For example, let us add one more observation at the midpoint 
of the fourth edge:
```{r, fig.show='hold',fig.align = "center",echo=TRUE}
obs.loc <- matrix(c(4, 0.5),1,2)
obs <- c(5)
df_new <- data.frame(y=obs, edge_number = obs.loc[,1], 
                          distance_on_edge = obs.loc[,2])
graph$add_observations(data=df_new, normalized = TRUE)
graph$plot(data = "y")
```

An alternative method is to specify the observations as spatial points objects,
where the locations are given in Euclidean coordinates. In this case the
observations are added to the closes location on the graph:
```{r, fig.show='hold',fig.align = "center",echo=TRUE}
obs.loc <- rbind(c(0.7, 0), c(0, 0.2))
obs <- c(6,7)
points <- SpatialPointsDataFrame(coords = obs.loc,
                                 data = data.frame(y = obs))
graph$add_observations(points)
graph$plot(data = "y")
```

If we want to replace the data in the object, we can use `clear_observations()`
to remove all current data: 
```{r}
graph$clear_observations()
```

The `metric_graph` object can hold multiple variables of data, and one can also 
specify a `group` argument that it useful for specifying that data is grouped, 
which for example is the case with data observed at different time points.
As an example, let us add two variables observed at two separate time points: 
```{r}
obs.loc <- cbind(c(1:4, 1:4), c(rep(0.5, 4), rep(0.7, 4)))
df_rep <- data.frame(y = c(1, 2, NA, 3, 4, 6, 5, 7),
                     x = c(NA, 8, 9, 10, 11, 12, 13, NA),
                     edge_number = obs.loc[,1], 
                     distance_on_edge = obs.loc[,2],
                     time = c(rep(1, 4), rep(2, 4)))
graph$add_observations(data = df_rep, group = "time")
```
If `NA` is given in some variable, this indicates that this specific variable is
not measured at that location and group. We can now plot the individual
variables by specifying their names in the  plot function together with the
group number that we want to see. By default the first group is shown. 
```{r, fig.show='hold',fig.align = "center",echo=TRUE}
graph$plot(data = "y", group = 2)
```

In some cases, we might want to add the observation locations as vertices 
in the graph. This can be done as follows: 
```{r, fig.show='hold',fig.align = "center",echo=TRUE}
graph$observation_to_vertex()
graph$plot(data = "x", group = 1)
```

One can note that the command adds all observation locations, from all groups
as vertices. 

Finally, let us obtain a summary of the graph containing data:

```{r}
summary(graph)
```

# Working with functions on metric graphs

When working with data on metric graphs, one often wants to display functions
on the graph. The best way to visualize functions on the graph is to evaluate
them on a fine mesh over the graph and then use `plot_function`. To illustrate
this procedure, let us consider the following graph:
```{r, fig.show='hold',fig.align = "center",echo=TRUE}
V <- rbind(c(0, 0), c(1, 0), c(1, 1), c(0, 1), c(-1, 1), c(-1, 0), c(0, -1))
E <- rbind(c(1, 2), c(2, 3), c(3, 4), c(4, 5),
           c(5, 6), c(6, 1), c(4, 1),c(1, 7))
graph <- metric_graph$new(V = V, E = E)
graph$build_mesh(h = 0.5)
graph$plot(mesh=TRUE)
```

In the command `build_mesh`, the argument `h` decides the largest spacing between nodes in
the mesh. As can be seen in the plot, the mesh is very coarse, so let's reduce the value of `h` and rebuild the mesh:
```{r}
  graph$build_mesh(h = 0.01)
```

Suppose now that we want to display the function $f(x, y) = x^2 - y^2$ on this
graph. We then first evaluate it on the vertices of the mesh and then use the 
function `plot_function` to display it:
```{r, fig.show='hold',fig.align = "center",echo=TRUE}
x <- graph$mesh$V[, 1]
y <- graph$mesh$V[, 2]
f <- x^2 - y^2
graph$plot_function(X = f)
```

Alternatively, we can set `plotly = TRUE` in the plot command to get a 3D 
visualization of the function:
```{r, fig.show='hold',fig.align = "center",fig.height=7,echo=TRUE}
graph$plot_function(X = f, plotly = TRUE)
```

When the first argument of `plot_function` is a vector, the function assumes that 
the values in the vector are the values of the function evaluated at the vertices
of the mesh. As an alternative, one can also provide a new dataset.
To illustrate this, let us first construct a set of locations that are evenly 
spaced over each edge. Then, we will convert the data to Euclidean coordinates 
so that we can evaluate the function above. Then, process the data and finally plot the result:
```{r, fig.show='hold',fig.align = "center",fig.height=7,echo=TRUE}
n.e <- 30
PtE <- cbind(rep(1:graph$nE, each = n.e), 
             rep(seq(from = 0, to = 1, length.out = n.e), graph$nE))        
XY <- graph$coordinates(PtE)
f <- XY[, 1]^2 - XY[, 2]^2
data_f <- data.frame(f = f, "edge_number" = PtE[,1], 
            "distance_on_edge" = PtE[,2])
data_f <- graph$process_data(data = data_f, normalized=TRUE)            
graph$plot_function(data = "f", newdata = data_f, plotly = TRUE)
```

Alternatively, we can add the data to the graph, and directly plot by using the `data` argument.
To illustration purposes, we will build this next plot as a 2d plot:

```{r, fig.show='hold',fig.align = "center",fig.height=7,echo=TRUE}
graph$add_observations(data = data_f, clear_obs = TRUE, normalized = TRUE)
graph$plot_function(data = "f")
```
