---
title: "An example with multiple likelihoods in INLA and inlabru"
author: "David Bolin, Alexandre B. Simas, and Jonas Wallin"
date: "Created: 2024-07-06. Last modified: `r Sys.Date()`."
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{An example with multiple likelihoods in INLA and inlabru}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
references:
- id: BSW2022a
  title: "Gaussian Whittle--Matérn fields on metric graphs"
  author:
  - family: Bolin
    given: David
  - family: Simas
    given: Alexandre B.
  - family: Wallin
    given: Jonas
  container-title: Bernoulli
  type: article
  issue: 30
  pages: 1611-1639
  issued:
    year: 2024
- id: BSW2022b
  title: "Statistical properties of Gaussian Whittle--Matérn fields on metric graphs"
  author:
  - family: Bolin
    given: David
  - family: Simas
    given: Alexandre B.
  - family: Wallin
    given: Jonas
  container-title: arXiv:2304.10372
  type: preprint
  issued:
    year: 2023
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
set.seed(1)
library(MetricGraph)
```

# Introduction

In this vignette we will show how to fit a model with multiple likelihoods with our `INLA` and `inlabru` implementations. We will consider the following model:
$$Y_{1i} = \beta_1 + u(s_i) + \varepsilon_{1i},$$
$$Y_{2i} = \beta_2 + u(s_i) + \varepsilon_{2i},$$
where $s_1,\ldots,s_n$ are locations on a compact metric graph $\Gamma$, $u(\cdot)$ is a Whittle--Matérn field with `alpha=1`,  $i=1,\ldots,n$, $\epsilon_{11},\ldots, \epsilon_{1n}$ are i.i.d. random variables following $N(0, \sigma_1^2)$, and $\epsilon_{21}, \ldots, \epsilon_{2n}$ are i.i.d. random variables following $N(0,\sigma_2^2)$, finally, we will take $n=400$. 

# A toy dataset

We will start by generating the dataset. Let us load the `MetricGraph` package and create the metric graph:

```{r, message=FALSE}
library(MetricGraph)

edge1 <- rbind(c(0,0),c(1,0))
edge2 <- rbind(c(0,0),c(0,1))
edge3 <- rbind(c(0,1),c(-1,1))
theta <- seq(from=pi,to=3*pi/2,length.out = 20)
edge4 <- cbind(sin(theta),1+ cos(theta))
edges = list(edge1, edge2, edge3, edge4)
graph <- metric_graph$new(edges = edges)
```

Let us add 100 random locations in each edge where we will 
have observations:

```{r}
obs_per_edge <- 100
obs_loc <- NULL
for(i in 1:(graph$nE)) {
  obs_loc <- rbind(obs_loc,
                   cbind(rep(i,obs_per_edge), 
                   runif(obs_per_edge)))
}
```

We will now sample in these observation locations
and plot the latent field:

```{r, fig.show='hold',fig.align = "center",echo=TRUE}
sigma <- 2
alpha <- 1
nu <- alpha - 0.5
r <- 0.15 # r stands for range

u <- sample_spde(range = r, sigma = sigma, alpha = alpha,
                 graph = graph, PtE = obs_loc)
graph$plot(X = u, X_loc = obs_loc)
```

Let us now generate the observed responses for both likelihoods, which we will
call, respectively, `y1` and `y2`. We will also plot the observed responses on the
metric graph.

```{r}
beta1 = 2
beta2 = -2
n_obs <- length(u)
sigma1.e <- 0.2
sigma2.e <- 0.5

y1 <- beta1 + u + sigma1.e * rnorm(n_obs)
y2 <- beta2 + u + sigma2.e * rnorm(n_obs)
```

Let us plot the observations from `y1`:
```{r, fig.show='hold',fig.align = "center",echo=TRUE}
graph$plot(X = y1, X_loc = obs_loc)
```

and from `y2`:
```{r, fig.show='hold',fig.align = "center",echo=TRUE}
graph$plot(X = y2, X_loc = obs_loc)
```

# Fitting models with multiple likelihoods in `R-INLA`

We are now in a position to fit the model with our
`R-INLA` implementation. To this end,
we need to add the observations to the graph, which we 
will do with the `add_observations()` method. We will create a column on the `data.frame` to indicate which likelihood the observed variable belongs to. We will also the intercepts as columns.

```{r, fig.show='hold',fig.align = "center",echo=TRUE}
df_graph1 <- data.frame(y = y1, intercept_1 = 1, intercept_2 = NA,
                        edge_number = obs_loc[,1],
                        distance_on_edge = obs_loc[,2],
                        likelihood = 1)
df_graph2 <- data.frame(y = y2, intercept_1 = NA,
                        intercept_2 = 1,
                        edge_number = obs_loc[,1],
                        distance_on_edge = obs_loc[,2],
                        likelihood = 2)      
df_graph <- rbind(df_graph1, df_graph2)               
```

Let us now add the observations and set the `likelihood` column as `group`:

```{r, message=FALSE, fig.show='hold',fig.align = "center",echo=TRUE}
graph$add_observations(data=df_graph, normalized=TRUE, group = "likelihood")
graph$plot(data="y")
```

Now, we load the `R-INLA` package and
create the `inla` model object with the
`graph_spde` function. By default we have
`alpha=1`. 

```{r, message=FALSE}
library(INLA)
spde_model <- graph_spde(graph)
```

Now, we need to create the data object with the `graph_data_spde()` function, in which we need to provide a name for the random effect, which we will call `field`, and we need to provide the covariates. We also need to pass the column that contains the number of the likelihood for the data 

```{r}
data_spde <- graph_data_spde(graph_spde = spde_model, 
                name = "field", likelihood_col = "likelihood",
                resp_col = "y",
                covariates = c("intercept_1", "intercept_2"))
```

The remaining is standard in `R-INLA`. We create the
formula object and the
`inla.stack` objects with the `inla.stack()` function.

Let us start by creating the formula:

```{r}
f.s <- y ~ -1 + intercept_1 + intercept_2 + f(field, model = spde_model)
```

Let us now create the `inla.stack` objects, one for each likelihood. To such an end, we simply supply the data in `data_spde` obtained from using `graph_data_spde`:

```{r}
stk_dat1 <- inla.stack(data = data_spde[[1]][["data"]], 
                        A = data_spde[[1]][["basis"]], 
                        effects = data_spde[[1]][["index"]]
    )
stk_dat2 <- inla.stack(data = data_spde[[2]][["data"]], 
                        A = data_spde[[2]][["basis"]], 
                        effects = data_spde[[2]][["index"]]
    )
stk_dat <- inla.stack(stk_dat1, stk_dat2)    
```

Now, we use the `inla.stack.data()`:
```{r}
data_stk <- inla.stack.data(stk_dat)
```

Finally, we fit the model:

```{r, message=FALSE, warning=FALSE}
spde_fit <- inla(f.s, family = c("gaussian", "gaussian"), 
    data = data_stk, control.predictor=list(A=inla.stack.A(stk_dat)))
```

Let us now obtain the estimates in the original scale
by using the `spde_metric_graph_result()` function, then
taking a `summary()`:
      
```{r}
spde_result <- spde_metric_graph_result(spde_fit, "field", spde_model)

summary(spde_result)
```


We will now compare the means of the estimated values with
the true values:

```{r}
  result_df <- data.frame(
    parameter = c("std.dev", "range"),
    true = c(sigma, r),
    mean = c(
      spde_result$summary.sigma$mean,
      spde_result$summary.range$mean
    ),
    mode = c(
      spde_result$summary.sigma$mode,
      spde_result$summary.range$mode
    )
  )
  print(result_df)
```

Let us now look at the estimates of the measurement errors and compare with the true ones:

```{r}
meas_err_df <- data.frame(
    parameter = c("sigma1.e", "sigma2.e"),
    true = c(sigma1.e, sigma2.e),
    mean = sqrt(1/spde_fit$summary.hyperpar$mean[1:2]),
    mode = sqrt(1/spde_fit$summary.hyperpar$mode[1:2])
  )
print(meas_err_df)
```

Finally, let us look at the estimates of the intercepts:

```{r}
intercept_df <- data.frame(
    parameter = c("beta1", "beta2"),
    true = c(beta1, beta2),
    mean = spde_fit$summary.fixed$mean,
    mode = spde_fit$summary.fixed$mode
  )
print(intercept_df)
```


# Fitting models with multiple likelihoods in `inlabru`

Under development.

# A toy dataset with multiple likelihoods and replicates

Let us now proceed similarly, but now we will consider a case in which we have multiple likelihoods and replicates.

To simplify exposition, we will use the same base graph.
So, we begin by clearing the observations.
```{r}
graph$clear_observations()
```

We will use the same observation locations as for the
previous cases. Let us sample 10 replicates:
```{r}
sigma_rep <- 1.5
alpha_rep <- 1
nu_rep <- alpha_rep - 0.5
r_rep <- 0.2 # r stands for range
kappa_rep <- sqrt(8 * nu_rep) / r_rep

n_repl <- 10

u_rep <- sample_spde(range = r_rep, sigma = sigma_rep,
                 alpha = alpha_rep,
                 graph = graph, PtE = obs_loc,
                 nsim = n_repl)
```

Let us now generate the observed responses, which we will
call `y_rep`.

```{r, fig.show='hold',fig.align = "center",echo=TRUE}
beta1 = 2
beta2 = -2

sigma1.e <- 0.2
sigma2.e <- 0.5

n_obs_rep <- nrow(u_rep)

y1_rep <- beta1 + u_rep + sigma1.e * matrix(rnorm(n_obs_rep * n_repl),
                                    ncol=n_repl)     
y2_rep <- beta2 + u_rep + sigma2.e * matrix(rnorm(n_obs_rep * n_repl),
                                    ncol=n_repl)            
```

# Fitting the model with multiple likelihoods and replicates in `R-INLA`

The `sample_spde()` function returns a matrix in which each replicate
is a column. We need to stack the columns together and a column to
indicate the replicat. Further, we need to do it for each likelihood:
```{r, fig.show='hold',fig.align = "center",echo=TRUE}
dl1_graph <- lapply(1:ncol(y1_rep), function(i){data.frame(y = y1_rep[,i],
                                          edge_number = obs_loc[,1],
                                          distance_on_edge = obs_loc[,2],
                                          likelihood = 1,
                                          intercept_1 = 1,
                                          intercept_2 = NA,
                                          repl = i)})
dl1_graph <- do.call(rbind, dl1_graph)
```
and
```{r, fig.show='hold',fig.align = "center",echo=TRUE}
dl2_graph <- lapply(1:ncol(y2_rep), function(i){data.frame(y = y2_rep[,i],
                                          edge_number = obs_loc[,1],
                                          distance_on_edge = obs_loc[,2],
                                          likelihood = 2,
                                          intercept_1 = NA,
                                          intercept_2 = 1,
                                          repl = i)})
dl2_graph <- do.call(rbind, dl2_graph)
```

We now join them:
```{r}
dl_graph <- rbind(dl1_graph, dl2_graph)
```

We can now add the the observations by setting the `group`
argument to `c("repl", "likelihood")`:
```{r, message=FALSE}
graph$add_observations(data = dl_graph, normalized=TRUE, 
                            group = c("repl", "likelihood"),
                            edge_number = "edge_number",
                            distance_on_edge = "distance_on_edge")
```                            

Let us now create the model object:

```{r}
spde_model_rep <- graph_spde(graph)
```

Let us first consider a case in which we do not use all
replicates. Then, we consider the case in which we use all
replicates. 

Thus, let us assume we want only to consider replicates
1, 3, 5, 7 and 9. To this end,
we the index object by using the
`graph_data_spde()` function with the
argument `repl` set to the replicates we want,
in this case `c(1,3,5,7,9)`. Observe that here we need to pass `repl_col`, as the internal grouping variable is not the replicate variable. 
```{r}
data_spde_repl <- graph_data_spde(graph_spde=spde_model_rep,
                      name="field", repl = c(1,3,5,7,9), repl_col = "repl", 
                      likelihood_col = "likelihood", resp_col = "y",
                      covariates = c("intercept_1", "intercept_2"))
```

Next, we create the stack objects, remembering that
we need to input the components from `data_spde` for each likelihood:
```{r}
stk_dat_rep1 <- inla.stack(data = data_spde_repl[[1]][["data"]], 
                        A = data_spde_repl[[1]][["basis"]], 
                        effects = data_spde_repl[[1]][["index"]]
    )
stk_dat_rep2 <- inla.stack(data = data_spde_repl[[2]][["data"]], 
                        A = data_spde_repl[[2]][["basis"]], 
                        effects = data_spde_repl[[2]][["index"]]
    )

stk_dat_rep <- inla.stack(stk_dat_rep1, stk_dat_rep2)
```

We now create the formula object, adding the
name of the field (in our case `field`)
attached with `.repl` a the `replicate` argument
inside the `f()` function.
```{r}
f_s_rep <- y ~ -1 + intercept_1 + intercept_2 + 
    f(field, model = spde_model_rep, 
        replicate = field.repl)
```
Then, we create the stack object with The
`inla.stack.data()` function:
```{r}
data_stk_rep <- inla.stack.data(stk_dat_rep)
```
Now, we fit the model:
```{r}
spde_fit_rep <- inla(f_s_rep, family = c("gaussian", "gaussian"), 
                data = data_stk_rep, 
                control.predictor=list(A=inla.stack.A(stk_dat_rep)))
```

Let us see the estimated values in the original scale:
```{r}
spde_result_rep <- spde_metric_graph_result(spde_fit_rep, 
                        "field", spde_model_rep)

summary(spde_result_rep)
```

Let us compare with the true values:

```{r}
  result_df_rep <- data.frame(
    parameter = c("std.dev", "range"),
    true = c(sigma_rep, r_rep),
    mean = c(
      spde_result_rep$summary.sigma$mean,
      spde_result_rep$summary.range$mean
    ),
    mode = c(
      spde_result_rep$summary.sigma$mode,
      spde_result_rep$summary.range$mode
    )
  )
  print(result_df_rep)
```



Let us now look at the estimates of the measurement errors and compare with the true ones:

```{r}
meas_err_df <- data.frame(
    parameter = c("sigma1.e", "sigma2.e"),
    true = c(sigma1.e, sigma2.e),
    mean = sqrt(1/spde_fit_rep$summary.hyperpar$mean[1:2]),
    mode = sqrt(1/spde_fit_rep$summary.hyperpar$mode[1:2])
  )
print(meas_err_df)
```

Finally, let us look at the estimates of the intercepts:

```{r}
intercept_df <- data.frame(
    parameter = c("beta1", "beta2"),
    true = c(beta1, beta2),
    mean = spde_fit_rep$summary.fixed$mean,
    mode = spde_fit_rep$summary.fixed$mode
  )
print(intercept_df)
```

# Fitting models with multiple likelihoods and replicates in `inlabru`

Under development.

