---
title: "Log-Gaussian Cox processes on metric graphs"
author: "David Bolin, Alexandre B. Simas"
date: "Created: 2023-01-30. Last modified: `r Sys.Date()`."
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Log-Gaussian Cox processes on metric graphs}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
bibliography: references.bib
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
set.seed(1)
```

## Introduction
        
In this vignette we will introduce how to work with log-Gaussian Cox processes
based on Whittle--Matérn fields on metric graphs. To simplify the integration
with `R-INLA` and `inlabru` hese models are constructed using finite element 
approximations as implemented in the `rSPDE` package.  The theoretical details will be given in the forthcoming article [@BSW2023PPST].
   
## Constructing the graph and the mesh
We begin by loading the `rSPDE`, `MetricGraph` and `INLA` packages:

```{r, message=FALSE, warning=FALSE}
library(rSPDE)
library(MetricGraph)
library(INLA)
```

As an example, we consider the default graph in the package:

```{r, fig.show='hold',fig.align = "center",echo=TRUE, warning=FALSE, message=FALSE}
graph <- metric_graph$new(tolerance = list(vertex_vertex = 1e-1, vertex_edge = 1e-3, edge_edge = 1e-3),
                          remove_deg2 = TRUE)
graph$plot()
```

To construct a FEM approximation of a Whittle--Matérn field,
we must first construct a mesh on the graph.
```{r, fig.show='hold',fig.align = "center",echo=TRUE,  warning=FALSE}
graph$build_mesh(h = 0.1)
graph$plot(mesh=TRUE)
```

The next step is to build the mass and stiffness matrices for the FEM basis.
```{r}
  graph$compute_fem()
```

We are now ready to specify the and sample from a log-Gaussian Cox process model
with intensity $\lambda = \exp(\beta + u)$ where $\beta$ is an intercept and 
$u$ is a Gaussian Whittle--Matérn field specified by
$$
(\kappa^2 - \Delta)^{\alpha/2} \tau u = \mathcal{W}.
$$
For this we can use the function `graph_lgcp` as follows:
```{r}
  sigma <- 0.5
  range <- 2
  alpha <- 2
  cov_lgcp <- graph$mesh$VtE[,1]/max(graph$mesh$VtE[,1])
  lgcp_sample <- graph_lgcp_sim(intercept = -1 + 2*cov_lgcp, sigma = sigma,
                            range = range, alpha = alpha,
                            graph = graph)
```
The object returned by the function is a list with the simulated Gaussian process and the points on the graph. We can plot the simulated intensity function as
```{r, fig.show='hold',fig.align = "center",echo=TRUE,  warning=FALSE}
graph$plot_function(X = exp(lgcp_sample$u), vertex_size = 0)
```

To plot the simulated points, we can add them to the graph and then plot:
```{r, fig.show='hold',fig.align = "center",echo=TRUE,  warning=FALSE}
graph$add_observations(data = data.frame(y=rep(1,length(lgcp_sample$edge_loc)),
                                         edge_number = lgcp_sample$edge_number,
                                         distance_on_edge = lgcp_sample$edge_loc,
                                         cov_lgcp = lgcp_sample$edge_number),
                       normalized = TRUE)
graph$plot(vertex_size = 0, data = "y")
```

In order to fit a log-Gaussian Cox process model, we need to specify integration points on the graph, be able to evaluate the covariates at such integration points. This process is a bit involved, so we have created an interface to simplify the process. In this interface, by default, the covariates are interpolated from the data provided in the `graph` object to obtain their values at the integration points. 

The integration points are defined, by default, as the mesh locations if the metric graph object has a mesh. If the mesh is not provided, one must either provide the integration points manually or build a mesh.

At the end of the vignette, we will also show how to fit the model in `INLA` without using our `INLA` interface for LGCP models.


# Fitting LGCP models with our `INLA` interface

We will now fit the model using our `INLA` interface. To such an end, we will clear the observations from the graph and add the data to the graph.

```{r}
graph$clear_observations()

#Add the data together with the exposure terms
graph$add_observations(data = data.frame(y = rep(1,length(lgcp_sample$edge_loc)),
                                         edge_number = lgcp_sample$edge_number,
                                         distance_on_edge = lgcp_sample$edge_loc,
                                         Intercept = 1,
                                         cov_lgcp = lgcp_sample$edge_number/max(lgcp_sample$edge_number)),
                       normalized = TRUE)
```

We have added the response variable $y$, however, this is not strictly necessary. If such a response variable is not provided, it will be assumed that all locations correspond to observed points.

Let us now create the `rSPDE` model object:

```{r}
rspde_model <- rspde.metric_graph(graph, nu = alpha - 1/2)
```

We can now fit the model using the `lgcp_graph()` function:

```{r}
inla_fit <- lgcp_graph(y ~ -1 + Intercept + cov_lgcp + 
                      f(field, model = rspde_model), graph=graph)
```

Let us observe the `inla_fit` object:

```{r}
summary(inla_fit)
```

Let us extract the estimates in the original scale
by using the `spde_metric_graph_result()` function, then
taking a `summary()`:
      
```{r}
spde_result <- spde_metric_graph_result(inla_fit, "field", rspde_model)

summary(spde_result)
```

We will now compare the means of the estimated values with
the true values:

```{r}
  result_df <- data.frame(
    parameter = c("std.dev", "range"),
    true = c(sigma, range),
    mean = c(
      spde_result$summary.std.dev$mean,
      spde_result$summary.range$mean
    ),
    mode = c(
      spde_result$summary.std.dev$mode,
      spde_result$summary.range$mode
    )
  )
  print(result_df)
```

If we have the actual values of the covariates at the integration points, we can pass them to the `lgcp_graph()` function via the `manual_covariates` argument.

```{r}
manual_covariates <- data.frame(Intercept = 1,
                      cov_lgcp = graph$mesh$VtE[,1]/max(lgcp_sample$edge_number),
                      .group = 1)
inla_fit <- lgcp_graph(y ~ -1 + Intercept + cov_lgcp + f(field, model = rspde_model), 
             graph=graph, manual_covariates = manual_covariates, interpolate = FALSE)
```

Let us observe the new `inla_fit` object:

```{r}
summary(inla_fit)
```

We can now extract the estimates in the original scale

```{r}
spde_result <- spde_metric_graph_result(inla_fit, "field", rspde_model)

summary(spde_result)
```


We can also plot the posterior marginal densities with the 
help of the `gg_df()` function:
```{r, fig.show='hold',fig.align = "center",echo=TRUE,  warning=FALSE}
  posterior_df_fit <- gg_df(spde_result)

  library(ggplot2)

  ggplot(posterior_df_fit) + geom_line(aes(x = x, y = y)) + 
  facet_wrap(~parameter, scales = "free") + labs(y = "Density")
```

Finally, we can plot the estimated field $u$:
```{r, fig.show='hold',fig.align = "center",echo=TRUE,  warning=FALSE}
n.obs <- length(graph$get_data()$y)
n.field <- dim(graph$mesh$VtE)[1]
u_posterior <- inla_fit$summary.linear.predictor$mean[(n.obs+1):(n.obs+n.field)]
graph$plot_function(X = u_posterior, vertex_size = 0)
```

This can be compared with the field that was used to generate the data:
```{r, fig.show='hold',fig.align = "center",echo=TRUE,  warning=FALSE}
graph$plot_function(X = lgcp_sample$u, vertex_size = 0)
```


We can also fit the model using the exact model by using the `graph_spde()` function as the SPDE model:

```{r}
spde_model <- graph_spde(graph, alpha = 1)
```

Let us now fit the model:

```{r}
inla_fit_spde <- lgcp_graph(y ~ -1 + Intercept + cov_lgcp + 
                          f(field, model = spde_model), graph=graph)
```

Let us observe the new `inla_fit_spde` object:

```{r}
summary(inla_fit_spde)
```

We can now extract the estimates in the original scale

```{r}
spde_result <- spde_metric_graph_result(inla_fit_spde, "field", spde_model)

summary(spde_result)
```

# An example with replicates in our `INLA` interface

We start by simulating the data.

```{r}
  n.rep <- 30
  sigma <- 0.5
  range <- 2
  alpha <- 2
  cov_lgcp <- graph$mesh$VtE[,1]/max(graph$mesh$VtE[,1])  
  max_edge_num <- max(graph$mesh$VtE[,1])
  lgcp_sample_rep <- graph_lgcp_sim(n = n.rep, intercept = -1 + 2*cov_lgcp, sigma = sigma,
                            range = range, alpha = alpha,
                            graph = graph)
```

Let us clear the observations from the graph and add the simulated data.

```{r}  
graph$clear_observations()

df_rep <- data.frame(y=rep(1,length(lgcp_sample_rep[[1]]$edge_loc)),
                     edge_number = lgcp_sample_rep[[1]]$edge_number,
                     distance_on_edge = lgcp_sample_rep[[1]]$edge_loc,
                     Intercept = 1,
                     cov_lgcp = lgcp_sample_rep[[1]]$edge_number/max_edge_num,
                     rep = rep(1,length(lgcp_sample_rep[[1]]$edge_loc)))

for(i in 2:n.rep){
  df_rep <- rbind(df_rep, data.frame(y=rep(1,length(lgcp_sample_rep[[i]]$edge_loc)),
                                       edge_number = lgcp_sample_rep[[i]]$edge_number,
                                       distance_on_edge = lgcp_sample_rep[[i]]$edge_loc,
                                       Intercept = 1,
                                       cov_lgcp = lgcp_sample_rep[[i]]$edge_number/max_edge_num,
                                       rep = rep(i,length(lgcp_sample_rep[[i]]$edge_loc))))
}

graph$add_observations(data = df_rep,
                       normalized = TRUE,
                        group = "rep")

```

Let us now fit the model. In this case, the default column for the replicates is `.group`, and also, by default, all replicates will be used.

```{r}
inla_fit <- lgcp_graph(y ~ -1 + Intercept + cov_lgcp + f(field, model = rspde_model, replicate = field.repl), graph=graph)
```

Let us observe the `inla_fit` object:

```{r}
summary(inla_fit)
```

Let us now extract the estimates in the original scale

```{r}
spde_result <- spde_metric_graph_result(inla_fit, "field", rspde_model)

summary(spde_result)
```

As in the previous case, we can also supply the covariates manually:

```{r}
manual_covariates <- data.frame(
      Intercept = 1,
      cov_lgcp = rep(graph$mesh$VtE[,1]/max(graph$mesh$VtE[,1]), 30),
      .group = rep(1:30, each = nrow(graph$mesh$VtE))
    )
inla_fit <- lgcp_graph(y ~ -1 + Intercept + cov_lgcp + 
            f(field, model = rspde_model, replicate = field.repl), 
              graph=graph, manual_covariates = manual_covariates)
```

Let us observe the `inla_fit` object:

```{r}
summary(inla_fit)
```

Let us now extract the estimates in the original scale

```{r}
spde_result <- spde_metric_graph_result(inla_fit, "field", rspde_model)

summary(spde_result)
```

We can also fit the model with replicates using the exact model:

```{r}
inla_fit_spde_rep <- lgcp_graph(y ~ -1 + Intercept + cov_lgcp + 
                               f(field, model = spde_model, replicate = field.repl), 
                               graph=graph)
```

Let us observe the `inla_fit_spde_rep` object:

```{r}
summary(inla_fit_spde_rep)
```

We can now extract the estimates in the original scale

```{r}
spde_result_rep <- spde_metric_graph_result(inla_fit_spde_rep, "field", spde_model)

summary(spde_result_rep)
```

# Fitting LGCP models without our `INLA` interface

We are now in a position to fit the model with our `R-INLA` implementation, without using our `INLA` interface for LGCP models. 
When working with log-Gaussian Cox processes, the likelihood has a term 
$\int_\Gamma \exp(u(s)) ds$ that needs to be handled separately. 
This is done by using the mid-point rule as suggested for SPDE models by [@lgcp](https://academic.oup.com/biomet/article/103/1/49/2389990)
where we approximate 
$$
\int_\Gamma \exp(u(s)) ds \approx \sum_{i=1}^p \widetilde{a}_i \exp\left(u(\widetilde{s}_i)\right).
$$
Using the fact that $u(s) = \sum_{j=1}^n \varphi(s) u_i$ from the FEM approximation, 
we can write the integral as $\widetilde{\alpha}^T\exp(\widetilde{A}u)$ where
$\widetilde{A}_{ij} = \varphi_j(\widetilde{s}_i)$ and $\widetilde{a}$ is a vector
with integration weights. These quantities can be obtained as
```{r}
Atilde <- graph$fem_basis(graph$mesh$VtE)
atilde <- graph$mesh$weights
```
The weights are used as exposure terms in the Poisson likelihood in R-INLA. Because 
of this, the easiest way to construct the model is to add the integration points 
as zero observations in the graph, with corresponding exposure weights. We also 
need to add the exposure terms (which are zero) for the actual observation locations:
```{r}
#clear the previous data in the graph
graph$clear_observations()

#Add the data together with the exposure terms
graph$add_observations(data = data.frame(y = rep(1,length(lgcp_sample$edge_loc)),
                                         e = rep(0,length(lgcp_sample$edge_loc)),
                                         edge_number = lgcp_sample$edge_number,
                                         distance_on_edge = lgcp_sample$edge_loc,
                                         Intercept = 1,
                                         cov_lgcp = lgcp_sample$edge_number/max(lgcp_sample$edge_number)),
                       normalized = TRUE)

#Add integration points
graph$add_observations(data = data.frame(y = rep(0,length(atilde)),
                                         e = atilde,
                                         edge_number = graph$mesh$VtE[,1],
                                         distance_on_edge = graph$mesh$VtE[,2],
                                         Intercept = 1,
                                         cov_lgcp = graph$mesh$VtE[,1]/max(lgcp_sample$edge_number)),
                       normalized = TRUE)

```
We now create the `inla` model object with the
`graph_spde` function. For simplicity, we assume that $\alpha$ is known and fixed
to the true value in the model.  

```{r}
rspde_model <- rspde.metric_graph(graph, nu = alpha - 1/2)
```

Next, we compute the auxiliary data:

```{r}
data_rspde <- graph_data_spde(rspde_model, name="field", covariates = c("Intercept","cov_lgcp"))
```

We now create the `inla.stack` object with the `inla.stack()` function. 
At this stage, it is important that the data has been added to the `graph` since
it is supplied to the stack by using the `graph_spde_data()` function. 

```{r}
stk <- inla.stack(data = data_rspde[["data"]], 
                  A = data_rspde[["basis"]],
                  effects = data_rspde[["index"]])
```

We can now fit the model using `R-INLA`:

```{r, message=FALSE, warning=FALSE}
spde_fit <- inla(y ~ -1 + Intercept + cov_lgcp + f(field, model = rspde_model), 
                 family = "poisson", data = inla.stack.data(stk),
                 control.predictor = list(A = inla.stack.A(stk), compute = TRUE),
                 E = inla.stack.data(stk)$e)
```

Let us extract the estimates in the original scale
by using the `spde_metric_graph_result()` function, then
taking a `summary()`:
      
```{r}
spde_result <- rspde.result(spde_fit, "field", rspde_model)

summary(spde_result)
```

We will now compare the means of the estimated values with
the true values:

```{r}
  result_df <- data.frame(
    parameter = c("std.dev", "range"),
    true = c(sigma, range),
    mean = c(
      spde_result$summary.std.dev$mean,
      spde_result$summary.range$mean
    ),
    mode = c(
      spde_result$summary.std.dev$mode,
      spde_result$summary.range$mode
    )
  )
  print(result_df)
```


# An example with replicates

We now clear the previous data and add the new data together with the exposure terms
```{r}
  graph$clear_observations()
  df_rep <- data.frame(y=rep(1,length(lgcp_sample_rep[[1]]$edge_loc)),
                                             e = rep(0,length(lgcp_sample_rep[[1]]$edge_loc)),
                                         edge_number = lgcp_sample_rep[[1]]$edge_number,
                                         distance_on_edge = lgcp_sample_rep[[1]]$edge_loc,
                                         Intercept = 1,
                                         cov_lgcp = lgcp_sample_rep[[1]]$edge_number/max_edge_num,
                                         rep = rep(1,length(lgcp_sample_rep[[1]]$edge_loc)))

  df_rep <- rbind(df_rep, data.frame(y = rep(0,length(atilde)),
                                         e = atilde,
                                         edge_number = graph$mesh$VtE[,1],
                                         distance_on_edge = graph$mesh$VtE[,2],
                                         Intercept = 1,
                                         cov_lgcp = graph$mesh$VtE[,1]/max_edge_num,
                                         rep = rep(1,length(atilde))))
  for(i in 2:n.rep){
    df_rep <- rbind(df_rep, data.frame(y=rep(1,length(lgcp_sample_rep[[i]]$edge_loc)),
                                             e = rep(0,length(lgcp_sample_rep[[i]]$edge_loc)),
                                         edge_number = lgcp_sample_rep[[i]]$edge_number,
                                         distance_on_edge = lgcp_sample_rep[[i]]$edge_loc,
                                         Intercept = 1,
                                         cov_lgcp = lgcp_sample_rep[[i]]$edge_number/max_edge_num,
                                         rep = rep(i,length(lgcp_sample_rep[[i]]$edge_loc))))
    df_rep <- rbind(df_rep, data.frame(y = rep(0,length(atilde)),
                                         e = atilde,
                                         edge_number = graph$mesh$VtE[,1],
                                         distance_on_edge = graph$mesh$VtE[,2],
                                         Intercept = 1,
                                         cov_lgcp = graph$mesh$VtE[,1]/max_edge_num,
                                         rep = rep(i,length(atilde))))                                        

  }

      graph$add_observations(data = df_rep,
                       normalized = TRUE,
                        group = "rep")

```
We can now define and fit the model as previously
```{r}
rspde_model <- rspde.metric_graph(graph, nu = alpha - 1/2)

data_rspde <- graph_data_spde(rspde_model, name = "field", 
                                  repl = ".all", repl_col = "rep", 
                                  covariates = c("Intercept","cov_lgcp"))

stk <- inla.stack(data = data_rspde[["data"]], 
                  A = data_rspde[["basis"]], 
                  effects = data_rspde[["index"]])

spde_fit <- inla(y ~ -1 + Intercept + cov_lgcp + 
                 f(field, model = rspde_model, replicate = field.repl), 
                 family = "poisson", data = inla.stack.data(stk),
                 control.predictor = list(A = inla.stack.A(stk), compute = TRUE),
                 E = inla.stack.data(stk)$e)
```
Let's look at the summaries
```{r}
spde_result <- rspde.result(spde_fit, "field", rspde_model)
summary(spde_result)
result_df <- data.frame(
    parameter = c("std.dev", "range"),
    true = c(sigma, range),
    mean = c(
      spde_result$summary.std.dev$mean,
      spde_result$summary.range$mean
    ),
    mode = c(
      spde_result$summary.std.dev$mode,
      spde_result$summary.range$mode
    )
  )
  print(result_df)
```


# Using precomputed data for efficient model fitting

When fitting multiple LGCP models with different formulas but the same spatial structure and covariates, it can be very efficient to precompute the expensive quantities once and reuse them. This is particularly useful for model selection, cross-validation, or exploring different covariate combinations.

The `precompute_lgcp_graph()` function allows us to precompute integration points, mesh setup, and SPDE model structures. Then `lgcp_graph()` can use this precomputed data to fit models much faster. This approach is especially beneficial when using exact SPDE models created with the `graph_spde()` function, as these models require more computationally expensive setup operations compared to the rational SPDE models from the rSPDE package.

## Example without replicates

Let's start with a timing comparison for the single replicate case. First, we'll set up the data and create a precomputed object:

```{r}
# Clear and add the single replicate data
graph$clear_observations()
graph$add_observations(data = data.frame(y = rep(1,length(lgcp_sample$edge_loc)),
                                         edge_number = lgcp_sample$edge_number,
                                         distance_on_edge = lgcp_sample$edge_loc,
                                         Intercept = 1,
                                         cov_lgcp = lgcp_sample$edge_number/max(lgcp_sample$edge_number)),
                       normalized = TRUE)

# Create precomputed object with all available covariates
precomputed_data <- precompute_lgcp_graph(
  resp_variable_name = "y",
  spde_model = spde_model,
  model_name = "field",
  graph = graph,
  covariates = c("Intercept", "cov_lgcp"),
  use_current_mesh = TRUE
)
```

Now let's compare timings between the original approach and using precomputed data:

```{r}
# Time the original approach (multiple fits)
time_original <- system.time({
  fit1_orig <- lgcp_graph(y ~ -1 + Intercept + f(field, model = spde_model), 
                          graph = graph)
  fit2_orig <- lgcp_graph(y ~ -1 + Intercept + cov_lgcp + f(field, model = spde_model), 
                          graph = graph)
  fit3_orig <- lgcp_graph(y ~ -1 + cov_lgcp + f(field, model = spde_model), 
                          graph = graph)
})

# Time the precomputed approach (multiple fits)
time_precomputed <- system.time({
  fit1_precomp <- lgcp_graph(y ~ -1 + Intercept + f(field, model = spde_model), 
                             graph = graph, precomputed_data = precomputed_data)
  fit2_precomp <- lgcp_graph(y ~ -1 + Intercept + cov_lgcp + f(field, model = spde_model), 
                             graph = graph, precomputed_data = precomputed_data)
  fit3_precomp <- lgcp_graph(y ~ -1 + cov_lgcp + f(field, model = spde_model), 
                             graph = graph, precomputed_data = precomputed_data)
})

# Time for creating precomputed object
time_precompute <- system.time({
  precomputed_temp <- precompute_lgcp_graph(
  resp_variable_name = "y",
  spde_model = rspde_model,
  model_name = "field",
  graph = graph,
  covariates = c("Intercept", "cov_lgcp"),
  use_current_mesh = TRUE
)
})

# Create timing comparison table
timing_single <- data.frame(
  Method = c("Original (3 fits)", "Precomputation", "Precomputed (3 fits)", "Total precomputed"),
  Time_seconds = c(
    time_original[["elapsed"]], 
    time_precompute[["elapsed"]], 
    time_precomputed[["elapsed"]],
    time_precompute[["elapsed"]] + time_precomputed[["elapsed"]]
  )
)

print("Timing comparison for single replicate:")
print(timing_single)
```

Let's verify that the results are equivalent:

```{r}
# Compare log marginal likelihoods to verify equivalence
cat("Log marginal likelihood comparison (fit2):\n")
cat("Original:", fit2_orig$mlik[1], "\n")
cat("Precomputed:", fit2_precomp$mlik[1], "\n")
cat("Difference:", abs(fit2_orig$mlik[1] - fit2_precomp$mlik[1]), "\n")
```

### Using manual covariates with precomputation

We can also use manual covariates with precomputation. This is useful when you have the exact values of covariates at the integration points (mesh nodes in `graph$mesh$VtE`):

```{r}
# Create manual covariates based on mesh nodes
manual_covariates <- data.frame(
  Intercept = 1,
  cov_lgcp = graph$mesh$VtE[,1]/max(graph$mesh$VtE[,1]),
  .group = 1
)

# Precompute with manual covariates (interpolate = FALSE)
precomputed_manual <- precompute_lgcp_graph(
  graph = graph,
  resp_variable_name = "y",
  model_name = "field",
  covariates = c("Intercept", "cov_lgcp"),
  spde_model = spde_model,
  manual_covariates = manual_covariates,
  interpolate = FALSE,
  use_current_mesh = TRUE
)

# Fit model using manual covariates precomputed data
time_manual <- system.time({
  fit_manual <- lgcp_graph(y ~ -1 + Intercept + cov_lgcp + f(field, model = spde_model), 
                          graph = graph, precomputed_data = precomputed_manual)
})

cat("Manual covariates fit time:", time_manual[["elapsed"]], "seconds\n")
cat("Manual covariates log marginal likelihood:", fit_manual$mlik[1], "\n")
```

### Performance optimization: avoiding graph cloning

For maximum performance, you can avoid cloning the graph by setting `clone_graph = FALSE`. This works directly on the original graph, which is faster but modifies the input:

```{r}
# Performance comparison: with and without cloning
time_with_clone <- system.time({
  fit_clone <- lgcp_graph(y ~ -1 + Intercept + cov_lgcp + f(field, model = spde_model), 
                         graph = graph, clone_graph = TRUE)
})

time_without_clone <- system.time({
  fit_no_clone <- lgcp_graph(y ~ -1 + Intercept + cov_lgcp + f(field, model = spde_model), 
                            graph = graph, clone_graph = FALSE)
})

cat("Time with cloning:", time_with_clone[["elapsed"]], "seconds\n")
cat("Time without cloning:", time_without_clone[["elapsed"]], "seconds\n")
cat("Speedup factor:", round(time_with_clone[["elapsed"]] / time_without_clone[["elapsed"]], 2), "x\n")

# The results should be identical
cat("Log marginal likelihood difference:", 
    abs(fit_clone$mlik[1] - fit_no_clone$mlik[1]), "\n")
```

You can also use `clone_graph = FALSE` with precomputation for even better performance:

```{r}
# Precompute without cloning (faster but modifies the graph)
time_precomp_no_clone <- system.time({
  precomputed_no_clone <- precompute_lgcp_graph(
    resp_variable_name = "y",
    model_name = "field",
    graph = graph,
    covariates = c("Intercept", "cov_lgcp"),
    spde_model = spde_model,    
    clone_graph = FALSE
  )
})

# Fit using precomputed data without cloning
time_fit_no_clone <- system.time({
  fit_precomp_no_clone <- lgcp_graph(y ~ -1 + Intercept + cov_lgcp + f(field, model = rspde_model), 
                                    graph = graph, 
                                    precomputed_data = precomputed_no_clone)
})

cat("Precomputation time (no clone):", time_precomp_no_clone[["elapsed"]], "seconds\n")
cat("Fit time with precomputed data (no clone):", time_fit_no_clone[["elapsed"]], "seconds\n")
```



## Example with replicates

Now let's do the same comparison for the replicated data:

```{r}
# Clear and add the replicated data
graph$clear_observations()
graph$add_observations(data = df_rep, normalized = TRUE, group = "rep")

# Create precomputed object for replicated data
precomputed_data_rep <- precompute_lgcp_graph(
  graph = graph,
  resp_variable_name = "y",
  model_name = "field",
  spde_model = spde_model,
  covariates = c("Intercept", "cov_lgcp"),
  repl = ".all",
  repl_col = "rep",
  use_current_mesh = TRUE
)
```

Time comparisons for replicated data:

```{r}
# Time the original approach (multiple fits with replicates)
time_original_rep <- system.time({
  fit1_orig_rep <- lgcp_graph(y ~ -1 + Intercept + f(field, model = spde_model, replicate = field.repl), 
                              graph = graph)
  fit2_orig_rep <- lgcp_graph(y ~ -1 + Intercept + cov_lgcp + f(field, model = spde_model, replicate = field.repl), 
                              graph = graph)
  fit3_orig_rep <- lgcp_graph(y ~ -1 + cov_lgcp + f(field, model = spde_model, replicate = field.repl), 
                              graph = graph)
})

# Time the precomputed approach (multiple fits with replicates)
time_precomputed_rep <- system.time({
  fit1_precomp_rep <- lgcp_graph(y ~ -1 + Intercept + f(field, model = spde_model, replicate = field.repl), 
                                 graph = graph, precomputed_data = precomputed_data_rep)
  fit2_precomp_rep <- lgcp_graph(y ~ -1 + Intercept + cov_lgcp + f(field, model = spde_model, replicate = field.repl), 
                                 graph = graph, precomputed_data = precomputed_data_rep)
  fit3_precomp_rep <- lgcp_graph(y ~ -1 + cov_lgcp + f(field, model = spde_model, replicate = field.repl), 
                                 graph = graph, precomputed_data = precomputed_data_rep)
})

# Time for creating precomputed object with replicates
time_precompute_rep <- system.time({
  precomputed_temp_rep <- precompute_lgcp_graph(
    graph = graph,
    resp_variable_name = "y",
    model_name = "field",
    spde_model = spde_model,
    covariates = c("Intercept", "cov_lgcp"),
    repl = ".all",
    repl_col = "rep",
    use_current_mesh = TRUE
  )
})

# Create timing comparison table for replicates
timing_replicates <- data.frame(
  Method = c("Original (3 fits)", "Precomputation", "Precomputed (3 fits)", "Total precomputed"),
  Time_seconds = c(
    time_original_rep[["elapsed"]], 
    time_precompute_rep[["elapsed"]], 
    time_precomputed_rep[["elapsed"]],
    time_precompute_rep[["elapsed"]] + time_precomputed_rep[["elapsed"]]
  )
)

print("Timing comparison for replicated data:")
print(timing_replicates)
```

Let's verify that the results are equivalent:

```{r}
# Compare log marginal likelihoods to verify equivalence
cat("Log marginal likelihood comparison (fit2):\n")
cat("Original:", fit2_orig_rep$mlik[1], "\n")
cat("Precomputed:", fit2_precomp_rep$mlik[1], "\n")
cat("Difference:", abs(fit2_orig_rep$mlik[1] - fit2_precomp_rep$mlik[1]), "\n")
```

### Using manual covariates with replicates

For replicated data, manual covariates must include the replicate structure:

```{r}
# Create manual covariates for replicated data based on mesh nodes
manual_covariates_rep <- data.frame(
  Intercept = 1,
  cov_lgcp = rep(graph$mesh$VtE[,1]/max(graph$mesh$VtE[,1]), 30),
  rep = rep(1:30, each = nrow(graph$mesh$VtE))
)

# Precompute with manual covariates for replicates
precomputed_manual_rep <- precompute_lgcp_graph(
  graph = graph,
  resp_variable_name = "y",
  model_name = "field",
  spde_model = rspde_model,
  covariates = c("Intercept", "cov_lgcp"),
  manual_covariates = manual_covariates_rep,
  interpolate = FALSE,
  repl = ".all",
  repl_col = "rep",
  use_current_mesh = TRUE
)

# Fit model using manual covariates for replicates
time_manual_rep <- system.time({
  fit_manual_rep <- lgcp_graph(y ~ -1 + Intercept + cov_lgcp + f(field, model = rspde_model, replicate = field.repl), 
                              graph = graph, precomputed_data = precomputed_manual_rep)
})

cat("Manual covariates (replicates) fit time:", time_manual_rep[["elapsed"]], "seconds\n")
cat("Manual covariates (replicates) log marginal likelihood:", fit_manual_rep$mlik[1], "\n")
```

## Summary and speedup analysis

Let's create a comprehensive comparison table and calculate speedups. Observe that this is a small graph and simple model, so the speedup is not as pronounced as it would be for larger graphs and more complex models. Also, the speedup is not as pronounced for replicated data as it is for single replicate data.

```{r}
# Calculate speedups
speedup_single <- time_original[["elapsed"]] / (time_precompute[["elapsed"]] + time_precomputed[["elapsed"]])
speedup_replicates <- time_original_rep[["elapsed"]] / (time_precompute_rep[["elapsed"]] + time_precomputed_rep[["elapsed"]])

# If we do more fits, the speedup becomes more pronounced
# Calculate break-even point (when precomputation becomes beneficial)
breakeven_single <- time_precompute[["elapsed"]] / (time_original[["elapsed"]]/3 - time_precomputed[["elapsed"]]/3)
breakeven_replicates <- time_precompute_rep[["elapsed"]] / (time_original_rep[["elapsed"]]/3 - time_precomputed_rep[["elapsed"]]/3)

summary_table <- data.frame(
  Scenario = c("Single replicate", "Multiple replicates"),
  Original_3fits = c(time_original[["elapsed"]], time_original_rep[["elapsed"]]),
  Precomputed_total = c(
    time_precompute[["elapsed"]] + time_precomputed[["elapsed"]],
    time_precompute_rep[["elapsed"]] + time_precomputed_rep[["elapsed"]]
  ),
  Speedup = c(speedup_single, speedup_replicates),
  Breakeven_fits = pmax(1, ceiling(c(breakeven_single, breakeven_replicates)))
)

print("Summary of timing results:")
print(summary_table)

cat("\nKey observations:\n")
cat("- Precomputation becomes beneficial when fitting", min(summary_table$Breakeven_fits), "or more models\n")
cat("- For 3 fits, speedup ranges from", sprintf("%.2f", min(summary_table$Speedup)), "x to", sprintf("%.2f", max(summary_table$Speedup)), "x\n")
cat("- The more models you fit, the greater the relative benefit of precomputation\n")
```

## When to use precomputation

The `precompute_lgcp_graph()` function is most beneficial when:

1. **Multiple model fits**: You plan to fit several models with different formulas but the same spatial structure
2. **Model selection**: Comparing models with different covariate combinations
3. **Sensitivity analysis**: Testing different covariate subsets while keeping the spatial component fixed

**Important considerations**:
- You can only use covariates that were included in the original precomputation
- The spatial structure (mesh, integration points) is fixed during precomputation
- The benefit increases with the number of model fits you perform
- Manual covariates can be provided based on mesh nodes (`graph$mesh$VtE`) for exact covariate values at integration points
- Use `clone_graph = FALSE` for maximum performance when you don't need to preserve the original graph state

**Example use case**: If you have 10 potential covariates and want to compare all possible models with 1-3 covariates, precomputation would allow you to fit dozens of models very efficiently after the initial setup cost.

## References
