---
title: "Log-Gaussian Cox processes on metric graphs"
author: "David Bolin, Alexandre B. Simas"
date: "Created: 2023-01-30. Last modified: `r Sys.Date()`."
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Log-Gaussian Cox processes on metric graphs}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
bibliography: references.bib
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
set.seed(1)
```

## Introduction
        
In this vignette we will introduce how to work with log-Gaussian Cox processes
based on Whittle--Matérn fields on metric graphs. To simplify the integration
with `R-INLA` and `inlabru` hese models are constructed using finite element 
approximations as implemented in the `rSPDE` package.  The theoretical details will be given in the forthcoming article [@BSW2023PPST].
   
## Constructing the graph and the mesh
We begin by loading the `rSPDE`, `MetricGraph` and `INLA` packages:

```{r, message=FALSE, warning=FALSE}
library(rSPDE)
library(MetricGraph)
library(INLA)
```

As an example, we consider the default graph in the package:

```{r, fig.show='hold',fig.align = "center",echo=TRUE, warning=FALSE, message=FALSE}
graph <- metric_graph$new(tolerance = list(vertex_vertex = 1e-1, vertex_edge = 1e-3, edge_edge = 1e-3),
                          remove_deg2 = TRUE)
graph$plot()
```

To construct a FEM approximation of a Whittle--Matérn field,
we must first construct a mesh on the graph.
```{r, fig.show='hold',fig.align = "center",echo=TRUE,  warning=FALSE}
graph$build_mesh(h = 0.1)
graph$plot(mesh=TRUE)
```

The next step is to build the mass and stiffness matrices for the FEM basis.
```{r}
  graph$compute_fem()
```

We are now ready to specify the and sample from a log-Gaussian Cox process model
with intensity $\lambda = \exp(\beta + u)$ where $\beta$ is an intercept and 
$u$ is a Gaussian Whittle--Matérn field specified by
$$
(\kappa^2 - \Delta)^{\alpha/2} \tau u = \mathcal{W}.
$$
For this we can use the function `graph_lgcp` as follows:
```{r}
  sigma <- 0.5
  range <- 2
  alpha <- 2
  cov_lgcp <- graph$mesh$VtE[,1]/max(graph$mesh$VtE[,1])
  lgcp_sample <- graph_lgcp_sim(intercept = -1 + 2*cov_lgcp, sigma = sigma,
                            range = range, alpha = alpha,
                            graph = graph)
```
The object returned by the function is a list with the simulated Gaussian process and the points on the graph. We can plot the simulated intensity function as
```{r, fig.show='hold',fig.align = "center",echo=TRUE,  warning=FALSE}
graph$plot_function(X = exp(lgcp_sample$u), vertex_size = 0)
```

To plot the simulated points, we can add them to the graph and then plot:
```{r, fig.show='hold',fig.align = "center",echo=TRUE,  warning=FALSE}
graph$add_observations(data = data.frame(y=rep(1,length(lgcp_sample$edge_loc)),
                                         edge_number = lgcp_sample$edge_number,
                                         distance_on_edge = lgcp_sample$edge_loc,
                                         cov_lgcp = lgcp_sample$edge_number),
                       normalized = TRUE)
graph$plot(vertex_size = 0, data = "y")
```

In order to fit a log-Gaussian Cox process model, we need to specify integration points on the graph, be able to evaluate the covariates at such integration points. This process is a bit involved, so we have created an interface to simplify the process. In this interface, by default, the covariates are interpolated from the data provided in the `graph` object to obtain their values at the integration points. 

The integration points are defined, by default, as the mesh locations if the metric graph object has a mesh. If the mesh is not provided, one must either provide the integration points manually or build a mesh.

At the end of the vignette, we will also show how to fit the model in `INLA` without using our `INLA` interface for LGCP models.


# Fitting LGCP models with our `INLA` interface

We will now fit the model using our `INLA` interface. To such an end, we will clear the observations from the graph and add the data to the graph.

```{r}
graph$clear_observations()

#Add the data together with the exposure terms
graph$add_observations(data = data.frame(y = rep(1,length(lgcp_sample$edge_loc)),
                                         edge_number = lgcp_sample$edge_number,
                                         distance_on_edge = lgcp_sample$edge_loc,
                                         Intercept = 1,
                                         cov_lgcp = lgcp_sample$edge_number/max(lgcp_sample$edge_number)),
                       normalized = TRUE)
```

We have added the response variable $y$, however, this is not strictly necessary. If such a response variable is not provided, it will be assumed that all locations correspond to observed points.

Let us now create the `rSPDE` model object:

```{r}
rspde_model <- rspde.metric_graph(graph, nu = alpha - 1/2)
```

We can now fit the model using the `lgcp_graph()` function:

```{r}
inla_fit <- lgcp_graph(y ~ -1 + Intercept + cov_lgcp + 
                      f(field, model = rspde_model), graph=graph)
```

Let us observe the `inla_fit` object:

```{r}
summary(inla_fit)
```

Let us extract the estimates in the original scale
by using the `spde_metric_graph_result()` function, then
taking a `summary()`:
      
```{r}
spde_result <- spde_metric_graph_result(inla_fit, "field", rspde_model)

summary(spde_result)
```

We will now compare the means of the estimated values with
the true values:

```{r}
  result_df <- data.frame(
    parameter = c("std.dev", "range"),
    true = c(sigma, range),
    mean = c(
      spde_result$summary.std.dev$mean,
      spde_result$summary.range$mean
    ),
    mode = c(
      spde_result$summary.std.dev$mode,
      spde_result$summary.range$mode
    )
  )
  print(result_df)
```

If we have the actual values of the covariates at the integration points, we can pass them to the `lgcp_graph()` function via the `manual_covariates` argument.

```{r}
manual_covariates <- data.frame(Intercept = 1,
                      cov_lgcp = graph$mesh$VtE[,1]/max(lgcp_sample$edge_number),
                      .group = 1)
inla_fit <- lgcp_graph(y ~ -1 + Intercept + cov_lgcp + f(field, model = rspde_model), 
             graph=graph, manual_covariates = manual_covariates, interpolate = FALSE)
```

Let us observe the new `inla_fit` object:

```{r}
summary(inla_fit)
```

We can now extract the estimates in the original scale

```{r}
spde_result <- spde_metric_graph_result(inla_fit, "field", rspde_model)

summary(spde_result)
```


We can also plot the posterior marginal densities with the 
help of the `gg_df()` function:
```{r, fig.show='hold',fig.align = "center",echo=TRUE,  warning=FALSE}
  posterior_df_fit <- gg_df(spde_result)

  library(ggplot2)

  ggplot(posterior_df_fit) + geom_line(aes(x = x, y = y)) + 
  facet_wrap(~parameter, scales = "free") + labs(y = "Density")
```

Finally, we can plot the estimated field $u$:
```{r, fig.show='hold',fig.align = "center",echo=TRUE,  warning=FALSE}
n.obs <- length(graph$get_data()$y)
n.field <- dim(graph$mesh$VtE)[1]
u_posterior <- inla_fit$summary.linear.predictor$mean[(n.obs+1):(n.obs+n.field)]
graph$plot_function(X = u_posterior, vertex_size = 0)
```

This can be compared with the field that was used to generate the data:
```{r, fig.show='hold',fig.align = "center",echo=TRUE,  warning=FALSE}
graph$plot_function(X = lgcp_sample$u, vertex_size = 0)
```


We can also fit the model using the exact model by using the `graph_spde()` function as the SPDE model:

```{r}
spde_model <- graph_spde(graph, alpha = 1)
```

Let us now fit the model:

```{r}
inla_fit_spde <- lgcp_graph(y ~ -1 + Intercept + cov_lgcp + 
                          f(field, model = spde_model), graph=graph)
```

Let us observe the new `inla_fit_spde` object:

```{r}
summary(inla_fit_spde)
```

We can now extract the estimates in the original scale

```{r}
spde_result <- spde_metric_graph_result(inla_fit_spde, "field", spde_model)

summary(spde_result)
```

# An example with replicates in our `INLA` interface

We start by simulating the data.

```{r}
  n.rep <- 30
  sigma <- 0.5
  range <- 2
  alpha <- 2
  cov_lgcp <- graph$mesh$VtE[,1]/max(graph$mesh$VtE[,1])  
  max_edge_num <- max(graph$mesh$VtE[,1])
  lgcp_sample_rep <- graph_lgcp_sim(n = n.rep, intercept = -1 + 2*cov_lgcp, sigma = sigma,
                            range = range, alpha = alpha,
                            graph = graph)
```

Let us clear the observations from the graph and add the simulated data.

```{r}  
graph$clear_observations()

df_rep <- data.frame(y=rep(1,length(lgcp_sample_rep[[1]]$edge_loc)),
                     edge_number = lgcp_sample_rep[[1]]$edge_number,
                     distance_on_edge = lgcp_sample_rep[[1]]$edge_loc,
                     Intercept = 1,
                     cov_lgcp = lgcp_sample_rep[[1]]$edge_number/max_edge_num,
                     rep = rep(1,length(lgcp_sample_rep[[1]]$edge_loc)))

for(i in 2:n.rep){
  df_rep <- rbind(df_rep, data.frame(y=rep(1,length(lgcp_sample_rep[[i]]$edge_loc)),
                                       edge_number = lgcp_sample_rep[[i]]$edge_number,
                                       distance_on_edge = lgcp_sample_rep[[i]]$edge_loc,
                                       Intercept = 1,
                                       cov_lgcp = lgcp_sample_rep[[i]]$edge_number/max_edge_num,
                                       rep = rep(i,length(lgcp_sample_rep[[i]]$edge_loc))))
}

graph$add_observations(data = df_rep,
                       normalized = TRUE,
                        group = "rep")

```

Let us now fit the model. In this case, the default column for the replicates is `.group`, and also, by default, all replicates will be used.

```{r}
inla_fit <- lgcp_graph(y ~ -1 + Intercept + cov_lgcp + f(field, model = rspde_model, replicate = field.repl), graph=graph)
```

Let us observe the `inla_fit` object:

```{r}
summary(inla_fit)
```

Let us now extract the estimates in the original scale

```{r}
spde_result <- spde_metric_graph_result(inla_fit, "field", rspde_model)

summary(spde_result)
```

As in the previous case, we can also supply the covariates manually:

```{r}
manual_covariates <- data.frame(
      Intercept = 1,
      cov_lgcp = rep(graph$mesh$VtE[,1]/max(graph$mesh$VtE[,1]), 30),
      .group = rep(1:30, each = nrow(graph$mesh$VtE))
    )
inla_fit <- lgcp_graph(y ~ -1 + Intercept + cov_lgcp + 
            f(field, model = rspde_model, replicate = field.repl), 
              graph=graph, manual_covariates = manual_covariates)
```

Let us observe the `inla_fit` object:

```{r}
summary(inla_fit)
```

Let us now extract the estimates in the original scale

```{r}
spde_result <- spde_metric_graph_result(inla_fit, "field", rspde_model)

summary(spde_result)
```


# Fitting LGCP models without our `INLA` interface

We are now in a position to fit the model with our `R-INLA` implementation, without using our `INLA` interface for LGCP models. 
When working with log-Gaussian Cox processes, the likelihood has a term 
$\int_\Gamma \exp(u(s)) ds$ that needs to be handled separately. 
This is done by using the mid-point rule as suggested for SPDE models by [@lgcp](https://academic.oup.com/biomet/article/103/1/49/2389990)
where we approximate 
$$
\int_\Gamma \exp(u(s)) ds \approx \sum_{i=1}^p \widetilde{a}_i \exp\left(u(\widetilde{s}_i)\right).
$$
Using the fact that $u(s) = \sum_{j=1}^n \varphi(s) u_i$ from the FEM approximation, 
we can write the integral as $\widetilde{\alpha}^T\exp(\widetilde{A}u)$ where
$\widetilde{A}_{ij} = \varphi_j(\widetilde{s}_i)$ and $\widetilde{a}$ is a vector
with integration weights. These quantities can be obtained as
```{r}
Atilde <- graph$fem_basis(graph$mesh$VtE)
atilde <- graph$mesh$weights
```
The weights are used as exposure terms in the Poisson likelihiood in R-INLA. Because 
of this, the easiest way to construct the model is to add the integration points 
as zero observations in the graph, with corresponding exposure weights. We also 
need to add the exposure terms (which are zero) for the actual observation locations:
```{r}
#clear the previous data in the graph
graph$clear_observations()

#Add the data together with the exposure terms
graph$add_observations(data = data.frame(y = rep(1,length(lgcp_sample$edge_loc)),
                                         e = rep(0,length(lgcp_sample$edge_loc)),
                                         edge_number = lgcp_sample$edge_number,
                                         distance_on_edge = lgcp_sample$edge_loc,
                                         Intercept = 1,
                                         cov_lgcp = lgcp_sample$edge_number/max(lgcp_sample$edge_number)),
                       normalized = TRUE)

#Add integration points
graph$add_observations(data = data.frame(y = rep(0,length(atilde)),
                                         e = atilde,
                                         edge_number = graph$mesh$VtE[,1],
                                         distance_on_edge = graph$mesh$VtE[,2],
                                         Intercept = 1,
                                         cov_lgcp = graph$mesh$VtE[,1]/max(lgcp_sample$edge_number)),
                       normalized = TRUE)

```
We now create the `inla` model object with the
`graph_spde` function. For simplicity, we assume that $\alpha$ is known and fixed
to the true value in the model.  

```{r}
rspde_model <- rspde.metric_graph(graph, nu = alpha - 1/2)
```

Next, we compute the auxiliary data:

```{r}
data_rspde <- graph_data_spde(rspde_model, name="field", covariates = c("Intercept","cov_lgcp"))
```

We now create the `inla.stack` object with the `inla.stack()` function. 
At this stage, it is important that the data has been added to the `graph` since
it is supplied to the stack by using the `graph_spde_data()` function. 

```{r}
stk <- inla.stack(data = data_rspde[["data"]], 
                  A = data_rspde[["basis"]],
                  effects = data_rspde[["index"]])
```

We can now fit the model using `R-INLA`:

```{r, message=FALSE, warning=FALSE}
spde_fit <- inla(y ~ -1 + Intercept + cov_lgcp + f(field, model = rspde_model), 
                 family = "poisson", data = inla.stack.data(stk),
                 control.predictor = list(A = inla.stack.A(stk), compute = TRUE),
                 E = inla.stack.data(stk)$e)
```

Let us extract the estimates in the original scale
by using the `spde_metric_graph_result()` function, then
taking a `summary()`:
      
```{r}
spde_result <- rspde.result(spde_fit, "field", rspde_model)

summary(spde_result)
```

We will now compare the means of the estimated values with
the true values:

```{r}
  result_df <- data.frame(
    parameter = c("std.dev", "range"),
    true = c(sigma, range),
    mean = c(
      spde_result$summary.std.dev$mean,
      spde_result$summary.range$mean
    ),
    mode = c(
      spde_result$summary.std.dev$mode,
      spde_result$summary.range$mode
    )
  )
  print(result_df)
```


# An example with replicates

We now clear the previous data and add the new data together with the exposure terms
```{r}
  graph$clear_observations()
  df_rep <- data.frame(y=rep(1,length(lgcp_sample_rep[[1]]$edge_loc)),
                                             e = rep(0,length(lgcp_sample_rep[[1]]$edge_loc)),
                                         edge_number = lgcp_sample_rep[[1]]$edge_number,
                                         distance_on_edge = lgcp_sample_rep[[1]]$edge_loc,
                                         Intercept = 1,
                                         cov_lgcp = lgcp_sample_rep[[1]]$edge_number/max_edge_num,
                                         rep = rep(1,length(lgcp_sample_rep[[1]]$edge_loc)))

  df_rep <- rbind(df_rep, data.frame(y = rep(0,length(atilde)),
                                         e = atilde,
                                         edge_number = graph$mesh$VtE[,1],
                                         distance_on_edge = graph$mesh$VtE[,2],
                                         Intercept = 1,
                                         cov_lgcp = graph$mesh$VtE[,1]/max_edge_num,
                                         rep = rep(1,length(atilde))))
  for(i in 2:n.rep){
    df_rep <- rbind(df_rep, data.frame(y=rep(1,length(lgcp_sample_rep[[i]]$edge_loc)),
                                             e = rep(0,length(lgcp_sample_rep[[i]]$edge_loc)),
                                         edge_number = lgcp_sample_rep[[i]]$edge_number,
                                         distance_on_edge = lgcp_sample_rep[[i]]$edge_loc,
                                         Intercept = 1,
                                         cov_lgcp = lgcp_sample_rep[[i]]$edge_number/max_edge_num,
                                         rep = rep(i,length(lgcp_sample_rep[[i]]$edge_loc))))
    df_rep <- rbind(df_rep, data.frame(y = rep(0,length(atilde)),
                                         e = atilde,
                                         edge_number = graph$mesh$VtE[,1],
                                         distance_on_edge = graph$mesh$VtE[,2],
                                         Intercept = 1,
                                         cov_lgcp = graph$mesh$VtE[,1]/max_edge_num,
                                         rep = rep(i,length(atilde))))                                        

  }

      graph$add_observations(data = df_rep,
                       normalized = TRUE,
                        group = "rep")

```
We can now define and fit the model as previously
```{r}
rspde_model <- rspde.metric_graph(graph, nu = alpha - 1/2)

data_rspde <- graph_data_spde(rspde_model, name = "field", 
                                  repl = ".all", repl_col = "rep", 
                                  covariates = c("Intercept","cov_lgcp"))

stk <- inla.stack(data = data_rspde[["data"]], 
                  A = data_rspde[["basis"]], 
                  effects = data_rspde[["index"]])

spde_fit <- inla(y ~ -1 + Intercept + cov_lgcp + 
                 f(field, model = rspde_model, replicate = field.repl), 
                 family = "poisson", data = inla.stack.data(stk),
                 control.predictor = list(A = inla.stack.A(stk), compute = TRUE),
                 E = inla.stack.data(stk)$e)
```
Let's look at the summaries
```{r}
spde_result <- rspde.result(spde_fit, "field", rspde_model)
summary(spde_result)
result_df <- data.frame(
    parameter = c("std.dev", "range"),
    true = c(sigma, range),
    mean = c(
      spde_result$summary.std.dev$mean,
      spde_result$summary.range$mean
    ),
    mode = c(
      spde_result$summary.std.dev$mode,
      spde_result$summary.range$mode
    )
  )
  print(result_df)
```

## References
