---
  title: "Whittle--Matérn fields with general smoothness"
author: "David Bolin, Alexandre B. Simas"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Whittle--Matérn fields with general smoothness}
%\VignetteEngine{knitr::rmarkdown}
%\VignetteEncoding{UTF-8}
references:
  - id: xiong22
title: "Covariance-based rational approximations of fractional SPDEs for computationally efficient Bayesian inference"
author:
  - family: Xiong
given: Zhen
- family: Simas
given: Alexandre B.
- family: Bolin
given: David
container-title: arXiv:2209.04670
type: preprint
issued:
  year: 2022
- id: graph_fem
title: "Regularity and numerical approximation of fractional elliptic differential equations on compact metric graphs"
author:
  - family: Bolin
given: David
- family: Kovács
given: Mihály
- family: Kumar
given: Vivek
- family: Simas
given: Alexandre B.
container-title: arXiv:??
  type: preprint
issued:
  year: 2022
---

  ```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
set.seed(1)
```

## Introduction

In this vignette we will introduce how to work with log-Gaussian Cox processes
based on Whittle--Matérn fields on metric graphs. To simplify the integration
with `R-INLA` and `inlabru` hese models are implemented using finite element 
approximations. 

For further details on the construction of metric graphs, see
[Working with metric graphs](metric_graphs.html)

## Constructing the graph and the mesh
We begin by loading the `rSPDE` and `MetricGraph` packages:

```{r, message=FALSE, warning=FALSE}
library(rSPDE)
library(MetricGraph)
```

As an example, we consider the following metric graph
```{r, fig.show='hold',fig.align = "center",fig.width=7,echo=TRUE}
library(sp)
line1 <- Line(rbind(c(0,0),c(1,0)))
line2 <- Line(rbind(c(0,0),c(0,1)))
line3 <- Line(rbind(c(0,1),c(-1,1)))
line4 <- Line(rbind(c(0,0),c(-1,0)))
line5 <- Line(rbind(c(-1,0),c(-1,1)))
Lines = SpatialLines(list(Lines(list(line1),ID="1"),
                          Lines(list(line2),ID="2"),
                          Lines(list(line3),ID="3"),
                          Lines(list(line4),ID="4"),
                          Lines(list(line5),ID="5")))
graph <- metric_graph$new(lines = Lines)
graph$plot()
```

To construct a FEM approximation of a Whittle--Matérn field,
we must first construct a mesh on the graph.
```{r}
graph$build_mesh(h = 0.01)
graph$plot(mesh=TRUE)
```

The next step is to build the mass and stiffness matrices for the FEM basis.
```{r}
  graph$compute_fem()
```

We are now ready to specify the and sample from a log-Gaussian Cox process model
with intensity $\lambda = \exp(u)$ where $u$ is a Gaussian Whittle--Matérn field specified by
$$
(\kappa^2 - \Delta)^{\alpha/2} \tau u = \mathcal{W}.
$$
For this we can use the function `graph_lgcp` as follows:
```{r}
  tau <- 1
  kappa <- 1
  lgcp_sample <- graph_lgcp(intercept = 1, kappa = kappa, tau = tau, alpha = 2, graph = graph)
```
The object returned by the function is a list with the simulated Gaussian process and the points on the graph. We can plot the simulated intensity function as
```{r, message=FALSE}
graph$plot_function(exp(lgcp_sample$u), vertex_size = 0)
```
To plot the points, we can add them to the graph and then plot:
```{r}
graph$add_observations(data = data.frame(y=rep(0,length(lgcp_sample$edge_loc)),
                                         edge_number = lgcp_sample$edge_number,
                                         distance_on_edge = lgcp_sample$edge_loc),
                       normalized = TRUE)
graph$plot(vertex_size = 0, data = "y")
```





## References
