---
title: "Gaussian random fields on metric graphs"
author: "David Bolin, Alexandre B. Simas, and Jonas Wallin"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Random fields on metric graphs}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
references:
- id: BSW2022a
  title: "Gaussian Whittle--Matéern fields on metric graphs"
  author:
  - family: Bolin
    given: David
  - family: Simas
    given: Alexandre B.
  - family: Wallin
    given: Jonas
  container-title: arXiv:2205.06163
  type: preprint
  issued:
  year: 2022
- id: BSW2022b
  title: "Statistical properties of Gaussian Whittle--Matérn fields on metric graphs"
  author:
  - family: Bolin
    given: David
  - family: Simas
    given: Alexandre B.
  - family: Wallin
    given: Jonas
  container-title: arXiv:??
  type: preprint
  issued:
  year: 2022
- id: Anderes2020
  title: "Isotropic covariance functions on graphs and their edges"
  author:
  - family: Anderes
    given: Ethan
  - family: Møller
    given: Jesper
  - family: Rasmussen
    given: Jakob G
  container-title: Annals of Statistics
  type: article
  issued: 48
  pages: 2478--2503
  year: 2020
- id: Borovitskiy2021
  title: "Matérn Gaussian processes on graphs"
  author:
  - family: Borovitskiy
    given: Viacheslav
  - family: Azangulov
    given: Iskander
  - family: Terenin
    given: Alexander 
  - family: Mostowsky
    given: Peter
  - family: Deisenroth
    given: Marc
  - family: Durrande
    given: Nicolas
  container-title: International Conference on Artificial Intelligence and Statistics
  type: article
  issued: 
  pages: 2593--2601
  year: 2021
---

	
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
set.seed(1)
library(GPGraph)
```

# Introduction

In this vignette we will introduce how to work with Gaussian random fields on
metric graphs. The main models are the Whittle--Matérn fields introduced in 
[@BSW2022a](https://arxiv.org/abs/2205.06163) and [@BSW2022b](??). 

The package also has support for isotropic Gaussian processes, and in particular
Gaussian processes with isotropic exponential covariance functions as introduced
by [@Anderes2020](https://projecteuclid.org/journals/annals-of-statistics/volume-48/issue-4/Isotropic-covariance-functions-on-graphs-and-their-edges/10.1214/19-AOS1896.full). Finally, Gaussian processes based on the graph Laplacian, as introduced
by [@Borovitskiy2021](http://proceedings.mlr.press/v130/borovitskiy21a/borovitskiy21a.pdf) are also supported. 

As an example throughout the vignette, we consider the following metric graph:
```{r}
library(sp)
line1 <- Line(rbind(c(0,0),c(1,0)))
line2 <- Line(rbind(c(0,0),c(0,1)))
line3 <- Line(rbind(c(0,1),c(-1,1)))
theta <- seq(from=pi,to=3*pi/2,length.out = 20)
line4 <- Line(cbind(sin(theta),1+ cos(theta)))
Lines = sp::SpatialLines(list(Lines(list(line1),ID="1"),
                              Lines(list(line2),ID="2"),
                              Lines(list(line3),ID="3"),
                              Lines(list(line4),ID="4")))
graph <- metric_graph$new(Lines = Lines)
graph$plot(line_width = 0.3)
```

For further details on the construction of metric graphs, see
[Working with metric graphs](metric_graphs.html)

# Whittle--Matérn fields 
The Whittle--Matérn fields are specified as solutions to the stochastic
differential equation 
$$
  (\kappa^2 - \Delta)^{\alpha} \tau u = \mathcal{W}
$$
on the metric graph $\Gamma$. We can work with these models without
and approximations if the smoothness parameter $\alpha$ is an integer, and this 
is what we focus on in this vignette. For details on the case of a general 
smoothness parameter, see [Whittle--Matérn fields with general smoothness](fem_models.html). 

## Sampling 

As an example, let us simulate the field $u$ on the graph using $\alpha = 1$. 
To do so, we first need to specify
where to sample it. As a first example, let us specify some locations manually:
```{r}
PtE <- cbind(rep(1:4, each = 4),
             rep(c(0.2, 0.4, 0.6, 0.8), times = 4))

sigma <- 2
alpha <- 1
kappa <- 5
u <- sample_spde(kappa = kappa, sigma = sigma, alpha = alpha,
                 graph = graph, PtE = PtE)
graph$plot(X = u, X_loc = PtE)
```

In many cases, one wants to sample the field at evenly spaced locations over
the graph. To avoid having to specify such locations manually, we can first
create a mesh on the graph
```{r}
graph$build_mesh(h = 0.1)
graph$plot(mesh=TRUE)
```

In the command `build_mesh`, the argument `h` decides the largest spacing between nodes in the mesh. We can now sample the field on this mesh and plot the 
result as a function as follows:
```{r}
u <- sample_spde(kappa = kappa, sigma = sigma, alpha = alpha,
                 graph = graph, type = "mesh")
graph$plot_function_mesh(u)
```

Let us construct a finer mesh, simulate the field, and visualize the simulation
in 3D by specifying the `plotly` argument in the plot function:
```{r}
graph$build_mesh(h = 0.01)
u <- sample_spde(kappa = kappa, sigma = sigma, alpha = alpha,
                 graph = graph, type = "mesh")
graph$plot_function_mesh(u, plotly = TRUE)
```

Since $\alpha=1$, these sample paths are continuous but not differentiable. 
To visualize the correlation structure of the field, we can compute and plot
the covariances between some point and all other points in the graph as follows:
```{r}
C <- covariance_alpha1_mesh(c(2, 0.2), kappa = kappa, sigma = sigma,
                            graph = graph)
graph$plot_function_mesh(C, plotly = TRUE)
```

To obtain a field with differentiable sample paths, we can change to $\alpha=2$.
The corresponding covariance function then looks as follows: 
```{r}
C <- covariance_alpha2_mesh(c(2, 0.2), kappa = kappa, sigma = sigma,
                            graph = graph)
graph$plot_function_mesh(C, plotly = TRUE)
```

Let us simulate a process with $\alpha=2$ as well:
```{r}
u <- sample_spde(kappa = kappa, sigma = sigma, alpha = 2,
                 graph = graph, type = "mesh")
graph$plot_function_mesh(u, plotly = TRUE)
```

## Inference

Mostly for illustration purposes, the `GPGraph` package contains implementations
of likelihoods for Whittle--Matérn fields observed under Gaussian measurement 
noise. In this section we will illustrate these methods. For the use of the 
Whittle--Matérn fields in more complicated hierarchical models, we recommend
using the interfaces to the `INLA` and `inlabru` packages. See [INLA and inlabru interfaces](inla_interface.html) for further details on these.

Suppose that we want to estimate the model parameters of a Whittle--Matérn field
$u(s)$ observed under Gaussian measurement noise. That is, we assume that we
are given observations
$$
y_i = u(s_i) + \varepsilon_i, \quad i=1,\ldots,n
$$
where $s_i\in \Gamma$ are observation locations and $\varepsilon_i$ are
independent centered Gaussian variables $N(0,\sigma_e^2)$ representing 
measurement noise. 

Let us start by generating some data like this and adding it to the 
metric graph:

```{r}
kappa <- 10
sigma <- 2
sigma_e <- 0.1
alpha <- 1
theta <-  c(sigma_e, kappa, sigma)

n.obs.per.edge <- 10
PtE <- NULL
for(i in 1:graph$nE){
  #add locations sampled at random to each edge
  PtE <- rbind(PtE, cbind(rep(i, n.obs.per.edge), runif(n.obs.per.edge)))
}

u <- sample_spde(kappa = kappa, sigma = sigma, alpha = alpha,
                 graph = graph, PtE = PtE)

y <- u + sigma_e*rnorm(n.obs.per.edge * graph$nE)

graph$add_observations2(y,PtE, normalized = TRUE)
graph$plot(data = TRUE)
```

We can now use the `likelihood_graph_spde` function to evaluate the 
log-likelihood function for this data based on a Whittle--Matérn field 
model for $u$. Assuming that $\alpha=1$, we optimize this likelihood using 
`optim` as follows:

```{r}
sigma_e_start <- 0.2
sigma_start <- 2
kappa_start <- 10

theta0 <- c(sigma_e_start, sigma_start, kappa_start) 
res <- optim(log(theta0), function(x) -likelihood_graph_spde(exp(x),
                                                                   graph,
                                                                   alpha = 1))
sigma_e_est <- exp(res$par[1])
sigma_est <- exp(res$par[2])
kappa_est <- exp(res$par[3])
results <- data.frame(sigma_e = c(sigma_e, sigma_e_est),
                      sigma = c(sigma, sigma_est),
                      kappa = c(kappa, kappa_est),
                      row.names = c("Truth", "Estimate"))
print(results)
```
Given these estimated parameters, we can now do kriging to estimate the 
field at locations in the graph. As an example, we now estimate the field 
on the regular mesh that we previously constructed. 
```{r}
u_est <- spde_posterior_mean(c(sigma_e_est, sigma_est, kappa_est),
                             graph, alpha = 1, type = "mesh")

graph$plot_function_mesh(u_est, plotly = TRUE)
```

The same procedure can be done with $\alpha = 2$. One can also estimate $\alpha$
from data as described in the vignette 
[Whittle--Matérn fields with general smoothness](fem_models.html).

# Isotropic Gaussian processes

Let us now see how we can fit a model with an isotropic exponential covariance
function 
$$
r(d(s,t)) = \sigma^2\exp(-\kappa d(s,t))
$$
to the data we generated above. For this we must first add the 
observation locations to the graph. Let us first copy the graph to a new 
graph object and then add the observation locations to it

```{r}
graph2 <- graph$clone()
graph2$observation_to_vertex()
```

The model is based on using the so-called resistance metric as the distance 
$d(s,t)$ between the observations when calculating the covariance function. 
This can be done as follows:
```{r}
graph2$compute_resdist()
```
We can now fit the model to the data by using `optim` in combination with the 
function `likelihood_graph_covariance()` which implements, among other models,
Gaussian processes with isotropic exponential covariance functions.
```{r}
res.exp <- optim(log(theta0), 
                 function(x) -likelihood_graph_covariance(exp(x), graph2,
                                                          model = "isoExp"))
theta.exp <- exp(res.exp$par)

```

Let us now compute the posterior mean for the field at the observation 
locations:

```{r}
u_est_exp <- posterior_mean_covariance(theta.exp, graph2, model = "isoExp")
graph2$plot(X = u_est_exp, X_loc = graph$PtE)
```

Let us plot the estimated covariance function at the mesh nodes in the 
original mesh. For this, we first compute the resistance metric between 
the vertices in the mesh. We can then construct and plot the covariance 
between the field at the first mesh location and the rest as follows:
```{r}
graph$compute_resdist_mesh()
C <- theta.exp[2]^2*exp(-theta.exp[3]*graph$mesh$res.dist[1,])
graph$plot_function_mesh(C, plotly = TRUE)
```
# Models based on the Graph Laplacian


# References
