
---
title: "Spatio-temporal Models on Compact Metric Graphs"
author: "David Bolin, Alexandre B. Simas, and Jonas Wallin"
date: "Created: 2024-10-23. Last modified: `r Sys.Date()`."
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Spatio-temporal Models on Compact Metric Graphs}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
references:
- id: lindgren24
  title: "A Diffusion-based Spatio-temporal Extension of Gaussian Matérn Fields"
  author:
  - family: Lindgren
    given: Finn
  - family: Bakka
    given: Haakon  
  - family: Bolin
    given: David
  - family: Krainski
    given: Elias
  - family: Rue
    given: Håvard
  container-title: SORT Statistics and Operations Research Transactions
  volume: 48
  issue: 1
  page: 3-66
  type: article-journal
  issued:
    year: 2024
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
library(rSPDE)
set.seed(123)
```

```{r inla_link, include = FALSE}
inla_link <- function() {
  sprintf("[%s](%s)", "`R-INLA`", "https://www.r-inla.org")
}
```

## Introduction

The `MetricGraph` package, combined with the `rSPDE` package, implements the following spatio-temporal model on compact metric graphs:
$$
d u + \gamma(\kappa^2 + \rho \cdot \nabla - \Delta)^{\alpha} u = dW_Q, \quad \text{on } T \times \Gamma
$$
where $T$ is a temporal interval, and $\Gamma$ is a compact metric graph. Here, $\kappa > 0$ is a spatial range parameter, $\rho$ is a drift parameter, and $W_Q$ is a $Q$-Wiener process with spatial covariance operator $\sigma^2(\kappa^2 - \Delta)^{-\beta}$, where $\sigma^2$ is a variance parameter. The model has two smoothness parameters, $\alpha$ and $\beta$, which are assumed to be integers. This model generalizes the spatio-temporal models introduced in 
[@lindgren24](https://www.raco.cat/index.php/SORT/article/view/428665), where the generalization is to allow for drift and to allow for metric graphs as spatial domains. The model is implemented using a finite element discretization of the corresponding precision operator:
$$
\sigma^{-2} \left(d + \gamma(\kappa^2 + \rho \cdot \nabla - \Delta)^{\alpha}\right) \left(d + \gamma(\kappa^2 - \rho \cdot \nabla - \Delta)^{\alpha}\right) (\kappa^2 - \rho \cdot \nabla - \Delta)^{\beta}
$$
in both space and time, following a similar discretization to [@lindgren24](https://www.raco.cat/index.php/SORT/article/view/428665).

## Implementation Details

The function `spacetime.operators()`, provided by the `rSPDE` package, constructs the spatio-temporal precision matrix for models on metric graphs and enables interaction with the `MetricGraph` package. This function requires inputs that define the spatial structure via a `MetricGraph` object and the temporal domain through a vector of discretized time points. Key parameters like `kappa` and `sigma` control the range and variance of the field, respectively, whereas `alpha` and `beta` govern the smoothness of the field.

This function processes these inputs to create a `spacetime.operators` object containing the discretized precision matrix, covariance structure, and essential methods for model evaluation, simulation, and kriging on metric graphs.

```{r, message=FALSE}
library(sp)
library(rSPDE)
library(MetricGraph)

# Define graph edges
line1 <- Line(rbind(c(1, 0), c(0, 0)))
line2 <- Line(rbind(c(0, 0), c(0, 1)))
line3 <- Line(rbind(c(0, 0), c(0, -1)))
line4 <- Line(rbind(c(0, 1), c(1, 0)))
lines <- sp::SpatialLines(list(
  Lines(list(line1), ID="1"),
  Lines(list(line2), ID="2"),
  Lines(list(line3), ID="3"),
  Lines(list(line4), ID="4")
))

# Initialize metric graph
graph <- metric_graph$new(edges = lines)
graph$plot(direction = TRUE)

# Create finite element mesh and compute matrices
h <- 0.01
graph$build_mesh(h = h)
graph$compute_fem()

# Define model parameters and construct spatio-temporal operator
t <- seq(from = 0, to = 10, length.out = 31)
kappa <- 5
sigma <- 10
gamma <- 0.1
rho <- 0.3
alpha <- 1
beta <- 1

op <- spacetime.operators(graph = graph, time = t,
                          kappa = kappa, sigma = sigma, alpha = alpha,
                          beta = beta, rho = rho, gamma = gamma)
op$plot_covariances(t.ind = 15, s.ind = 50, t.shift = c(-1, 0, 1))
```

### Simulating Observations and Estimating Parameters

We simulate some data from the model:

```{r, message=FALSE}
x <- simulate(op, nsim = 1) 

n.obs <- 500
PtE <- cbind(sample(1:graph$nE, size = n.obs, replace = TRUE), runif(n.obs))
t.obs <- max(t) * runif(n.obs)
A <- op$make_A(PtE, t.obs)
sigma.e <- 0.01
Y <- as.vector(A %*% x + sigma.e * rnorm(n.obs))
df <- data.frame(y = as.matrix(Y), edge_number = PtE[,1], 
                    distance_on_edge = PtE[,2], 
                    time = t.obs)
graph$add_observations(data = df, normalized=TRUE,
                          group = "time")
```

Next, we estimate the model parameters:

```{r}
res <- rspde_lme(y ~ 1, loc_time = "time",
                 model = op, 
                 parallel = TRUE)
```

After fitting the model, we can view a summary of the results:

```{r}
summary(res)
```

Compare the estimated results with the true values:

```{r}
results <- data.frame(kappa = c(kappa, res$coeff$random_effects[1]), 
                      sigma = c(sigma, res$coeff$random_effects[2]),
                      gamma = c(gamma, res$coeff$random_effects[3]), 
                      rho = c(rho, res$coeff$random_effects[4]),
                      sigma.e = c(sigma.e, res$coeff$measurement_error),
                      intercept = c(0, res$coeff$fixed_effects),
                      row.names = c("True", "Estimate"))
                      
print(results)
```

## `INLA` and `inlabru` Implementation

For completeness, we fit the model using `INLA` and `inlabru` packages. First, we create a temporal mesh:

```{r, message=FALSE}
library(INLA)
library(inlabru)
library(fmesher)
mesh_time <- fm_mesh_1d(t)
st_graph <- rspde.spacetime(mesh_space = graph, 
                                mesh_time = mesh_time, 
                                alpha = 1, beta = 1)
```

We now define the data object, by using the `graph_data_rspde()` function, with `bru` set to `FALSE`.

```{r}
st_data_inla <- graph_data_rspde(st_graph, name = "field", time = "time", bru = FALSE)
```

Now we create the `inla.stack` object. To such an end, observe that `st_data` contains the dataset as the `data` component, the index as the `index` component and the so-called `A` matrix as the `basis` component. We will now create the stack using these components:
```{r}
  stk.dat <- inla.stack(
    data = st_data_inla[["data"]], 
    A = st_data_inla[["basis"]],
    effects =
      c(st_data_inla[["index"]],
        list(Intercept = 1))
    )
```

Finally, we create the formula object and fit the model:

```{r}
  f_st <- y ~ -1 + Intercept + f(field, model = st_graph)

  rspde_fit <- inla(f_st, data = inla.stack.data(stk.dat),
    control.inla = list(int.strategy = "eb"),
    control.predictor = list(A = inla.stack.A(stk.dat), compute = TRUE),
    num.threads = "1:1"
  )
```

Let us compare the fitted values with the true values:
```{r}
param_st <- transform_parameters_spacetime(rspde_fit$summary.hyperpar$mean[2:6], 
                                            st_graph)

results <- data.frame(kappa = c(kappa,  param_st[[1]]), 
                      sigma = c(sigma,  param_st[[2]]),
                      gamma = c(gamma,  param_st[[3]]), 
                      rho = c(rho, param_st[[4]]),
                      sigma.e = c(sigma.e, sqrt(1/rspde_fit$summary.hyperpar$mean[1])),
                      intercept = c(0, rspde_fit$summary.fixed$mean),
                      row.names = c("True", "Estimate"))
                      
print(results)
```

Now, we will proceed with the `inlabru` implementation. Define the data, in which we use the function `graph_data_rspde()`, set `bru` to `TRUE`. Next, we define the model component for `inlabru`, where we pass a list with two elements, `space` which is given by the spatial locations, for which we pass `cbind(.edge_number, .distance_on_edge)`, and `time` which is given by the time points.

```{r}
st_data_bru <- graph_data_rspde(st_graph, bru = TRUE)

cmp_graph <- y ~ 1 + field(list(
                    space = cbind(.edge_number, .distance_on_edge), 
                                time = time), 
                                model = st_graph)
```

Fit the model passing `st_data_bru` to the `data` argument:

```{r}
bru_fit_graph <- bru(cmp_graph, data = st_data_bru[["data"]])
```

Compare the estimated parameters with the true values:

```{r}
param_st <- transform_parameters_spacetime(bru_fit_graph$summary.hyperpar$mean[2:6], 
                                            st_graph)

results <- data.frame(kappa = c(kappa,  param_st[[1]]), 
                      sigma = c(sigma,  param_st[[2]]),
                      gamma = c(gamma,  param_st[[3]]), 
                      rho = c(rho, param_st[[4]]),
                      sigma.e = c(sigma.e, sqrt(1/bru_fit_graph$summary.hyperpar$mean[1])),
                      intercept = c(0, bru_fit_graph$summary.fixed$mean),
                      row.names = c("True", "Estimate"))
                      
print(results)
```

## Fitting model with replicates in `INLA` and `inlabru` Implementation

We will now simulate from the same model, however, this time, with replicates. We will assume we have 10 replicates.

Let us start by simulating the field:

```{r}
n_rep = 10
x <- simulate(op, nsim = n_rep) 

n.obs <- 500
PtE <- cbind(sample(1:graph$nE, size = n.obs, replace = TRUE), runif(n.obs))
t.obs <- max(t) * runif(n.obs)
A <- op$make_A(PtE, t.obs)
sigma.e <- 0.01
Y <- as.vector(A %*% x + sigma.e * rnorm(n.obs))
df <- data.frame(y = as.matrix(Y), edge_number = PtE[,1], 
                    distance_on_edge = PtE[,2], time = t.obs,
                    repl = rep(1:n_rep, each = n.obs))

graph_rep <- graph$clone()
```

Let us now add the observations into the metric graph object. Observe that in this time we have two grouping variables, `time` and `repl`. We need grouping variable whenever we want to allow the storage of different observations at the exact same location.

```{r}
graph_rep$add_observations(data = df, normalized=TRUE, 
                clear_obs = TRUE, group = c("time", "repl"))
```

Let us now create the `rSPDE` model object. Observe that we cannot use the previous one as the data inside the graph is now different.

```{r}
st_graph_repl <- rspde.spacetime(mesh_space = graph_rep, 
                                mesh_time = mesh_time, 
                                alpha = 1, beta = 1)
```

We will start with `INLA`'s implementation. To this end, let us create the data object.

```{r}
st_data_inla_repl <- graph_data_rspde(st_graph_repl, 
                                  repl = ".all",
                                  repl_col = "repl",
                                  time = "time",
                                  bru = FALSE)
```

Let us now create the corresponding `inla.stack` object:
```{r}
st_dat_rep <- inla.stack(
  data = st_data_inla_repl[["data"]],
  A = st_data_inla_repl[["basis"]],
  effects = c(st_data_inla_repl[["index"]],
          list(Intercept = 1))
)
```

Now, we can create the formula object and fit the model:

```{r}
f_st_rep <-
  y ~ -1 + Intercept + f(field,
    model = st_graph_repl,
    replicate = field.repl
  )

rspde_fit_rep <-
  inla(f_st_rep,
    data = inla.stack.data(st_dat_rep),
    control.predictor =
      list(A = inla.stack.A(st_dat_rep))
  )
```


Let us compare the fitted values with the true values:
```{r}
param_st_rep <- transform_parameters_spacetime(rspde_fit_rep$summary.hyperpar$mean[2:6], 
                                            st_graph)

results_inla_repl <- data.frame(kappa = c(kappa,  param_st_rep[[1]]), 
                      sigma = c(sigma,  param_st_rep[[2]]),
                      gamma = c(gamma,  param_st_rep[[3]]), 
                      rho = c(rho, param_st_rep[[4]]),
                      sigma.e = c(sigma.e, sqrt(1/rspde_fit_rep$summary.hyperpar$mean[1])),
                      intercept = c(0, rspde_fit_rep$summary.fixed$mean),
                      row.names = c("True", "Estimate"))
                      
print(results_inla_repl)
```

We will now proceed to `inlabru`'s implementation. Let us now create the data object, by using the `graph_data_rspde()` function. In this case, we also need to pass `repl_col`:

```{r}
st_data_bru_repl <- graph_data_rspde(st_graph_repl, 
                                  repl = ".all",
                                  repl_col = "repl",
                                  bru = TRUE)
```

Next, we create `inlabru` component by also passing the replicate argument with the `repl` element of the object returned by `graph_data_rspde()`:
```{r}
cmp_graph_repl <- y ~ 1 + field(list(space = cbind(.edge_number, .distance_on_edge), 
                                time = time),
                                model = st_graph_repl,
                                replicate = st_data_bru_repl[["repl"]]
                                )
```

Fit the model passing `st_data_bru_repl` to the `data` argument:

```{r}
bru_fit_graph_repl <- bru(cmp_graph_repl, 
                      data = st_data_bru_repl[["data"]])
```

Compare the estimated parameters with the true values:

```{r}
param_st_repl <- transform_parameters_spacetime(
                              bru_fit_graph_repl$summary.hyperpar$mean[2:6], 
                              st_graph_repl)

results_repl <- data.frame(kappa = c(kappa,  param_st_repl[[1]]), 
                      sigma = c(sigma,  param_st_repl[[2]]),
                      gamma = c(gamma,  param_st_repl[[3]]), 
                      rho = c(rho, param_st_repl[[4]]),
                      sigma.e = c(sigma.e, sqrt(1/bru_fit_graph_repl$summary.hyperpar$mean[1])),
                      intercept = c(0, bru_fit_graph_repl$summary.fixed$mean),
                      row.names = c("True", "Estimate"))
                      
print(results_repl)
```


## Example: Setting `bounded_rho = FALSE`

The `bounded_rho` parameter in `rspde.spacetime()` ensures that \(
ho\) stays within a range that guarantees numerical stability and well-posedness. However, in some cases, better fits can be achieved by setting `bounded_rho = FALSE`. Below, we show how to modify the model to handle this case.

### Using `rspde_lme()`

Here we set `bounded_rho = FALSE` when creating the spatio-temporal operator and fit the model using `rspde_lme()`:

```{r}
# Create the spatio-temporal operator with bounded_rho = FALSE
op_unbounded <- spacetime.operators(
  graph = graph,
  time = t,
  kappa = kappa,
  sigma = sigma,
  alpha = alpha,
  beta = beta,
  rho = rho,
  gamma = gamma,
  bounded_rho = FALSE
)

# Fit the model using rspde_lme
res <- rspde_lme(y ~ 1, loc_time = "time", model = op_unbounded, parallel = TRUE)
```

#### Comparing Results

```{r}
# Compare results for rspde_lme
results <- data.frame(
  kappa = c(kappa, res$coeff$random_effects[1]),
  sigma = c(sigma, res$coeff$random_effects[2]),
  gamma = c(gamma, res$coeff$random_effects[3]),
  rho = c(rho, res$coeff$random_effects[4]),
  sigma.e = c(sigma.e, res$coeff$measurement_error),
  intercept = c(0, res$coeff$fixed_effects),
  row.names = c("True", "Estimate")
)

print(results)
```

### Using `inlabru`

Here we set `bounded_rho = FALSE` when creating the spatio-temporal operator and prepare data for `inlabru`:

```{r}
# Create the spatio-temporal operator for inlabru
st_graph_unbounded <- rspde.spacetime(
  mesh_space = graph,
  mesh_time = mesh_time,
  alpha = 1,
  beta = 1,
  bounded_rho = FALSE
)

# Prepare data for inlabru
st_data_bru_unbounded <- graph_data_rspde(
  st_graph_unbounded,
  bru = TRUE
)

# Define inlabru component with unbounded rho
cmp_graph_unbounded <- y ~ 1 + field(
  list(
    space = cbind(.edge_number, .distance_on_edge),
    time = time
  ),
  model = st_graph_unbounded
)
```

#### Fitting the Model with `inlabru`

```{r}
bru_fit_graph_unbounded <- bru(cmp_graph_unbounded, data = st_data_bru_unbounded[["data"]])
```

#### Comparing Results

```{r}
# Transform parameters and compare results
param_st_unbounded <- transform_parameters_spacetime(
  bru_fit_graph_unbounded$summary.hyperpar$mean[2:6],
  st_graph_unbounded
)

results_unbounded <- data.frame(
  kappa = c(kappa, param_st_unbounded[[1]]),
  sigma = c(sigma, param_st_unbounded[[2]]),
  gamma = c(gamma, param_st_unbounded[[3]]),
  rho = c(rho, param_st_unbounded[[4]]),
  sigma.e = c(sigma.e, sqrt(1 / bru_fit_graph_unbounded$summary.hyperpar$mean[1])),
  intercept = c(0, bru_fit_graph_unbounded$summary.fixed$mean),
  row.names = c("True", "Estimate")
)

print(results_unbounded)
```


## References
